---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

## Multivoxel Pattern Analysis in R

[![codecov](https://codecov.io/gh/rMVPA/NA/branch/master/graph/badge.svg)](https://codecov.io/github/bbuchsbaum/rMVPA?branch=master)
[![Build Status](https://travis-ci.org/rMVPA/rMVPA.svg?branch=master)](https://travis-ci.org/bbuchsbaum/rMVPA)

```{r, echo = FALSE}                                                                                                        
version <- as.vector(read.dcf('DESCRIPTION')[, 'Version'])                                                                 
version <- gsub('-', '.', version)                                                                                      
```                                                                                                                        

### This package is under development.

### Introduction

`rMVPA` is an R library for multivariate pattern analysis of neuroimaging data. The goal of this library is to make MVPA analyses easy. It can be used both programmatically from within R or using a command line interface. 'rMVPA' leverages the 'caret' library for the underlying machine learning interface. What `rMVPA` provides is the infrastructure for conducting machine learning analyses on neuroimaging data. 

Documentation and vignettes: https://bbuchsbaum.github.io/rMVPA/

### Installation

### Using devtools

To install `rMVPA` from within R, use the `devtools` function `install_github`. You will need the development version of `neuroim2` as well.

From within R:

```
library(devtools)
install_github("bbuchsbaum/neuroim2")
install_github("bbuchsbaum/rMVPA")
```

### Using `git` from the command line

```
git clone git@github.com:bbuchsbaum/rMVPA.git
R CMD install rMVPA
```

### Optionally install command line scripts for "coding-free" MVPA analysis:

`wget https://raw.githubusercontent.com/bbuchsbaum/rMVPA/master/scripts/MVPA_Searchlight.R`

`wget https://raw.githubusercontent.com/bbuchsbaum/rMVPA/master/scripts/MVPA_Regional.R`

Then, move these files to a folder on your `PATH` and make them executable:

`chmod +x MVPA_Searchlight.R`

`chmod +x MVPA_Regional.R`



