---
title: "Advanced RSA Methods: Feature-Based and Vector-Based Approaches"
author: "Bradley Buchsbaum"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Advanced RSA Methods: Feature-Based and Vector-Based Approaches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, message=FALSE}
suppressPackageStartupMessages({
  library(neuroim2)
  library(rMVPA)
  library(dplyr)
})
```

# Introduction

While standard Representational Similarity Analysis (RSA) compares neural patterns with model-based similarity structures, there are cases where more specialized approaches are needed. The `rMVPA` package provides two advanced RSA methods:

1. **Feature-Based RSA**: Projects neural patterns into a predefined feature space
2. **Vector-Based RSA**: Directly compares vectorized distance matrices

This vignette explains when and why to use these specialized approaches instead of standard RSA.

# Feature-Based RSA

## Overview

Feature-Based RSA is designed for cases where:

1. You have a rich feature space describing your stimuli
2. You want to directly map neural patterns to these features
3. You're interested in reconstructing stimulus features from neural patterns

Unlike standard RSA which compares similarity matrices, Feature-Based RSA attempts to find a mapping between neural patterns and a feature space.

## Key Differences from Standard RSA

1. **Direct Feature Mapping**: 
   - Standard RSA: Compares similarity structures
   - Feature RSA: Maps neural patterns to feature dimensions

2. **Output**:
   - Standard RSA: Correlation/regression coefficients between RDMs
   - Feature RSA: Predicted feature values for each stimulus

3. **Use Cases**:
   - Standard RSA: Testing theoretical models of representation
   - Feature RSA: Reconstructing stimulus features from brain activity

## Implementation Example

Let's walk through a complete example:

```{r}
## Generate data and weakly-correlated features
# Synthetic dataset: 6x6x6, 50 observations (4 blocks)
data_out <- rMVPA::gen_sample_dataset(D = c(6,6,6), nobs = 50, blocks = 4, nlevels = 2)
print(data_out)

set.seed(123)
n_stimuli  <- 50
n_features <- 5

# Latent factors shared between features and brain data (weak signal)
Z <- matrix(rnorm(n_stimuli * 2), n_stimuli, 2)

# Build a weakly-correlated feature matrix: F = 0.5 * (Z %*% B) + 0.5 * noise
B <- matrix(rnorm(2 * n_features), 2, n_features)
feature_matrix <- 0.5 * (Z %*% B) + 0.5 * matrix(rnorm(n_stimuli * n_features), n_stimuli, n_features)
feature_matrix <- scale(feature_matrix)
colnames(feature_matrix) <- paste0("feature_", 1:n_features)

# Create stimulus labels
stim_labels <- paste0("stim_", 1:n_stimuli)

# Create a subtle latent signal in the brain data driven by the same Z
mask_vol <- data_out$dataset$mask
mask_idx <- which(mask_vol > 0)

# Extract masked voxel-by-time matrix from NeuroVec (V x T)
vol_list <- neuroim2::vols(data_out$dataset$train_data)
datamat  <- do.call(cbind, lapply(vol_list, function(v) as.numeric(v[mask_idx])))

# Two random spatial patterns within the mask
set.seed(124)
p1 <- rnorm(length(mask_idx))
p2 <- rnorm(length(mask_idx))
p1 <- p1 / sd(p1); p2 <- p2 / sd(p2)

# Inject weak signal into brain data: X <- X + 0.2 * (p1 %*% t(Z1)) + 0.2 * (p2 %*% t(Z2))
datamat <- datamat + p1 %*% t(0.2 * Z[,1]) + p2 %*% t(0.2 * Z[,2])

# Repackage as a NeuroVec using the existing space and mask
train_vec <- neuroim2::SparseNeuroVec(datamat, neuroim2::space(data_out$dataset$train_data), mask = mask_vol)
dset <- mvpa_dataset(train_vec, mask = mask_vol)

# Create feature RSA design from the correlated feature matrix
feature_design <- feature_rsa_design(
  F = feature_matrix,
  labels = stim_labels
)

# Create cross-validation structure using the block information
crossval <- blocked_cross_validation(data_out$design$block_var)

# Create feature RSA model
feature_model <- feature_rsa_model(
  dataset = dset,
  design = feature_design,
  method = "pls",  # Partial Least Squares
  crossval = crossval  # Add cross-validation
)

# Create proper region mask from the dataset's mask
mask_vol <- data_out$dataset$mask
nvox <- sum(mask_vol)
region_mask <- neuroim2::NeuroVol(
  sample(1:3, size = nvox, replace = TRUE),  # 3 regions
  space(mask_vol),
  indices = which(mask_vol > 0)
)

# Run regional analysis
results <- run_regional(feature_model, region_mask)

# Examine results
print(results$performance_table)
```

## Available Methods

Feature-Based RSA supports multiple analysis methods:

1. **Partial Least Squares (PLS)**:
   - Finds latent components that maximize covariance between neural patterns and features
   - Good for prediction when features are highly correlated

2. **Principal Component Analysis (PCA)**:
   - Uses PCA on features followed by regression
   - Good for dimensionality reduction with clear interpretability

3. **GLMNet (Elastic Net)**:
   - Elastic net regression with L1 and L2 regularization
   - Useful for feature selection and handling correlated features

```{r}
# Compare different methods
methods <- c("pls", "pca", "glmnet")
results_list <- lapply(methods, function(method) {
  model <- feature_rsa_model(
    dataset = dset,
    design = feature_design,
    method = method,
    crossval = crossval  # Add cross-validation
  )
  run_regional(model, region_mask)
})

# Compare performance
for (i in seq_along(methods)) {
  cat("\nMethod:", methods[i], "\n")
  print(results_list[[i]]$performance_table)
}
```

# Vector-Based RSA

## Overview

Vector-Based RSA is designed for cases where:

1. You have pre-computed distance matrices
2. You want to compare across-block distances only
3. You need efficient computation for large datasets

Unlike standard RSA which works with full similarity matrices, Vector-Based RSA operates on vectorized distance matrices and can efficiently handle block structure.

## Key Differences from Standard RSA

1. **Computation**:
   - Standard RSA: Works with full similarity matrices
   - Vector RSA: Works with vectorized distances

2. **Block Handling**:
   - Standard RSA: Manual block exclusion
   - Vector RSA: Built-in efficient block handling

3. **Memory Usage**:
   - Standard RSA: Stores full matrices
   - Vector RSA: Stores only necessary comparisons

## Implementation Example

```{r}
# Create distance matrix for stimuli
stim_distances <- as.matrix(dist(feature_matrix))
rownames(stim_distances) <- stim_labels

# Create block structure (e.g., runs)
blocks <- rep(1:5, length.out = n_stimuli)

# Create vector RSA design
vector_design <- vector_rsa_design(
    D = stim_distances,
    labels = stim_labels,
    block_var = blocks
)

# Create vector RSA model
vector_model <- vector_rsa_model(
  dataset = dset,
  design = vector_design,
  distfun = cordist(),  # Correlation distance
  rsa_simfun = "pearson"
)

# Run analysis
results_vector <- run_regional(vector_model, region_mask)

# Examine results
print(results_vector$performance_table)
```

## Efficient Block Handling

Vector-Based RSA automatically handles block structure:

```{r}
# Compare with different block structures
block_sizes <- c(5, 10)
results_blocks <- lapply(block_sizes, function(size) {
  blocks <- rep(1:(n_stimuli/size), each = size)
  design <- vector_rsa_design(
    D = stim_distances,
    labels = stim_labels,
    block_var = blocks
  )
  model <- vector_rsa_model(
    dataset = dset,
    design = design,
    distfun = cordist()
  )
  run_regional(model, region_mask)
})

# Compare results
for (i in seq_along(block_sizes)) {
  cat("\nBlock size:", block_sizes[i], "\n")
  print(results_blocks[[i]]$performance_table)
}
```

# When to Use Each Method

## Feature-Based RSA

Feature-Based RSA is most appropriate when working with a well-defined feature space where the dimensions have meaningful interpretations. This approach excels at reconstructing and predicting specific stimulus features from neural activity patterns. Common applications include predicting visual features from visual cortex responses, reconstructing semantic dimensions from language area activity, and mapping motion parameters from MT/V5 activation patterns.

## Vector-Based RSA

Vector-Based RSA is ideal for analyzing large datasets with pre-computed distance matrices. It handles memory efficiently and provides sophisticated block handling capabilities. This makes it particularly valuable when comparing across multiple experimental runs, working with high-resolution fMRI data, or analyzing large-scale similarity structures.

## Standard RSA

Stick with standard RSA when:
- You're testing theoretical models
- You want to compare similarity structures
- Block structure isn't critical
- Memory usage isn't a concern

# Summary

The `rMVPA` package provides three complementary RSA approaches:

1. **Standard RSA**: General-purpose similarity analysis
2. **Feature-Based RSA**: Direct feature mapping and reconstruction
3. **Vector-Based RSA**: Efficient similarity comparison with block handling

Your choice of RSA method should align with your research goals, match your data's structure, fit within computational constraints, and reflect whether you're analyzing feature spaces or similarity patterns.

For implementation details, refer to:
- `feature_rsa_model.R`
- `vector_rsa_model.R`
- `rsa_model.R` 


