---
title: "Advanced RSA Methods: Feature-Based and Vector-Based Approaches"
author: "Bradley Buchsbaum"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Advanced RSA Methods: Feature-Based and Vector-Based Approaches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, message=FALSE}
suppressPackageStartupMessages({
  library(neuroim2)
  library(rMVPA)
  library(dplyr)
})
```

# Introduction

While standard Representational Similarity Analysis (RSA) compares neural patterns with model-based similarity structures, there are cases where more specialized approaches are needed. The `rMVPA` package provides two advanced RSA methods:

1. **Feature-Based RSA**: Projects neural patterns into a predefined feature space
2. **Vector-Based RSA**: Directly compares vectorized distance matrices

This vignette explains when and why to use these specialized approaches instead of standard RSA.

# Feature-Based RSA

## Overview

Feature-Based RSA is designed for cases where:

1. You have a rich feature space describing your stimuli
2. You want to directly map neural patterns to these features
3. You're interested in reconstructing stimulus features from neural patterns

Unlike standard RSA which compares similarity matrices, Feature-Based RSA attempts to find a mapping between neural patterns and a feature space.

## Key Differences from Standard RSA

1. **Direct Feature Mapping**: 
   - Standard RSA: Compares similarity structures
   - Feature RSA: Maps neural patterns to feature dimensions

2. **Output**:
   - Standard RSA: Correlation/regression coefficients between RDMs
   - Feature RSA: Predicted feature values for each stimulus

3. **Use Cases**:
   - Standard RSA: Testing theoretical models of representation
   - Feature RSA: Reconstructing stimulus features from brain activity

## Implementation Example

Let's walk through a complete example:

```{r}
# Generate a synthetic dataset with dimensions 6x6x6, 50 observations, divided into 4 blocks
# This ensures that the number of trials matches the number of stimuli used below
data_out <- rMVPA::gen_sample_dataset(D = c(6,6,6), nobs = 50, blocks = 4, nlevels = 2)
print(data_out)

# Generate synthetic feature space (e.g., visual features of stimuli)
set.seed(123)
n_stimuli <- 50
n_features <- 5

# Create feature matrix (stimuli × features)
feature_matrix <- matrix(rnorm(n_stimuli * n_features), n_stimuli, n_features)
colnames(feature_matrix) <- paste0("feature_", 1:n_features)

# Create stimulus labels
stim_labels <- paste0("stim_", 1:n_stimuli)

# Create feature RSA design
feature_design <- feature_rsa_design(
  F = feature_matrix,  # Direct feature matrix
  labels = stim_labels
)

# Create MVPA dataset from the generated data
dset <- mvpa_dataset(data_out$dataset$train_data, mask = data_out$dataset$mask)

# Create cross-validation structure using the block information
crossval <- blocked_cross_validation(data_out$design$block_var)

# Create feature RSA model
feature_model <- feature_rsa_model(
  dataset = dset,
  design = feature_design,
  method = "scca",  # Sparse Canonical Correlation Analysis
  crossval = crossval  # Add cross-validation
)

# Create proper region mask from the dataset's mask
mask_vol <- data_out$dataset$mask
nvox <- sum(mask_vol)
region_mask <- neuroim2::NeuroVol(
  sample(1:3, size = nvox, replace = TRUE),  # 3 regions
  space(mask_vol),
  indices = which(mask_vol > 0)
)

# Run regional analysis
results <- run_regional(feature_model, region_mask)

# Examine results
print(results$performance_table)
```

## Available Methods

Feature-Based RSA supports multiple analysis methods:

1. **Sparse Canonical Correlation Analysis (SCCA)**:
   - Finds canonical correlations between neural patterns and features
   - Useful when features may be correlated

2. **Partial Least Squares (PLS)**:
   - Projects data to maximize covariance with features
   - Good for high-dimensional feature spaces

3. **PCA Regression**:
   - Reduces neural patterns via PCA then predicts features
   - Useful for very high-dimensional neural data

```{r}
# Compare different methods
methods <- c("scca", "pca") ## pls is not implemented yet
results_list <- lapply(methods, function(method) {
  model <- feature_rsa_model(
    dataset = dset,
    design = feature_design,
    method = method,
    crossval = crossval  # Add cross-validation
  )
  run_regional(model, region_mask)
})

# Compare performance
for (i in seq_along(methods)) {
  cat("\nMethod:", methods[i], "\n")
  print(results_list[[i]]$performance_table)
}
```

# Vector-Based RSA

## Overview

Vector-Based RSA is designed for cases where:

1. You have pre-computed distance matrices
2. You want to compare across-block distances only
3. You need efficient computation for large datasets

Unlike standard RSA which works with full similarity matrices, Vector-Based RSA operates on vectorized distance matrices and can efficiently handle block structure.

## Key Differences from Standard RSA

1. **Computation**:
   - Standard RSA: Works with full similarity matrices
   - Vector RSA: Works with vectorized distances

2. **Block Handling**:
   - Standard RSA: Manual block exclusion
   - Vector RSA: Built-in efficient block handling

3. **Memory Usage**:
   - Standard RSA: Stores full matrices
   - Vector RSA: Stores only necessary comparisons

## Implementation Example

```{r}
# Create distance matrix for stimuli
stim_distances <- as.matrix(dist(feature_matrix))
rownames(stim_distances) <- stim_labels

# Create block structure (e.g., runs)
blocks <- rep(1:5, length.out = n_stimuli)

# Create vector RSA design
vector_design <- vector_rsa_design(
    D = stim_distances,
    labels = stim_labels,
    block_var = blocks
)

# Create vector RSA model
vector_model <- vector_rsa_model(
  dataset = dset,
  design = vector_design,
  distfun = cordist(),  # Correlation distance
  rsa_simfun = "pearson"
)

# Run analysis
results_vector <- run_regional(vector_model, region_mask)

# Examine results
print(results_vector$performance_table)
```

## Efficient Block Handling

Vector-Based RSA automatically handles block structure:

```{r}
# Compare with different block structures
block_sizes <- c(5, 10)
results_blocks <- lapply(block_sizes, function(size) {
  blocks <- rep(1:(n_stimuli/size), each = size)
  design <- vector_rsa_design(
    D = stim_distances,
    labels = stim_labels,
    block_var = blocks
  )
  model <- vector_rsa_model(
    dataset = dset,
    design = design,
    distfun = cordist()
  )
  run_regional(model, region_mask)
})

# Compare results
for (i in seq_along(block_sizes)) {
  cat("\nBlock size:", block_sizes[i], "\n")
  print(results_blocks[[i]]$performance_table)
}
```

# When to Use Each Method

## Feature-Based RSA

Choose Feature-Based RSA when:
- You have a well-defined feature space
- You want to reconstruct stimulus features
- Your features have meaningful dimensions
- You're interested in feature prediction

Example use cases:
- Predicting visual features from visual cortex activity
- Reconstructing semantic dimensions from language areas
- Mapping motion parameters from MT/V5 responses

## Vector-Based RSA

Choose Vector-Based RSA when:
- You have many pre-computed distance matrices
- Memory efficiency is important
- You need sophisticated block handling
- You're working with large datasets

Example use cases:
- Comparing across many experimental runs
- Working with high-resolution fMRI data
- Analyzing large-scale similarity structures

## Standard RSA

Stick with standard RSA when:
- You're testing theoretical models
- You want to compare similarity structures
- Block structure isn't critical
- Memory usage isn't a concern

# Summary

The `rMVPA` package provides three complementary RSA approaches:

1. **Standard RSA**: General-purpose similarity analysis
2. **Feature-Based RSA**: Direct feature mapping and reconstruction
3. **Vector-Based RSA**: Efficient similarity comparison with block handling

Choose the appropriate method based on:
- Your research question
- Data structure
- Computational requirements
- Whether you're interested in features vs. similarities

For implementation details, refer to:
- `feature_rsa_model.R`
- `vector_rsa_model.R`
- `rsa_model.R` 