---
title: 'Fit a Blocked Voxelwise Banded-Ridge Encoding Model'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: lapis
  preset: study
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts
vignette: |
  %\VignetteIndexEntry{Fit a Blocked Voxelwise Banded-Ridge Encoding Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 4.5,
  fig.align = "center"
)

library(rMVPA)
library(neuroim2)
set.seed(20260828)
```

```{r albers-classes, echo = FALSE, results = 'asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

You have several time-aligned feature spaces---for example, visual motion,
audio, and semantic features---and one BOLD response at every voxel. You want
to predict each voxel on observations the model never saw, while allowing
different feature spaces and different voxels to need different amounts of
regularization. Randomly splitting a time series would leak temporal context,
so the validation units must remain intact.

This guide builds that analysis with three contiguous stimulus segments. By
the end, you will have voxelwise outer out-of-fold predictions, MSE,
correlation, R2, the alpha and feature-space weights selected inside each outer
training set, and optional predictive leave-one-band-out delta R2 maps.

```{r example-builder, include = FALSE}
build_encoding_example <- function(seed = 20260828L) {
  set.seed(seed)
  n <- 36L
  dims <- c(3L, 3L, 2L)
  blocks_train <- rep(1:3, each = 12L)
  X <- matrix(rnorm(n * 8L), n, 8L)
  colnames(X) <- c(paste0("visual_", 1:4), paste0("semantic_", 1:4))
  beta <- matrix(0, 8L, prod(dims))
  beta[1:4, 1:9] <- matrix(rnorm(4L * 9L, sd = 0.7), 4L, 9L)
  beta[5:8, 10:15] <- matrix(rnorm(4L * 6L, sd = 0.7), 4L, 6L)
  beta[, 16:18] <- matrix(rnorm(8L * 3L, sd = 0.2), 8L, 3L)
  Y <- X %*% beta + matrix(rnorm(n * prod(dims), sd = 0.6), n, prod(dims))
  data_array <- array(as.vector(t(Y)), c(dims, n))
  dataset <- mvpa_dataset(
    NeuroVec(data_array, NeuroSpace(c(dims, n), c(1, 1, 1))),
    mask = NeuroVol(array(1, dims), NeuroSpace(dims, c(1, 1, 1)))
  )
  list(X = X, Y = Y, blocks = blocks_train, dataset = dataset)
}

example <- build_encoding_example()
```

## What are the inputs?

Rows of the predictor matrix must align exactly with volumes in the image
series. `blocks()` receives **numbers of columns**, not column indices. The
design owns both feature-band membership and the training block metadata, so
the model cannot receive a second, conflicting band specification.

```{r construct-design}
features <- feature_sets(
  example$X,
  blocks(visual = 4, semantic = 4)
)

design <- feature_sets_design(
  features,
  block_var_train = example$blocks,
  time_series = TRUE
)
```

The example has 36 observations, eight predictors in two named bands, three
independent temporal segments, and 18 active voxels.

```{r input-overview, echo = FALSE}
input_overview <- data.frame(
  observations = nrow(example$X),
  predictors = ncol(example$X),
  feature_bands = length(features$set_order),
  temporal_blocks = length(unique(example$blocks)),
  active_voxels = sum(example$dataset$mask > 0)
)
knitr::kable(input_overview, caption = "Aligned inputs for the worked example.")
```

## How do you fit the model safely?

The outer folds estimate generalization. The inner folds choose
hyperparameters using only the corresponding outer-training rows. Because the
design declares a time series, rMVPA refuses an unsafe random split unless you
provide intact blocks or an explicit temporal purge.

```{r fit-model}
model <- banded_ridge_model(
  example$dataset, design,
  outer_crossval = 3, tune_crossval = 2,
  theta_method = "grid", theta_grid_points = 3,
  target_batch_size = 7, return_predictions = TRUE,
  delta_sets = "all"
)

result <- run_banded_ridge(model)
```

`alphas` is left at its default, `"auto"`, which places nine penalties across
six decades around the scale at which the penalty starts to change this
design's fit. That scale is a property of the design, not a constant: the
solver standardizes every training column and scales band `b` by
`sqrt(theta_b)`, so alpha competes with cross-product eigenvalues that average
`(n - 1) * mean(D_b) / min(n - 1, p)`. It grows with the number of rows and
with band width, which is why a design with a few thousand timepoints needs
penalties orders of magnitude larger than one with a few dozen.

```{r resolved-alphas}
signif(model$alpha_grid, 3)
```

`outer_crossval` also accepts a `cross_validation` object, so
`blocked_cross_validation(example$blocks)` selects the same block-wise outer
folds used elsewhere in rMVPA. Integer counts and `cross_validation` objects
have `purge` applied to their training rows, whereas a hand-built fold list is
validated as supplied. `tune_crossval` can be omitted: the default uses at
most five inner folds, limited by the blocks (or rows, when none are declared)
available inside each outer-training set, and inner tuning is skipped entirely
when only one
candidate survives the alpha/theta scopes.

```{r validate-model, include = FALSE}
stopifnot(
  inherits(model, "banded_ridge_model"),
  inherits(result, "banded_ridge_result"),
  identical(dim(result$predictions), c(36L, 18L)),
  identical(result$active_response_ids, 1:18),
  nrow(result$metrics) == 18L,
  nrow(result$hyperparameters) == 18L * 3L,
  all(is.finite(result$metrics$mse)),
  stats::sd(result$metrics$correlation) > 0,
  identical(result$provenance$conceptual_model_count, 3L),
  identical(result$provenance$outer_fold_ids,
            names(result$predictive_leave_one_band_out$
                    provenance$matched_outer_test_rows)),
  setequal(
    names(result$predictive_leave_one_band_out$maps),
    c("delta_cv_r2_visual", "delta_cv_r2_semantic")
  ),
  any(abs(result$predictive_leave_one_band_out$effects[, 3:4]) > 1e-6)
)
```

The compact result table is already in original voxel order. These are pooled
outer out-of-fold metrics: every observation contributes once, from the outer
fold in which it was held out.

```{r inspect-metrics}
head(result$metrics)
```

The hyperparameter table is deliberately fold-specific. A single averaged
alpha or theta map is convenient for visualization, but this table is the
exact record of what generated each fold's predictions.

```{r inspect-hyperparameters}
head(result$hyperparameters[, c(
  "voxel_index", "outer_fold", "alpha",
  "theta_visual", "theta_semantic", "inner_score"
)])
```

```{r plot-summary, echo = FALSE}
plot_encoding_summary <- function(result) {
  old_par <- graphics::par(no.readonly = TRUE)
  on.exit(graphics::par(old_par), add = TRUE)
  graphics::par(mfrow = c(1, 2), mar = c(4, 4, 2, 0.5))
  graphics::plot(
    sort(result$metrics$correlation), pch = 19,
    xlab = "Voxel rank", ylab = "Outer-OOF correlation",
    main = "Predictive performance"
  )
  graphics::abline(h = 0, lty = 2, col = "grey50")
  effects <- result$predictive_leave_one_band_out$effects[, 3:4]
  graphics::boxplot(
    effects, names = c("visual", "semantic"),
    ylab = expression(Delta * R^2),
    main = "Predictive band drop-out"
  )
  graphics::abline(h = 0, lty = 2, col = "grey50")
}

plot_encoding_summary(result)
```

Positive and negative values are both meaningful. A negative delta means the
independently retuned reduced model predicted better in these finite samples;
the value is never clipped.

## Did the tuning grid contain the optimum?

A tuning grid that stops short of the optimum does not fail. Every response
simply takes the largest penalty on offer, the refit is under-penalized, and
the maps and delta R2 values come back fully populated and wrong. Read
`result$selection_diagnostics` before anything else.

```{r inspect-selection-diagnostics}
result$selection_diagnostics$alpha[, c(
  "model", "modal_alpha", "share_at_grid_min", "share_at_grid_max",
  "share_interior", "saturated"
)]
```

The signature of a usable grid is an **interior** modal selection: most
responses chose an alpha that is neither the smallest nor the largest
available, so the grid brackets the optimum rather than truncating it.
`saturated` is set when 95% of a model's selections take one end of the grid,
or when the two ends hold 95% between them, and `run_banded_ridge()` warns.
Both forms matter. Alpha is tuned per response by default, so a mask that
mixes predictable voxels with unpredictable ones sends its boundary mass to
opposite ends --- the noise voxels want more penalty than the ceiling offers,
the signal voxels less than the floor --- and neither end alone need look
alarming while almost nothing lands inside. Widen `alphas` until the modal
selection moves inside.

`$fit` answers the second question, which is independent of the grid: did the
model predict at all?

```{r inspect-fit-diagnostics}
result$selection_diagnostics$fit
```

A median outer out-of-fold R2 below `-0.05` means the fit predicts worse than
the mean of the data it was scored on, and `run_banded_ridge()` warns about
that separately.
Nothing downstream of such a fit describes explained variance --- in
particular, the leave-one-band-out deltas are then differences between two
models that both fail.

## Which loss and which score are you reading?

Three quantities answer different questions:

1. `inner_score` is the selection loss. The default is MSE computed from inner
   out-of-fold predictions within one outer-training set. Correlation and R2
   are explicit opt-in selection criteria.
2. `mse`, `correlation`, and `r2` in `result$metrics` are descriptive scores
   computed after concatenating the outer out-of-fold predictions. They are
   not the inner selection scores.
3. `delta_cv_r2_<band>` is `R2_full - R2_without_band`, using the same outer
   test rows for both models. Each reduced model is tuned again. This is a
   predictive leave-one-band-out effect, not an additive unique/shared
   variance decomposition, so effects need not sum to the full-model R2.

The fitted objective for a fixed candidate is

\[
  \widehat B = \arg\min_B
  \lVert Y - XB \rVert_F^2 +
  \sum_g \frac{\alpha}{\theta_g}\lVert B_g \rVert_F^2,
\]

where non-negative theta values sum to one. `theta_g = 0` excludes a band
exactly. The squared-error term is not divided by the number of observations,
so alpha is on the scale of the training cross-product. Centering and scaling
are learned separately inside every training split. Equal theta values recover
ordinary ridge after accounting for scale: with `G` bands and ordinary ridge
penalty `lambda`, use `theta_g = 1/G` and `alpha = lambda/G`.

## What is retained, and what can become large?

The constructor resolves the solver and estimates retained output storage
before fitting. The default keeps metrics and hyperparameters but not the full
prediction matrix, coefficients, or nested diagnostics. Request those outputs
only when you need them.

```{r allocation-plan, echo = FALSE}
allocation <- data.frame(
  chosen_solver = model$solver,
  reason = model$solver_plan$reason,
  predictions_MiB = model$retention_estimated_bytes[["predictions"]] / 1024^2,
  attribution_MiB = model$retention_estimated_bytes[["attribution"]] / 1024^2,
  total_retained_MiB = model$retention_estimated_bytes[["total"]] / 1024^2
)
knitr::kable(allocation, digits = 4,
             caption = "Decision and retained-storage estimate available before fitting.")
```

Use `weight_retention = "primal"` for outer-fold coefficient matrices or
`"dual"` for outer-fold sample-space weights. `max_retained_mb` refuses an
unsafe request before allocation; `weight_overflow = "none"` may explicitly
fall back to no weights. `memory_limit_mb` independently guards solver
intermediates and caps the retained decomposition cache: band Gram matrices are
cached once per training split and shared across theta candidates and drop-out
models, and least-recently-used entries are evicted (and recomputed on demand)
once the cap is reached, so the limit changes reuse but never results. Runtime
provenance records the chosen solver, chunk manifest, decomposition, band-Gram
reuse, and eviction counts, maximum chunk result size, and the B+1 cost of
requested drop-out models.

For `F` outer folds, `K` inner folds, `T` theta candidates, `A` alphas, and
response chunks of size `B`, tuning evaluates approximately `F * K * T * A`
candidate fits per conceptual model and then performs outer refits. Optimized
paths reuse a decomposition across alpha values only when the training rows
and theta are unchanged. Changing theta changes the scaled design/kernel and
requires another decomposition; there is no claim of one decomposition for
the entire search.

## What did the issue #70 simulation show when rerun?

Issue #70 reported a large relative gain from per-voxel penalties in one
simulation. The reproducibility script installed at
`benchmarks/banded_ridge_issue70.R` defines a concrete successor: two bands,
40 voxels at SNR 1.0/0.4/0.0, six contiguous outer segments, nested tuning, and
30 declared seeds. It compares shared alpha/theta with response-specific
alpha/theta and reports the **absolute** OOF-correlation change.

```{r issue70-load, include = FALSE}
resolve_inst_file <- function(directory, filename) {
  installed <- system.file(directory, filename, package = "rMVPA")
  if (nzchar(installed) && file.exists(installed)) return(installed)
  candidates <- c(
    file.path("..", "inst", directory, filename),
    file.path("inst", directory, filename)
  )
  found <- candidates[file.exists(candidates)]
  if (!length(found)) stop("Cannot locate installed certification artifact.")
  normalizePath(found[[1L]])
}

issue70 <- utils::read.csv(resolve_inst_file(
  "extdata", "banded_ridge_issue70_results.csv"
))
issue70_by_rep <- aggregate(
  cbind(shared_oof_correlation, response_oof_correlation,
        absolute_correlation_gain) ~ seed + snr,
  issue70, mean
)
issue70_summary <- do.call(rbind, lapply(
  split(issue70_by_rep, issue70_by_rep$snr),
  function(x) data.frame(
    SNR = x$snr[[1L]],
    shared_r = mean(x$shared_oof_correlation),
    response_specific_r = mean(x$response_oof_correlation),
    absolute_gain = mean(x$absolute_correlation_gain),
    gain_q025 = unname(stats::quantile(x$absolute_correlation_gain, 0.025)),
    gain_q975 = unname(stats::quantile(x$absolute_correlation_gain, 0.975))
  )
))

# Re-run the first declared seed and compare with the bundled receipt. This
# prevents the rendered table from silently surviving script/API drift.
benchmark_environment <- new.env(parent = globalenv())
sys.source(resolve_inst_file("benchmarks", "banded_ridge_issue70.R"),
           envir = benchmark_environment)
first_seed <- min(issue70$seed)
rechecked <- benchmark_environment$issue70_one_replication(first_seed)
stored_first <- issue70[issue70$seed == first_seed, ]
stored_first <- stored_first[match(rechecked$response, stored_first$response), ]
stopifnot(
  length(unique(issue70$seed)) == 30L,
  identical(sort(unique(issue70$snr)), c(0, 0.4, 1)),
  max(abs(rechecked$shared_oof_correlation -
            stored_first$shared_oof_correlation)) < 1e-12,
  max(abs(rechecked$response_oof_correlation -
            stored_first$response_oof_correlation)) < 1e-12,
  identical(rechecked$fold_manifest, stored_first$fold_manifest),
  identical(rechecked$candidate_manifest, stored_first$candidate_manifest)
)
```

```{r issue70-summary, echo = FALSE}
knitr::kable(
  issue70_summary, digits = 3,
  caption = paste0(
    "Issue #70 successor simulation. Intervals are empirical 2.5% and 97.5% ",
    "quantiles of replication-level mean absolute gains."
  )
)
```

This registered variant did **not** reproduce the earlier 78% claim. At SNR
0.4, response-specific tuning changed OOF correlation by about -0.048 in
absolute units (replication interval about -0.069 to -0.030). The result is
specific to this data-generating process, MSE selection rule, and candidate
grid; it is neither evidence that shared tuning is universally better nor a
correctness test. Across pure-noise voxels, the replication-level mean
response-specific correlation was close to zero (about -0.023; empirical 95%
interval about -0.092 to 0.013). The direct/SVD/dual oracles, leakage
mutations, and attribution adversaries provide correctness evidence instead.

## How does this relate to other ridge tools?

Choose by estimand and execution environment, then compare on identical folds,
preprocessing, candidate sets, and response scoring.

| Tool | Useful when | Important comparison boundary |
|---|---|---|
| Ordinary ridge | One common penalty across all feature columns is the scientific model | In rMVPA, use fixed equal theta and convert ordinary `lambda` to `alpha = lambda/G` |
| [glmnet](https://cran.r-project.org/package=glmnet) | You need elastic-net/lasso paths or its established sparse-matrix tooling | `mgaussian` is a multi-task group penalty across responses; its objective scaling and standardization conventions differ |
| [multiridge](https://cran.r-project.org/package=multiridge) | You want its multi-penalty CV or marginal-likelihood workflows for high-dimensional data blocks | Its optimization, preprocessing, and response model are not assumed to be numerically identical to this nested voxelwise estimand |
| [himalaya](https://gallantlab.org/himalaya/) | You want Python, CPU/GPU backends, or multiple-kernel/group-ridge solvers at very large target counts | It is an external parity target, not an rMVPA dependency; align folds, centering, alpha/theta parameterization, and precision before comparing |

The bundled fixed-shape performance receipt likewise characterizes, rather
than proves superiority. On the recorded Apple M3 Max / R 4.5.1 run, the
`n = 60, p = 512, V = 12` public workflow took median 0.226 seconds with the
direct backend and 0.099 seconds with each optimized backend over five runs.
The solver planner estimated 2.15 MB, 0.33 MB, and 0.17 MB of intermediate
storage respectively. Wall times are tracked as snapshots; CI enforces
structural shape, batching, allocation, cache, and numerical-parity contracts
instead of brittle timing cutoffs.

```{r performance-receipt, include = FALSE}
performance <- utils::read.csv(resolve_inst_file(
  "extdata", "banded_ridge_performance_results.csv"
))
stopifnot(
  setequal(performance$shape, c("small", "wide")),
  setequal(performance$solver, c("direct", "svd_primal", "dual_kernel")),
  all(performance$elapsed_median_seconds > 0),
  all(performance$estimated_solver_intermediate_peak_bytes > 0),
  max(performance$max_abs_prediction_error_vs_direct) < 1e-10
)
```

## Where should you go next?

Use `?banded_ridge_model` for the complete objective, candidate, storage, and
failure contracts; `?run_banded_ridge` for the result schema and regional
batching rules; and `vignette("CrossValidation")` for broader validation
design. For temporally ordered data, decide the independent blocks or purge
from the acquisition and stimulus design before choosing a solver.
