---
title: Constructing Datasets for MVPA Analysis
author: Bradley Buchsbaum
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 3
vignette: |
  %\VignetteIndexEntry{Constructing Datasets for MVPA Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-teal');});</script>

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(albersdown::theme_albers(params$family))
}
library(neuroim2)
library(rMVPA)
```

This vignette shows how to construct neuroimaging datasets for MVPA. We start with a runnable synthetic example, then show the pattern for real volumetric and surface data.

## Quick Start with Synthetic Data

The easiest way to get started is with `gen_sample_dataset()`, which creates a complete dataset and design in one call:

```{r synthetic-example}
ds <- gen_sample_dataset(D = c(6, 6, 6), nobs = 80, blocks = 4, nlevels = 2)
print(ds$dataset)
print(ds$design)
```

The dataset bundles a 4-D `NeuroVec` (voxels × time) and a binary `NeuroVol` mask. Two pictures make the shape concrete: a mid-axial slice of one trial's voxel pattern, and the trial × run grid of the design.

```{r synthetic-visual, echo = FALSE, fig.cap = "Left: a mid-axial slice of one trial's voxel pattern (signed activation, blue↔red). The synthetic mask covers the full 6x6x6 volume here; in real data the mask would be a brain envelope. Right: each trial coloured by run; runs split into equal-length blocks of trials.", fig.width = 7, fig.height = 3.4}
op <- par(mfrow = c(1, 2), mar = c(2.5, 2.5, 1.5, 0.5))
trial_vol <- as.array(neuroim2::vols(ds$dataset$train_data)[[1]])
mid_z <- ceiling(dim(trial_vol)[3] / 2)
sl <- trial_vol[, , mid_z]
zr <- max(abs(sl), na.rm = TRUE)
image(seq_len(nrow(sl)), seq_len(ncol(sl)), sl,
      col = hcl.colors(48, "Blue-Red 3"),
      zlim = c(-zr, zr),
      axes = FALSE, xlab = "", ylab = "", main = "trial 1, mid slice")
box()
plot(seq_along(ds$design$block_var), ds$design$block_var,
     pch = 16, cex = 0.7,
     col = scales::alpha(hcl.colors(length(unique(ds$design$block_var)),
                                     "Set 2"), 0.9)[ds$design$block_var],
     xlab = "trial index", ylab = "run", axes = FALSE,
     main = "design")
axis(1); axis(2, at = seq_along(unique(ds$design$block_var)))
par(op)
```

You can also assemble the pieces manually if you want explicit control over the design data frame:

```{r synthetic-manual}
dset <- mvpa_dataset(ds$dataset$train_data, mask = ds$dataset$mask)

design_df <- data.frame(
  Y     = ds$design$y_train,
  block = ds$design$block_var
)
mvdes <- mvpa_design(design_df, y_train = ~ Y, block_var = ~ block)
mvdes
```

## Creating a Real Volumetric (Image-Based) Dataset

The examples below use `eval = FALSE` because they reference file paths you should replace with your own data.

This example assumes you have a 4D fMRI file ("bold.nii.gz") and a corresponding 3D brain mask file ("mask.nii.gz").

```{r volumetric, eval=FALSE}
library(neuroim2)

# Read the fMRI data as a NeuroVec object using neuroim2::read_vec
train_neurovec <- neuroim2::read_vec("path/to/bold.nii.gz", mode = "normal")

# Read the brain mask and create a NeuroVol object
mask_vec <- neuroim2::read_vec("path/to/mask.nii.gz", mode = "normal")
mask_vol <- NeuroVol(as.array(mask_vec), NeuroSpace(dim(mask_vec), spacing = c(2, 2, 2)))

# Create the MVPA image dataset
real_dataset <- mvpa_dataset(train_data = train_neurovec, mask = mask_vol)

# Display dataset details
print(real_dataset)
```

## Creating a Real Surface-Based Dataset

This example assumes you have cortical geometry stored in a file (e.g., "subject.lh.smoothwm.asc") and a signal matrix in CSV format ("surface_data.csv"). The signal matrix should have dimensions corresponding to the number of vertices and the number of observations.

```{r surface, eval=FALSE}
## remotes::install_github("bbuchsbaum/neurosurf")
library(neurosurf)

# Load the cortical geometry
geom <- read_surf_geometry("path/to/subject.lh.smoothwm.asc")

# Read the surface data from a CSV file
# The CSV should not have a header and have dimensions: number of vertices x number of observations
data_matrix <- as.matrix(read.csv("path/to/surface_data.csv", header = FALSE))

# Verify that the number of rows in the data matches the geometry
nvert <- nrow(neurosurf::vertices(geom))
if(nrow(data_matrix) != nvert) {
  stop("The number of vertices in the data does not match the geometry.")
}

# Create a NeuroSurfaceVector using the geometry and the data matrix
real_neurosurf <- NeuroSurfaceVector(geom, 1:nvert, data_matrix)

# Create the MVPA surface dataset; if no mask is provided, one is generated automatically
real_surface_dataset <- mvpa_surface_dataset(train_data = real_neurosurf, name = "lh")

# Display dataset details
print(real_surface_dataset)
```

## Notes

- For volumetric datasets, the NeuroSpace object is used to define dimensions and voxel spacing. 
- For surface datasets, ensure that the signal matrix and cortical geometry are compatible.

Replace the file paths with actual paths to your real data files before running the code. 
