---
title: "Running Custom Analyses with rMVPA"
author: "Your Name Here"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Running Custom Analyses with rMVPA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(neuroim2)
  library(rMVPA)
  library(dplyr)
  library(tibble)
})
# Ensure sequential plan for examples unless specified
future::plan(future::sequential) 
```

# Introduction

The `rMVPA` package provides powerful tools for standard MVPA and RSA analyses. However, researchers often need to perform custom calculations within specific brain regions (ROIs) or searchlight spheres that go beyond the built-in models. For example, you might want to:

*   Calculate specific univariate statistics within an ROI.
*   Compute a custom connectivity metric within a searchlight sphere.
*   Integrate a novel analysis method not currently implemented as a standard `rMVPA` model.
*   Perform a simpler analysis without the overhead of defining a full `mvpa_model` or `rsa_model` object.

To facilitate this flexibility, `rMVPA` offers two functions: `run_custom_regional` and `run_custom_searchlight`. These functions allow you to apply any R function you define to the data extracted from ROIs or searchlight spheres, leveraging `rMVPA`'s data handling, iteration, parallel processing, and error management capabilities.

This vignette explains the rationale behind these functions and provides practical examples of their use.

# Rationale: Why Use Custom Analysis Functions?

Using `run_custom_regional` and `run_custom_searchlight` offers several advantages:

1.  **Flexibility**: Apply *any* R function to ROI or searchlight data, enabling bespoke analyses tailored to specific research questions.
2.  **Simplicity**: For analyses that don't fit the standard `mvpa_model` structure (e.g., simple descriptive statistics), these functions provide a more direct interface than creating a custom S3 model class.
3.  **Integration**: Seamlessly integrate custom metrics or methods from other R packages into the `rMVPA` workflow.
4.  **Efficiency**: Benefit from `rMVPA`'s optimized iteration (`mvpa_iterate`) and parallel processing capabilities (`future` framework) for your custom analyses.
5.  **Robustness**: Leverage built-in error handling that catches errors within your custom function for specific ROIs/spheres without halting the entire analysis.

# Core Functions

## `run_custom_regional`

This function applies a custom analysis function to data within predefined regions of interest (ROIs).

**Usage:**

```R
run_custom_regional(
  dataset,      # mvpa_dataset or mvpa_surface_dataset
  region_mask,  # NeuroVol or NeuroSurface mask defining ROIs
  custom_func,  # Your R function
  ...,          # Optional args passed to mvpa_iterate
  .cores = 1,   # Number of cores for parallel processing
  .verbose = FALSE # Print progress messages?
)
```

**The `custom_func` for Regional Analysis:**

Your custom function must accept two arguments:

1.  `roi_data`: A matrix or tibble containing the data (samples x features) for the *current* ROI.
2.  `roi_info`: A list containing information about the ROI, including `id` (the ROI number from the mask) and `indices` (the feature indices within the `dataset` corresponding to this ROI).

The function *must* return either:
*   A **named list** of scalar values (e.g., `list(mean = m, sd = s)`).
*   A **single-row data frame or tibble** where each column is a scalar value.

The names of the list elements or columns will become columns in the final output table.

## `run_custom_searchlight`

This function applies a custom analysis function to data within moving searchlight spheres across the brain.

**Usage:**

```R
run_custom_searchlight(
  dataset,      # mvpa_dataset or mvpa_surface_dataset
  custom_func,  # Your R function
  radius,       # Searchlight radius (mm for volume, connections for surface)
  method = "standard", # "standard" or "randomized"
  niter = 100,  # Iterations for "randomized" method
  ...,          # Optional args passed to mvpa_iterate
  .cores = 1,   # Number of cores for parallel processing
  .verbose = FALSE # Print progress messages?
)
```

**The `custom_func` for Searchlight Analysis:**

Your custom function must accept two arguments:

1.  `sl_data`: A matrix or tibble containing the data (samples x features_in_sphere) for the *current* searchlight sphere.
2.  `sl_info`: A list containing information about the sphere, including `center_index` (the index of the center voxel/vertex) and `indices` (the indices of all features within the sphere).

Similar to the regional version, the function *must* return either:
*   A **named list** of scalar values.
*   A **single-row data frame or tibble** with scalar columns.

All successfully processed spheres must return the *same set* of named metrics. The results will be aggregated into brain maps (`NeuroVol` or `NeuroSurface`).

# End-to-End Example

Let's demonstrate both functions with a simple custom analysis: calculating the mean and standard deviation of the signal within each ROI and searchlight sphere.

## 1. Setup: Data and Custom Function

First, we generate a sample volumetric dataset and define our custom function.

```{r setup_example}
# Generate sample dataset (e.g., 10x10x10 volume, 50 observations, 2 blocks)
dset_info <- gen_sample_dataset(D = c(10, 10, 10), nobs = 50, blocks = 2, nlevels=2)
dataset_obj <- dset_info$dataset

# Define our custom analysis function
# It calculates mean, sd, and number of features (voxels)
calculate_roi_stats <- function(data, info) {
  # Inputs:
  # data: matrix (samples x features) for the current ROI/sphere
  # info: list with id/center_index and feature indices
  
  # Perform calculations
  mean_signal <- mean(data, na.rm = TRUE)
  sd_signal <- sd(data, na.rm = TRUE)
  num_features <- ncol(data)
  
  # Return results as a named list of scalars
  list(
    mean_signal = mean_signal,
    sd_signal = sd_signal,
    n_features = num_features
  )
}

# Define a version that might fail for small ROIs/spheres
calculate_roi_stats_robust <- function(data, info) {
    if (ncol(data) < 5) {
        stop("Too few features (< 5) in this region!")
    }
    # If enough features, proceed as normal
    mean_signal <- mean(data, na.rm = TRUE)
    sd_signal <- sd(data, na.rm = TRUE)
    num_features <- ncol(data)
    list(
        mean_signal = mean_signal,
        sd_signal = sd_signal,
        n_features = num_features
    )
}

```

## 2. Custom Regional Analysis

Now, we create a region mask and run `run_custom_regional`.

```{r regional_example}
# Create a region mask with 4 ROIs covering parts of the dataset mask
mask_vol <- dataset_obj$mask
mask_arr <- array(0, dim(mask_vol))
active_indices <- which(mask_vol > 0)
n_active <- length(active_indices)

# Assign active voxels roughly to 4 regions
set.seed(456) # for reproducibility
roi_assignments <- sample(1:4, size = n_active, replace = TRUE) 
# Make ROI 4 potentially small to test error handling
roi_assignments[sample(which(roi_assignments==4), size=round(sum(roi_assignments==4)*0.9))] <- sample(1:3, size=round(sum(roi_assignments==4)*0.9), replace=TRUE)


mask_arr[active_indices] <- roi_assignments
region_mask_vol <- NeuroVol(mask_arr, space(mask_vol))

# Run the custom regional analysis (sequentially first)
custom_regional_results <- run_custom_regional(
  dataset = dataset_obj,
  region_mask = region_mask_vol,
  custom_func = calculate_roi_stats,
  .cores = 1, 
  .verbose = FALSE
)

# Print the results table
print(custom_regional_results)

# Example with error handling (using the robust function and potential small ROI 4)
# Suppress expected warnings from the logger about the error
suppressWarnings({
    custom_regional_error_results <- run_custom_regional(
      dataset = dataset_obj,
      region_mask = region_mask_vol,
      custom_func = calculate_roi_stats_robust, # Function that might error
      .cores = 1, 
      .verbose = FALSE
    )
})

cat("\nResults with potential errors:\n")
print(custom_regional_error_results)

```

**Explanation:**

The output of `run_custom_regional` is a `tibble`. Each row corresponds to an ROI defined in `region_mask_vol`. The columns are:
*   `id`: The ROI identifier (the integer value from the mask).
*   Columns based on the named list returned by `custom_func` (`mean_signal`, `sd_signal`, `n_features`).
*   `error`: A logical flag indicating if an error occurred within `custom_func` for this ROI.
*   `error_message`: The error message if `error` is `TRUE`.

In the second example, if `calculate_roi_stats_robust` encountered an ROI with fewer than 5 voxels (potentially ROI 4), the corresponding row would show `error = TRUE`, the specific error message, and `NA` values for the metrics.

## 3. Custom Searchlight Analysis

Next, we run a searchlight analysis using the same custom function.

```{r searchlight_example, message=FALSE, warning=FALSE}
# Run the custom searchlight analysis (standard method)
# Use a moderate radius; set cores > 1 for parallel execution if available
# Note: Running in parallel might print messages about the future plan setting.
custom_searchlight_results <- run_custom_searchlight(
  dataset = dataset_obj,
  custom_func = calculate_roi_stats,
  radius = 4, # e.g., 4mm radius
  method = "standard",
  .cores = 2, # Use 2 cores if available
  .verbose = FALSE
)

# Print the results object summary
print(custom_searchlight_results)

# Access the results for a specific metric (e.g., mean_signal)
mean_signal_map_obj <- custom_searchlight_results$results$mean_signal

# The map itself is stored in the 'data' slot
mean_signal_map_vol <- mean_signal_map_obj$data
cat("\nClass of the mean_signal map:", class(mean_signal_map_vol), "\n")
cat("Dimensions of the map:", dim(mean_signal_map_vol), "\n")

# Print summary stats for the map
print(mean_signal_map_obj$summary_stats)

# You can plot the map using neuroim2's plotting functions (example commented out)
# plot(mean_signal_map_vol) 
```

**Explanation:**

The output of `run_custom_searchlight` is a `searchlight_result` object (a list). Key components include:
*   `results`: A named list where each element corresponds to a metric returned by `custom_func` (e.g., `results$mean_signal`, `results$sd_signal`).
*   Each metric element (e.g., `results$mean_signal`) is a `searchlight_performance` object containing:
    *   `$data`: A `NeuroVol` (or `NeuroSurface`) object holding the metric values mapped back to brain space.
    *   `$summary_stats`: Basic statistics (mean, sd, min, max) calculated across the map values.
    *   Other metadata like `$metric_name`, `$n_nonzero`.
*   `metrics`: Names of the metrics computed.
*   `n_voxels`: total voxels/vertices defined by the mask.
*   `active_voxels`: number of voxels/vertices with results.

The values in the output maps (`$data`) represent the result of your `custom_func` applied to the searchlight sphere centered at each voxel. If the "randomized" method is used, the map values represent the *average* metric across all spheres that included that voxel.

# Key Considerations

*   **`custom_func` Requirements**: Ensure your function strictly adheres to the input argument requirements (`roi_data`/`sl_data`, `roi_info`/`sl_info`) and returns a named list or single-row tibble of *scalar* values. Inconsistent return types or names across ROIs/spheres will cause errors during result aggregation.
*   **Parallel Processing**: Set `.cores > 1` to speed up analysis. The `future` package backend will be used. For more control, set the `future::plan()` *before* calling the `run_custom_*` function (e.g., `future::plan(future::multisession, workers = 4)`).
*   **Error Handling**: Errors occurring *within* your `custom_func` for a specific ROI or sphere are caught automatically. The analysis will continue, and the affected ROI/voxel will be marked with an error flag or contain `NA` in the output. Use `futile.logger::flog.warn` or `flog.error` inside your `custom_func` or the `tryCatch` blocks in `rMVPA` for detailed debugging messages.
*   **Memory**: Searchlight analyses, especially with large radii or many randomized iterations, can be memory-intensive. Monitor usage accordingly.
*   **Randomized Searchlight**: If using `method = "randomized"`, ensure the metric returned by `custom_func` is meaningful when averaged across overlapping spheres.

# Summary

The `run_custom_regional` and `run_custom_searchlight` functions provide a powerful mechanism to extend `rMVPA`'s capabilities. They allow researchers to apply bespoke analyses within ROIs or searchlight spheres while benefiting from the package's iteration, parallelization, and error-handling infrastructure. By defining a custom function that meets the specified input and output requirements, you can easily integrate diverse analytical approaches into your neuroimaging workflow.

For implementation details, refer to the source code in `R/custom.R`.