---
title: 'ERA-RSA: Cross-Decoding Between Encoding and Retrieval'
author: rMVPA authors
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{ERA-RSA: Cross-Decoding Between Encoding and Retrieval}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE) && "albers_register_fonts" %in% getNamespaceExports("albersdown")) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE) && "theme_albers" %in% getNamespaceExports("albersdown")) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
if (isTRUE(getOption("knitr.in.progress"))) {
  options(
    crayon.enabled = FALSE,
    cli.num_colors = 1,
    cli.unicode = FALSE
  )
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
suppressPackageStartupMessages({
  library(rMVPA)
  library(neuroim2)
})
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

## Overview

The **ERA-RSA model** (`era_rsa_model()`) combines:

- **Encoding–Retrieval Accuracy (ERA)** — first-order similarity between
  encoding and retrieval patterns for each item, and
- **Representational Geometry** — second-order similarity between
  encoding and retrieval representational dissimilarity matrices (RDMs).

This vignette walks through a small synthetic example of
**encoding → retrieval cross-decoding** using ERA-RSA, and illustrates
how to run both **regional** and **searchlight** analyses.

ERA-RSA is designed for situations where you have:

- an encoding phase (e.g., stimulus viewing),
- a retrieval phase (e.g., recognition or recall),
- an item key that links encoding and retrieval trials.

## 1. Synthetic Encoding/Retrieval Dataset

We start from `gen_sample_dataset()` and treat its training split as
**encoding** and its external test split as **retrieval**.

```{r era_data}
set.seed(123)

toy <- gen_sample_dataset(
  D            = c(4, 4, 4),  # modest volume for speed
  nobs         = 24,
  nlevels      = 3,           # 3 item categories
  blocks       = 3,           # 3 runs
  external_test = TRUE,       # required for encoding → retrieval
  ntest_obs     = 24
)

# Add an item key column to both train and test designs
item_keys <- sprintf("item_%02d", seq_len(nrow(toy$design$train_design)))
toy$design$train_design$item <- factor(item_keys, levels = item_keys)
toy$design$test_design$item  <- factor(item_keys, levels = item_keys)

head(toy$design$train_design)
```

Each row in `train_design` and `test_design` corresponds to a trial.
The `item` column links encoding and retrieval presentations of the
same stimulus. The toy data use paired row IDs as item keys so the example has
one encoding and one retrieval observation per item; in an applied analysis,
this column would usually be a stimulus, image, word, or event identifier from
your design table.

### `key_var` vs `y_train` / `y_test` and `phase_var`

An `mvpa_design` already contains `y_train` / `y_test`, which are the
**response labels for each trial** (often category labels). ERA-RSA
adds two extra notions:

- **`key_var`**: an *item key* that links encoding and retrieval trials
  belonging to the same underlying stimulus (e.g., image ID). ERA
  metrics such as `era_top1_acc` operate at the item level: for each
  retrieval trial we ask whether its top match among encoding trials
  has the same `key_var`, not just the same category.
- **`phase_var`**: a phase label (e.g., encoding vs retrieval). In the
  external-test setup used here, the train split is treated as
  encoding and the test split as retrieval, so `phase_var` is optional.
  It is required when both phases live in the same image series.

In short, `y_train`/`y_test` describe *what was shown* on each trial
(class/category), while `key_var` defines *which item* that trial
belongs to; ERA-RSA uses the item keys to build encoding–retrieval
similarity and geometry.

For a combined series with alternating encoding and retrieval blocks, keep one
design row per image volume and identify phases explicitly:

```r
era_ms <- era_rsa_model(
  dataset = combined_dataset,
  design = combined_design,
  key_var = ~ item,
  phase_var = ~ phase,
  encoding_level = "E",
  retrieval_level = "R",
  pairing = "one_to_one"
)
```

Pairing is always by `key_var`, never by row adjacency or block order. rMVPA
splits the image series internally while preserving the original observation
order within each phase.

## 2. Regional ERA-RSA Model

We build an `era_rsa_model` and run a simple regional analysis with a
small number of regions.

The two similarity arguments operate at different levels:

- `distfun = cordist("pearson")` says how to build each within-phase RDM.
  Within a region, ERA-RSA first averages trials by `key_var`, then computes
  one encoding RDM and one retrieval RDM from those item-level patterns.
- `rsa_simfun = "spearman"` says how to compare those two RDMs. This is the
  second-order RSA step: the lower triangle of the encoding RDM is correlated
  with the lower triangle of the retrieval RDM. The resulting regional metric
  is `geom_cor`.

There is no external model RDM in this example. The "model" for the
second-order comparison is the encoding geometry itself: we ask whether the
relative distances among encoded items are preserved during retrieval.

```{r era_regional_model}
# Simple 3-region mask for demonstration
region_mask <- NeuroVol(
  sample(1:3, size = length(toy$dataset$mask), replace = TRUE),
  space(toy$dataset$mask)
)

era_ms <- era_rsa_model(
  dataset   = toy$dataset,
  design    = toy$design,
  key_var   = ~ item,          # item key linking encoding ↔ retrieval
  pairing   = "one_to_one",    # require one encoding and one retrieval row per item
  era_simfun = "pearson",      # item-level encoding-retrieval similarity
  distfun   = cordist("pearson"), # builds encoding and retrieval RDMs
  rsa_simfun = "spearman"         # compares those RDMs; output is geom_cor
)

era_ms

era_res <- run_regional(era_ms, region_mask)
era_res
```

The `performance_table` contains one row per region and several ERA-RSA
metrics.

```{r era_regional_perf}
head(era_res$performance_table)
```

```{r check-era-rsa-results, include=FALSE}
stopifnot(
  is.data.frame(era_res$performance_table),
  nrow(era_res$performance_table) == 3,
  all(era_res$performance_table$n_items == length(item_keys)),
  all(is.finite(era_res$performance_table$era_top1_acc)),
  all(era_res$performance_table$era_top1_acc >= 0),
  all(era_res$performance_table$era_top1_acc <= 1),
  "geom_cor" %in% names(era_res$performance_table),
  any(is.finite(era_res$performance_table$geom_cor))
)
```

Key metrics include:

- `era_top1_acc` — top-1 encoding→retrieval accuracy for the item key.
- `era_diag_mean` — mean encoding–retrieval similarity on the diagonal.
- `era_diag_minus_off` — mean item-specific similarity: for each retrieval
  item, matched similarity minus its own mean similarity to nonmatching
  encoding items. With complete similarities this equals diagonal mean minus
  the grand off-diagonal mean. If pairwise similarities are missing, each item
  with an estimable contrast receives equal weight.
- `geom_cor` — the second-order correlation between the encoding and
  retrieval RDMs. With `rsa_simfun = "spearman"`, this is a rank correlation
  over item-pair dissimilarities.

These quantify both cross-decoding performance and representational
geometry alignment between phases.

```{r era-regional-key-metrics}
era_res$performance_table[
  ,
  c("roinum", "era_top1_acc", "era_diag_minus_off", "geom_cor")
]
```

## 3. Relating item similarity to retrieval ratings

A retrieval rating can be related directly to an item-specific E–R score
computed in each region or searchlight sphere. By default, this is matched
similarity minus that retrieval item's mean nonmatch similarity, so every
retrieval trial has its own local background. Missing ratings do not remove
items from `era_diag_mean`; each zero-order correlation uses its own complete
pairs. Set `era_association_score = "matched"` only when the intended response
is raw matched E–R similarity without background correction.

```{r era-association-metadata}
set.seed(512)
toy$design$test_design$vividness <- sample(1:6, length(item_keys), replace = TRUE)
toy$design$test_design$vividness[c(4, 17)] <- NA_real_
toy$design$test_design$retrieval_run <- factor(
  rep(paste0("run_", 1:3), length.out = length(item_keys))
)
toy$design$test_design$trial_order <- seq_along(item_keys)
```

Use `era_correlates` for the unadjusted relationship. Use
`era_association` for the adjusted model, `era_effects` to select directional
one-degree-of-freedom effects, and `era_effects_block` for omnibus tests of
one or more terms:

```{r era-association-model}
era_assoc_ms <- era_rsa_model(
  dataset = toy$dataset,
  design = toy$design,
  key_var = ~ item,
  pairing = "one_to_one",
  era_simfun = "spearman",
  era_min_voxels = 3,
  era_components = "item",
  era_association_score = "item_specificity",
  era_correlates = ~ vividness,
  era_cor_method = "spearman",
  era_association = ~ vividness + retrieval_run + trial_order,
  era_effects = ~ vividness,
  era_effects_block = list(acquisition = ~ retrieval_run + trial_order),
  era_min_complete = 4
)

era_assoc_res <- run_regional(era_assoc_ms, region_mask)
era_assoc_res$performance_table[
  ,
  c("roinum", "era_diag_mean", "era_vividness_cor", "era_vividness_n",
    "era_assoc_part_r_vividness", "era_assoc_dr2_acquisition",
    "era_assoc_F_acquisition", "era_assoc_df1_acquisition",
    "era_assoc_n", "era_assoc_df_resid")
]
```

`era_vividness_cor` is the zero-order Spearman correlation with item
specificity.
`era_assoc_part_r_vividness` is the signed semi-partial correlation after
adjusting for retrieval run and trial order. It is invariant to linear
rescaling of vividness. Adjustment-only terms remain in the model but do not
create unsigned or reference-dependent maps. A multi-level factor needs an
explicit one-degree-of-freedom contrast before it can be requested in
`era_effects`. Blocks do not have that restriction: here, the
`acquisition` block jointly removes the multi-column retrieval-run factor and
trial order. `era_assoc_dr2_acquisition` is the block's incremental R-squared,
`era_assoc_F_acquisition` is its partial F statistic, and
`era_assoc_df1_acquisition` is the fitted rank difference between the full and
reduced models.

The association model uses one joint complete-case set across similarity and
all regression variables, reported as `era_assoc_n`; every block comparison
reuses exactly those rows. It does not emit raw beta or per-term unsigned
R-squared maps. Before conventional second-level analysis, Fisher-transform
subject-level correlation or part-r maps where that model's assumptions call
for it.

```{r check-era-association-results, include=FALSE}
assoc_metrics <- era_assoc_res$performance_table
stopifnot(
  all(c("era_vividness_cor", "era_vividness_n",
        "era_assoc_part_r_vividness", "era_assoc_dr2_acquisition",
        "era_assoc_F_acquisition", "era_assoc_df1_acquisition",
        "era_assoc_n") %in% names(assoc_metrics)),
  any(is.finite(assoc_metrics$era_vividness_cor)),
  any(is.finite(assoc_metrics$era_assoc_part_r_vividness)),
  any(is.finite(assoc_metrics$era_assoc_F_acquisition)),
  all(assoc_metrics$era_vividness_n == length(item_keys) - 2)
)
```

## 4. When do you need ERA partitioning?

The ERA-RSA summary above gives you direct item matching and a raw geometry
correlation. `era_partition_model()` asks a sharper follow-up question: how
much of the encoding-retrieval relationship is uniquely explained by
same-item transfer, and how much is uniquely explained by preserved
representational geometry after nuisance structure is accounted for?

This makes it useful when a single cross-decoding accuracy is too compressed.
For example, two regions could have similar top-1 accuracy, but one may carry
item-specific diagonal similarity while another mainly preserves the relative
geometry among items.

```{r era-partition-model}
partition_ms <- era_partition_model(
  dataset = toy$dataset,
  design  = toy$design,
  key_var = ~ item,
  distfun = cordist("pearson"),
  rsa_simfun = "spearman",
  xdec_link_by = "item",
  include_procrustes = FALSE # keep the toy example fast; see note below
)

partition_res <- run_regional(partition_ms, region_mask)
```

```{r era-partition-summary}
partition_cols <- c(
  "roinum",
  "n_items",
  "naive_top1_acc",
  "xdec_Accuracy",
  "first_order_delta_r2",
  "second_order_delta_r2",
  "geom_cor"
)
partition_res$performance_table[, partition_cols]
```

```{r check-era-partition-results, include=FALSE}
partition_metrics <- partition_res$performance_table
stopifnot(
  is.data.frame(partition_metrics),
  nrow(partition_metrics) == 3,
  all(partition_metrics$n_items == length(item_keys)),
  all(is.finite(partition_metrics$naive_top1_acc)),
  all(partition_metrics$naive_top1_acc >= 0),
  all(partition_metrics$naive_top1_acc <= 1),
  any(is.finite(partition_metrics$first_order_delta_r2)),
  any(is.finite(partition_metrics$second_order_delta_r2))
)
```

Read these metrics in layers:

- `naive_top1_acc` is the item-level direct-transfer score computed from the
  encoding-by-retrieval similarity matrix.
- `xdec_Accuracy` is the trial-level direct cross-decoding metric, matched to
  the same prototype scorer used by `naive_xdec_model()`.
- `first_order_delta_r2` is the unique variance explained by same-item
  transfer in the cross-state similarity matrix.
- `second_order_delta_r2` and `geom_cor` summarize whether the encoding
  item geometry is preserved at retrieval.

For this random toy dataset, the values should be near chance or near zero.
In real data, a region with high `first_order_delta_r2` and weak
`second_order_delta_r2` supports item-specific reinstatement without much
global geometry preservation. The reverse pattern suggests that the relative
arrangement of items is preserved even when direct item identification is
weak.

`era_partition_model()` can also include block, run, category, temporal, and
custom nuisance regressors through item-level vectors such as
`item_block_enc`, `item_block_ret`, `item_run_enc`, `item_run_ret`,
`item_time_enc`, `item_time_ret`, `item_category`, plus
`first_order_nuisance` and `second_order_nuisance`. By default, available
item-level nuisance groups are added automatically. Use
`auto_nuisance = FALSE` to use only custom nuisance regressors, or a character
vector such as `auto_nuisance = c("run", "time")` to keep selected groups.
If you have enough paired items, set `include_procrustes = TRUE` to add a
leakage-free orthogonal alignment decoder.

`include_procrustes` controls an optional decoder that first learns an
orthogonal Procrustes map from encoding prototypes into retrieval prototype
space. The fit is leave-one-item-out: when scoring item *i*, the rotation is
estimated from all other paired items, then the held-out retrieval item is
matched against the aligned encoding prototypes. This gives Procrustes metrics
such as `procrustes_top1_acc`, `procrustes_diag_mean`, and
`procrustes_diag_minus_off`. Use it when you want to ask whether encoding and
retrieval occupy the same representational space up to a global rotation or
reflection. Leave it off for small toy examples or when there are too few
paired items for stable held-out alignment; `min_procrustes_train_items`
sets that minimum.

## 5. Searchlight ERA-RSA

We can also run ERA-RSA in a searchlight mode to obtain whole-brain
maps of the same metrics.

```{r era_searchlight}
set.seed(456)

era_sl <- run_searchlight(
  era_assoc_ms,
  radius = 2,         # searchlight radius in voxels
  method = "standard" # ERA-RSA currently uses standard searchlight
)

era_sl
```

The `searchlight_result` contains:

- `metrics`: names of the output maps (e.g., `era_diag_minus_off`,
  `era_vividness_cor`, and `era_assoc_part_r_vividness`),
- `results`: a list of `NeuroVol` maps, one per metric.

```{r era_searchlight_metrics}
era_sl$metrics
```

We can save the searchlight maps using `save_results()`:

```{r era_save, eval=FALSE}
out_dir <- tempfile("era_rsa_sl_")
dir.create(out_dir, showWarnings = FALSE)

save_results(era_sl, out_dir, level = "standard")
list.files(file.path(out_dir, "maps"))
```

This will create one NIfTI file per requested metric (e.g.,
`era_diag_minus_off.nii.gz`, `era_vividness_cor.nii.gz`) that can be viewed in
your favorite neuroimaging software.

For association-focused HPC analyses, combine the item-only model above with
the shard backend:

```r
era_sl_hpc <- run_searchlight(
  era_assoc_ms,
  radius = 2,
  method = "standard",
  engine = "era_rsa_fast",
  backend = "shard"
)
```

`era_components = "item"` skips top-1 identification and both within-phase
RDMs. For a one-to-one volumetric standard searchlight, `engine = "auto"`
(the default) selects the same dedicated ERA-RSA engine shown explicitly
above. It slices train/test matrices directly, applies the established
train-only feature filter and center-preservation rule, and then calls the same
`fit_roi.era_rsa_model()` computation used by the general-purpose iterator. With
`backend = "shard"`, workers slice shared matrices directly rather than
constructing and serializing an ROI object for every sphere. Request
`era_components = c("item", "identification")` when top-1 accuracy is also
needed, or retain the default components when geometry maps are required; all
three component families use the same dedicated extraction engine. Repeated-
item averaging, non-volumetric data, and randomized/resampled searchlights
continue through the general-purpose iterator. Its existing compatibility key,
`engine = "legacy"`, remains available for an explicit reference run.

## 6. Adding Confounds and Lag Information

ERA-RSA can optionally incorporate item-level confounds and lag
variables. These are all defined at the **item** level, not the
trial level, and must align with the levels of `key_var`.

### Item-level confounds (`confound_rdms`)

`confound_rdms` is a named list of K×K matrices or `"dist"` objects
describing item-by-item nuisance structure (e.g., block/run/time), where
rows and columns correspond to item keys.

Run confounds can now be supplied in the simpler item-level form used by the
cross-run diagnostic. You pass `item_run_enc` and `item_run_ret`; ERA-RSA
derives the same-run RDMs internally for `geom_cor_run_partial`.

```{r era-confound-metadata}
items <- levels(toy$design$train_design$item)

Mode <- function(x) { ux <- unique(x); ux[which.max(tabulate(match(x, ux)))] }
item_run_enc <- sapply(items, function(it) {
  Mode(toy$design$train_design$block_var[toy$design$train_design$item == it])
})
item_run_ret <- sapply(items, function(it) {
  Mode(toy$design$test_design$block_var[toy$design$test_design$item == it])
})
names(item_run_enc) <- names(item_run_ret) <- items

item_run_enc <- factor(paste0("enc_", item_run_enc))
item_run_ret <- factor(paste0("ret_", item_run_ret))
names(item_run_enc) <- names(item_run_ret) <- items
```

You can add temporal or block RDMs to `confound_rdms` and choose which groups
enter the general partial-geometry metric with `partial_against`.

```{r era-confound-model}
time_enc <- setNames(seq_along(items), items)
time_enc_rdm <- abs(outer(time_enc, time_enc, "-"))
rownames(time_enc_rdm) <- colnames(time_enc_rdm) <- items

era_ms_conf <- era_rsa_model(
  dataset = toy$dataset,
  design  = toy$design,
  key_var = ~ item,
  phase_var = ~ block_var,
  confound_rdms = list(time_enc = time_enc_rdm),
  partial_against = c("run", "time"),
  item_run_enc = item_run_enc,
  item_run_ret = item_run_ret
)

era_conf_res <- run_regional(era_ms_conf, region_mask)
```

```{r era-confound-summary}
era_conf_res$performance_table[
  ,
  c("roinum", "geom_cor", "geom_cor_partial",
    "geom_cor_run_partial", "geom_cor_xrun", "beta_time_enc")
]
```

```{r check-era-confound-results, include=FALSE}
conf_metrics <- era_conf_res$performance_table
stopifnot(
  "geom_cor_partial" %in% names(conf_metrics),
  "geom_cor_run_partial" %in% names(conf_metrics),
  "geom_cor_xrun" %in% names(conf_metrics),
  "beta_time_enc" %in% names(conf_metrics),
  any(is.finite(conf_metrics$geom_cor_partial)),
  any(is.finite(conf_metrics$geom_cor_run_partial))
)
```

The three geometry metrics have different interpretations:

- `geom_cor`: raw correlation between encoding and retrieval RDMs.
- `geom_cor_partial`: residualized geometry correlation after regressing out
  the nuisance groups selected by `partial_against`.
- `geom_cor_run_partial`: legacy run-only partial geometry correlation. It is
  still reported for continuity and is derived from `item_run_enc` /
  `item_run_ret` when explicit `confound_rdms$run_enc` and
  `confound_rdms$run_ret` are absent.
- `geom_cor_xrun`: raw geometry correlation restricted to item pairs that
  differ in both encoding and retrieval run.

For temporal confounds, a common choice is `partial_against = c("run",
"time")`. If you want only a temporal residualization, use
`partial_against = "time"` or the exact RDM name, such as
`partial_against = "time_enc"`.

### Whole-mask global nuisance

If you want to remove broad similarity structure shared across the whole
analysis mask, set `global_nuisance = TRUE`. ERA-RSA computes item-level
whole-mask RDMs once and adds them as `global_enc` and `global_ret` nuisance
RDMs. When `partial_against` is left at its default, enabling
`global_nuisance` makes the effective partial model `c("run", "global")`;
set `partial_against` explicitly if you want a different nuisance set.

```{r era-global-nuisance-model}
item_block <- factor(item_run_enc, levels = sort(unique(item_run_enc)))

era_ms_global <- era_rsa_model(
  dataset = toy$dataset,
  design  = toy$design,
  key_var = ~ item,
  phase_var = ~ block_var,
  item_block = item_block,
  item_run_enc = item_run_enc,
  item_run_ret = item_run_ret,
  global_nuisance = TRUE
)

era_global_res <- run_regional(era_ms_global, region_mask)
```

```{r era-global-nuisance-summary}
era_global_res$performance_table[
  ,
  c("roinum", "geom_cor_partial", "geom_cor_run_partial",
    "beta_global_enc", "beta_global_ret")
]
```

```{r check-era-global-nuisance-results, include=FALSE}
global_metrics <- era_global_res$performance_table
stopifnot(
  all(c("geom_cor_partial", "beta_global_enc", "beta_global_ret") %in%
        names(global_metrics)),
  any(is.finite(global_metrics$geom_cor_partial))
)
```

### Block structure (`item_block`)

`item_block` encodes a per-item block/condition, typically derived from
`design$train_design$block_var`:

```r
era_ms_block <- era_rsa_model(
  dataset    = toy$dataset,
  design     = toy$design,
  key_var    = ~ item,
  phase_var  = ~ block_var,
  item_block = item_block
)
```

This enables block-limited ERA contrasts:

- `era_diag_minus_off_same_block`: each retrieval item's matched ERA minus its
  mean similarity to other encoding items in the same block, averaged equally
  over retrieval items.
- `era_diag_minus_off_diff_block`: each retrieval item's matched ERA minus its
  mean similarity to encoding items in different blocks, averaged equally over
  retrieval items.

### Lag information (`item_lag`)

`item_lag` is a numeric per-item lag between encoding and retrieval
(e.g., retrieval onset – encoding onset), aligned to item keys. It is
used to compute:

- `era_lag_cor`: Spearman correlation between the ERA diagonal and
  `item_lag`, using complete cases.

In applied analyses, you would construct `item_block`, `item_lag`,
`item_run_enc`, `item_run_ret`, and `confound_rdms` from your
experiment’s design tables following the patterns above, then pass
them into `era_rsa_model()`.

### Run and temporal nuisance in `era_partition_model()`

The variance-partition model uses the same item-level metadata but separates
first-order cross-state similarity from second-order geometry preservation.
When `auto_nuisance` includes `"run"`, the model adds same-run nuisance
regressors to both levels of the partition.

```{r era-partition-nuisance-model}
partition_nuis_ms <- era_partition_model(
  dataset = toy$dataset,
  design  = toy$design,
  key_var = ~ item,
  item_run_enc = item_run_enc,
  item_run_ret = item_run_ret,
  item_time_enc = time_enc,
  item_time_ret = time_enc + length(items),
  auto_nuisance = c("run", "time"),
  include_procrustes = FALSE
)

partition_nuis_res <- run_regional(partition_nuis_ms, region_mask)
```

```{r era-partition-nuisance-summary}
partition_nuis_res$performance_table[
  ,
  c("roinum", "first_order_delta_r2", "second_order_delta_r2",
    "nuisance_first_order_n", "nuisance_second_order_n")
]
```

```{r check-era-partition-nuisance-results, include=FALSE}
partition_nuis_metrics <- partition_nuis_res$performance_table
stopifnot(
  all(partition_nuis_metrics$nuisance_first_order_n == 4),
  all(partition_nuis_metrics$nuisance_second_order_n == 4),
  any(is.finite(partition_nuis_metrics$first_order_delta_r2)),
  any(is.finite(partition_nuis_metrics$second_order_delta_r2))
)
```

Here the first-order nuisance set contains `same_run_cross`, `enc_time`,
`ret_time`, and `abs_lag`. The second-order nuisance set contains
`same_run_enc`, `same_run_ret`, `temporal_distance_enc`, and
`temporal_distance_ret`. Set `auto_nuisance = FALSE` when you want to replace
these defaults with custom kernels through `first_order_nuisance` and
`second_order_nuisance`.

Whole-mask nuisance is also available in the partition model. When
`auto_nuisance` includes `"global"`, the first-order partition receives
`global_cross`; the second-order partition receives `global_enc` and
`global_ret`.

```{r era-partition-global-model}
partition_global_ms <- era_partition_model(
  dataset = toy$dataset,
  design  = toy$design,
  key_var = ~ item,
  global_nuisance = TRUE,
  auto_nuisance = "global",
  include_procrustes = FALSE
)

partition_global_res <- run_regional(partition_global_ms, region_mask)
```

```{r era-partition-global-summary}
partition_global_res$performance_table[
  ,
  c("roinum", "first_order_delta_r2", "second_order_delta_r2",
    "nuisance_first_order_n", "nuisance_second_order_n")
]
```

```{r check-era-partition-global-results, include=FALSE}
partition_global_metrics <- partition_global_res$performance_table
stopifnot(
  all(partition_global_metrics$nuisance_first_order_n == 1),
  all(partition_global_metrics$nuisance_second_order_n == 2),
  any(is.finite(partition_global_metrics$first_order_delta_r2)),
  any(is.finite(partition_global_metrics$second_order_delta_r2))
)
```

## 7. Summary

- `era_rsa_model()` provides a unified framework for:
  - cross-decoding between encoding and retrieval, and
  - comparing encoding and retrieval representational geometries, and
  - mapping zero-order and adjusted directional associations between
    item-specific matched-minus-nonmatch similarity and retrieval variables.
- `era_partition_model()` separates direct item transfer, variance uniquely
  explained by same-item similarity, and variance explained by preserved
  second-order geometry.
- It integrates naturally with:
  - `run_regional()` for ROI-based analyses, and
  - `run_searchlight()` for whole-brain mapping.
- Outputs are standard rMVPA result objects, so the same tooling
  (`save_results()`, plotting, etc.) applies as for other models such
  as `naive_xdec_model()` and `remap_rrr_model()`.

ERA-RSA is particularly useful when you want to go beyond simple
accuracy and ask **how similar the geometry of neural representations
is across phases**, and how that similarity varies across brain regions.
