---
title: ITEM Trial-Wise Decoding
author: rMVPA authors
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{ITEM Trial-Wise Decoding} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-teal');});</script>

---

```{r setup, include=FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
has_fmrilss <- requireNamespace("fmrilss", quietly = TRUE)
suppressPackageStartupMessages({
  library(rMVPA)
  library(neuroim2)
})
```

```{r fmrilss-note, echo=FALSE, results='asis', eval=!has_fmrilss}
cat("> **Note:** the `fmrilss` package is not installed, so the code in this",
    "vignette is shown but not executed. Install it with",
    "`remotes::install_github(\"bbuchsbaum/fmrilss\")` to run these examples.\n")
```

## Overview

ITEM (Inverse Transformed Encoding Model) decoding is a trial-wise pipeline
for ROI and searchlight analysis. Unlike standard MVPA, which decodes from a
single pattern per trial without accounting for how overlapping hemodynamic
responses correlate trial estimates, ITEM makes the trial-estimation step
explicit and then *corrects for the covariance it induces*.

The pipeline has two stages:

1. **Trial estimation.** A least-squares-all (LS-A) GLM regresses the
   TR-level data onto a trial-wise design matrix `X_t`, yielding one response
   pattern per trial (`Gamma`).
2. **Covariance-aware decoding.** Because nearby trials share overlapping
   HRFs, the trial estimates are correlated even when the underlying signals
   are not. ITEM computes this trial-by-trial covariance `U` analytically from
   `X_t` and uses it to whiten the decoder, then evaluates with
   leave-one-run-out cross-validation.

This differs from the archived `hrfdecoder` pathway, which fit a
continuous-time decoder directly on TR-level data and aggregated TR
predictions to events afterward. Reach for ITEM when you want an explicit
trial-estimation stage and direct control over trial covariance handling,
especially for trial-level diagnostics.

The implementation delegates the numerical work (`item_build_design`,
`item_compute_u`, `lsa`, `item_cv`) to the
[`fmrilss`](https://github.com/bbuchsbaum/fmrilss) package, which is an
optional dependency.

## The two ITEM objects

| Object | Constructor | Role |
|---|---|---|
| Design | `item_design()` | Couples the trial-wise design matrix `X_t` and trial targets `T_target` to TR-level observations. |
| Model spec | `item_model()` | Configures the decoder (mode, metric, ridge, solver) and precomputes `U`. Compatible with `run_regional()` and `run_searchlight()`. |

The key distinction from a standard `mvpa_design` is that supervised
information lives at the **trial** level (`T_target`, length `n_trials`),
while the data extracted per ROI/searchlight lives at the **TR** level
(`nrow(X_t)` rows). Cross-validation folds are derived internally from the
trial-level `run_id` -- you do not specify a `block_var`.

## Quick example

We simulate a small dataset, build a trial-wise design, and run a regression
decode over three ROIs.

```{r build, eval=has_fmrilss}
set.seed(42)

n_time   <- 80   # TR-level observations
n_trials <- 24   # trials to decode
n_runs   <- 4

data_info <- gen_sample_dataset(
  D = c(10, 10, 10), nobs = n_time, nlevels = 2, blocks = n_runs
)

# Trial-wise design matrix: one column per trial (n_time x n_trials).
# In real analyses this is built from your event onsets and an HRF.
X_t    <- matrix(rnorm(n_time * n_trials), nrow = n_time, ncol = n_trials)
run_id <- rep(seq_len(n_runs), each = n_trials / n_runs)

# Continuous trial-level target (regression mode).
T_target <- as.numeric(scale(rnorm(n_trials)))

des <- item_design(
  train_design = data_info$design$train_design,
  X_t          = X_t,
  T_target     = T_target,
  run_id       = run_id
)

print(des)
```

```{r model, eval=has_fmrilss}
model <- item_model(
  dataset = data_info$dataset,
  design  = des,
  mode    = "regression",
  metric  = "correlation",
  solver  = "svd"
)
```

```{r regional, eval=has_fmrilss}
roi_mask <- NeuroVol(
  sample(1:3, length(data_info$dataset$mask), replace = TRUE),
  space(data_info$dataset$mask)
)

results <- run_regional(model, roi_mask)
results$performance_table
```

Every ROI/searchlight location reports three scalars:

- `item_score_mean` -- mean of the chosen metric across folds,
- `item_score_sd` -- across-fold standard deviation,
- `item_n_folds` -- number of cross-validation folds (runs).

```{r check-regional, include=FALSE, eval=has_fmrilss}
stopifnot(
  is.data.frame(results$performance_table),
  "item_score_mean" %in% names(results$performance_table),
  all(is.finite(results$performance_table$item_score_mean))
)
```

With random data the correlation hovers near zero. With real fMRI data, an
`item_score_mean` reliably above zero (regression) or above chance
(classification) indicates decodable trial-level structure.

## Classification mode

For categorical targets, pass a factor (or character vector) as `T_target`
and set `mode = "classification"`:

```{r classification, eval=has_fmrilss}
T_class <- factor(rep(c("face", "house"), length.out = n_trials))

des_cls <- item_design(
  train_design = data_info$design$train_design,
  X_t          = X_t,
  T_target     = T_class,
  run_id       = run_id
)

model_cls <- item_model(
  dataset = data_info$dataset,
  design  = des_cls,
  mode    = "classification",
  metric  = "balanced_accuracy",
  solver  = "svd"
)

run_regional(model_cls, roi_mask)$performance_table
```

The `metric` argument is validated against `mode` at construction, so a
mismatch (for example `mode = "classification"` with `metric = "rmse"`)
fails immediately rather than producing a map of errored ROIs. Valid
choices are `"accuracy"` / `"balanced_accuracy"` for classification and
`"correlation"` / `"rmse"` for regression. `metric = NULL` uses the
`fmrilss` default for the mode.

## Searchlight

The same model spec drops straight into `run_searchlight()`. ITEM exposes an
output schema, so the searchlight produces one map per scalar metric:

```{r searchlight, eval=has_fmrilss}
sl <- run_searchlight(model, radius = 4, method = "standard")
names(sl$results)
```

For whole-brain searchlights, set `return_predictions = FALSE` on the model
spec. The searchlight schema combiner only keeps the scalar maps, but
`run_regional()` retains the full per-fold prediction payload on every ROI;
disabling predictions bounds memory when there are many ROIs or trials.

## Accounting for temporal autocorrelation

If you have an estimate of the TR-level noise covariance (or its precision),
pass it through `V` / `v_type` on the design. ITEM folds it into the analytic
trial covariance `U`:

```{r covariance, eval=FALSE}
# V_cov: n_time x n_time temporal covariance (e.g. from an AR model)
des_v <- item_design(
  train_design = train_design,
  X_t          = X_t,
  T_target     = T_target,
  run_id       = run_id,
  V            = V_cov,
  v_type       = "cov"      # or "precision" if you pass the inverse
)
```

Nuisance regressors that should be projected out during LS-A trial
estimation (motion, drift) go through the `Z` / `Nuisance` arguments of
`item_design()`.

## Trial-alignment guard

A common, hard-to-spot bug is a mismatch between the trial order in `X_t`
and the trial order in your behavioural targets. ITEM can guard against this
with a hash of the trial identifiers:

```{r hash, eval=FALSE}
des <- item_design(
  X_t        = X_t,
  T_target   = T_target,
  run_id     = run_id,
  trial_id   = trial_ids,
  trial_hash = fmrilss:::.item_simple_hash(trial_ids)
)

model <- item_model(dataset, des, mode = "regression", check_hash = TRUE)
```

When `check_hash = TRUE`, each fold verifies the trial hash before decoding
and the ROI fails loudly ("Trial hash mismatch") instead of silently
returning a meaningless score.

## Tuning knobs

| Argument | Default | Purpose |
|---|---|---|
| `ridge_u` | `0` | Ridge added when inverting to form `U`. |
| `ridge_w` | `1e-4` | Ridge on the per-fold ITEM weights. |
| `solver` | `"chol"` | Linear solver (`"chol"`, `"svd"`, `"pinv"`); falls back to `"svd"` on collinearity. |
| `u_storage` | `"matrix"` | Store `U` as a full matrix or per-run blocks (`"by_run"`); use blocks for many trials. |
| `lsa_method` | `"r"` | LS-A backend (`"r"` or `"cpp"`). |

## Troubleshooting

**"requires package 'fmrilss'"** -- ITEM delegates its numerics to
`fmrilss`. Install it with
`remotes::install_github("bbuchsbaum/fmrilss")`.

**"ROI rows ... must match nrow(X_t)"** -- the data extracted per ROI has a
different number of TR rows than `X_t`. Confirm `nrow(X_t) == nobs(dataset)`
and that the dataset and design describe the same scan.

**"requested 'chol' but used 'svd'"** -- a benign fallback warning emitted
when trial estimates are collinear within a fold; the decode still completes
via the SVD solver. Increase `ridge_w` to suppress it.

**Scores near chance everywhere** -- expected for null data. With real data,
check that `X_t` is built from the correct onsets/HRF and that `T_target` is
aligned to trials (consider the `check_hash` guard above).

## Next steps

- `?item_model`, `?item_design` -- full parameter documentation
- `vignette("Naive_Cross_Decoding")` -- a no-learning cross-domain baseline
- `vignette("Searchlight_Analysis")` -- general searchlight workflow
