---
title: "Model-Space Representational Connectivity"
author: Bradley Buchsbaum
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Model-Space Representational Connectivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width = 6,
  fig.height = 4.4,
  dpi = 110
)
```

```{r libs, include = FALSE}
library(rMVPA)
library(neuroim2)
```

## What this vignette is for

You've fit RSA in a handful of ROIs and learned how strongly each region tracks each model RDM. Two natural follow-up questions immediately appear:

1. **Where else in the brain does this same representational geometry live?** — i.e. ROI-to-ROI representational *connectivity* through your model RDMs.
2. **How do I ask the connectivity question across two different stimulus sets** — say items from movie A and items from movie B, where the items themselves are not the same?

rMVPA answers both with one fitted object. A single call to `rsa_model(..., return_fingerprint = TRUE)` produces, per ROI or per searchlight, a small *model-space fingerprint* vector. `model_space_connectivity()` then turns those fingerprints into the cross-ROI similarity summaries you want — without you having to refit anything or assemble RDM vectors by hand.

The shared abstraction is **pairwise model-space representation per analysis unit**. Within-unit RSA and across-unit connectivity are two views of the same fit.

## Two kinds of representational connectivity

rMVPA exposes two complementary ways to ask "do these ROIs share representational geometry?" Picking the right one is the single most important decision in this analysis:

| | **Model-space connectivity** *(this vignette)* | **Feature-RSA connectivity** *(`vignette("Feature_RSA_Connectivity")`)* |
|:--|:--|:--|
| **Question it answers** | Do ROIs project onto the *same axes of my declared model RDMs*? | Do ROIs *predict the same trial-by-trial similarity structure* through a learned feature space? |
| **What you supply** | One or more explicit model RDMs | A feature matrix `F` (or similarity matrix `S`) |
| **What gets compared across ROIs** | Whitened projections onto the model-RDM subspace (length-K fingerprints, K = number of model RDMs after orthogonalisation) | Predicted RDM vectors of length `n_trials × (n_trials − 1) / 2` |
| **Fitting required?** | No — just project the neural pair vector onto a fixed orthonormal basis | Yes — cross-validated PLS / PCA / glmnet inside `feature_rsa_model()` |
| **Memory per unit** | O(K), typically 2–10 numbers | O(n_pairs), can be 10⁴–10⁵ numbers |
| **Searchlight-friendly?** | Yes — fingerprints fit easily; k-means anchors avoid `n_centers²` blow-up | Marginal — requires file-backed RDM storage and is more expensive to compare across centers |
| **Best when** | You have specific theoretical RDMs (semantic, visual, motor…) and want to ask which regions express each | You don't have a clean theoretical RDM and want a model-free, data-driven similarity measure |
| **Diagonal of the connectivity matrix** | ROI's *strength* of model-RDM expression | ROI's CV-predicted-vs-observed RDM alignment |

A useful rule of thumb: **declared model → model-space connectivity; learned model → feature-RSA connectivity.** If you find yourself reaching for both, run them both — they make different statistical commitments and frequently agree only partially, which itself is informative.

## The mental model in one diagram

```
items / pair design   ───►   per-unit pair RSA fit   ───►   per-unit fingerprint f_u
                                       │
                                       ├───►  RSA scores (the usual map)
                                       │
                                       └───►  F = stack of f_u           ────►  F · Fᵀ
                                                                                ROI×ROI
                                                                                connectivity
```

- A *pair design* says which item pairs you compare and what your model RDMs predict for each pair.
- For each ROI / searchlight, `train_model.rsa_model()` computes the neural pair-dissimilarity vector and regresses it on your model RDMs. With `return_fingerprint = TRUE` it also stores the standardized projection of the neural pair vector onto an orthonormal basis of the model-RDM subspace.
- `model_space_connectivity()` collects those projections into a matrix `F` (ROIs × axes) and reports `F · Fᵀ` plus its standard decompositions.

The reader who internalizes this picture has already learned 80% of what this vignette teaches.

## A first win

We'll use a small synthetic dataset with 80 trials and four ROIs.

```{r toy-data}
set.seed(2026)
toy <- gen_sample_dataset(D = c(6, 6, 6), nobs = 80, blocks = 4, nlevels = 2)
dset <- toy$dataset
nvox <- sum(dset$mask)

# Four arbitrary ROIs over the active voxels.
region_mask <- NeuroVol(
  sample(seq_len(4), size = nvox, replace = TRUE),
  space(dset$mask), indices = which(dset$mask > 0)
)
```

Two model RDMs — say, a "category" RDM and a "feature" RDM:

```{r model-rdms}
n <- 80L
R_cat <- as.matrix(stats::dist(matrix(rnorm(n * 4), n, 4)))
R_feat <- as.matrix(stats::dist(matrix(rnorm(n * 4), n, 4)))
```

A `pair_rsa_design()` is a generalization of `rsa_design()`. For the within-domain case (one item set, lower-triangle pairs) the two are interchangeable; we use `pair_rsa_design()` here to keep one consistent design API throughout the vignette.

```{r pair-design}
des <- pair_rsa_design(
  items_a = paste0("trial", seq_len(n)),
  model   = list(category = R_cat, feature = R_feat)
)
```

Fit the RSA model with fingerprint capture turned on, and run it regionally.

```{r run-regional}
m <- rsa_model(
  dataset = dset, design = des,
  distmethod = "pearson", regtype = "pearson",
  return_fingerprint = TRUE
)

reg <- run_regional(m, region_mask, verbose = FALSE)
```

You already get the standard ROI×metric performance table:

```{r perf-table}
reg$performance_table
```

The new piece is the per-ROI fingerprint matrix, attached as an attribute:

```{r fingerprints}
fp <- attr(reg, "fingerprints")
fp$scores
```

And the connectivity summary is one line:

```{r conn}
conn <- model_space_connectivity(reg)
round(conn$similarity, 3)
```

`conn$similarity` is the ROI-by-ROI representational connectivity through the model-RDM subspace. Diagonal entries are each ROI's own model-explained strength; off-diagonal entries say how much two ROIs share that geometry.

```{r conn-heatmap, echo = FALSE, fig.cap = "ROI-by-ROI model-space connectivity. Diagonal = model-explained strength; off-diagonal = shared geometry through your model RDMs.", out.width = "75%"}
S <- conn$similarity
op <- par(mar = c(3.5, 3.5, 1.5, 4.5))
image(seq_len(nrow(S)), seq_len(ncol(S)), S,
      col = hcl.colors(64, "Blue-Red 3"), axes = FALSE,
      xlab = "", ylab = "")
axis(1, at = seq_len(nrow(S)), labels = rownames(S))
axis(2, at = seq_len(ncol(S)), labels = colnames(S), las = 1)
rng <- range(S, finite = TRUE)
ticks <- pretty(rng, 4)
ticks <- ticks[ticks >= rng[1] & ticks <= rng[2]]
grad <- seq(rng[1], rng[2], length.out = 64)
par(new = TRUE)
plot.new(); plot.window(c(0, 1), rng)
rasterImage(matrix(rev(hcl.colors(64, "Blue-Red 3")), ncol = 1),
            xleft = 0.95, ybottom = rng[1], xright = 1.0, ytop = rng[2])
axis(4, at = ticks, labels = format(ticks, digits = 2), las = 1)
par(op)
```

That's the whole loop. The rest of this vignette unpacks what each piece can do.

## Fingerprints, in detail

Per ROI, the fingerprint is a length-K vector where K is the rank of your model-RDM subspace (here K = 2 because we passed two RDMs). `model_space_connectivity()` standardizes the fingerprints and returns several views of `F · Fᵀ`:

```{r conn-fields}
conn$rank
dim(conn$scores)
names(conn)
```

| Field | Shape | What it is |
|:------|:------|:-----------|
| `scores` | ROIs × axes | The whitened fingerprint matrix `F` |
| `similarity` | ROIs × ROIs | `F · Fᵀ` — strength-sensitive ROI similarity |
| `profile_similarity` | ROIs × ROIs | Cosine of fingerprints — orientation only |
| `component_similarity` | list of ROIs × ROIs | One rank-1 matrix per orthogonal axis |
| `common_similarity` | ROIs × ROIs | The first component (often the shared model axis) |
| `difference_similarity` | ROIs × ROIs | Sum of the rest (differentiating axes) |
| `model_axis_cor` | axes × models | How each axis loads on your original RDMs |

If you only care about *who is similar to whom in geometry*, look at `profile_similarity`. If you care about *who has strong representations of these models*, look at `similarity` and the diagonal.

## Pair designs in two flavors

`pair_rsa_design()` is the unified pair-coordinate system. It supports two modes.

### Within-domain pairs (default)

The classical RSA layout: one item set, lower-triangle pairs. Equivalent to `rsa_design()`:

```{r pair-within-equiv}
classic <- rsa_design(~ R_cat, list(R_cat = R_cat))
paired  <- pair_rsa_design(
  items_a = paste0("trial", seq_len(n)),
  model   = list(R_cat = R_cat)
)

identical(paired$model_mat$R_cat, classic$model_mat$R_cat)
```

Both produce the same length-`n*(n-1)/2` pair vectors and feed into `rsa_model()` interchangeably. You can keep using `rsa_design()` if you prefer; `pair_rsa_design()` is here so the design layer can also handle the cases below.

### Cross-domain (between) pairs

When item set A and item set B differ — for example, items shown in *encoding* vs *retrieval*, or items from *movie A* vs *movie B* — the natural pair space is the rectangular `n_a × n_b` table, not the lower triangle of one square RDM.

```{r pair-between}
n_a <- 6L; n_b <- 8L
items_A <- paste0("A", seq_len(n_a))
items_B <- paste0("B", seq_len(n_b))

# Model: a rectangular cross-set RDM (e.g., similarity in a learned embedding)
M_cross <- matrix(rnorm(n_a * n_b), n_a, n_b)
rownames(M_cross) <- items_A
colnames(M_cross) <- items_B

cross_des <- pair_rsa_design(
  items_a   = items_A,
  items_b   = items_B,
  model     = list(M = M_cross),
  pairs     = "between",
  row_idx_a = seq_len(n_a),                # which dataset rows are in domain A
  row_idx_b = n_a + seq_len(n_b)           # which are in domain B
)

length(cross_des$model_mat$M)              # n_a * n_b — full rectangular block
```

Used with `rsa_model()`, this design causes the engine to compute the *rectangular* neural pair-dissimilarity block — `1 - cor(rows_A, rows_B)` — instead of a lower triangle. You get back per-ROI scores and (with `return_fingerprint = TRUE`) per-ROI fingerprints exactly as in the within-domain case. ROI-to-ROI connectivity through *cross-domain* geometry then drops out of the same `model_space_connectivity()` call.

### Function-valued model entries

Sometimes the model RDM is easier to define as a function of item attributes than as a precomputed matrix:

```{r pair-functional}
items <- c("apple", "pear", "plum", "fig")
features <- data.frame(length = c(5, 4, 4, 3), vowels = c(2, 2, 1, 1))

fdiff <- function(a, b, fa, fb) {
  abs(fa$length - fb$length) + abs(fa$vowels - fb$vowels)
}

fdes <- pair_rsa_design(
  items_a    = items,
  features_a = features,
  model      = list(feat_distance = fdiff)
)
fdes$model_mat$feat_distance
```

The function is evaluated once over the pair table; you don't need to materialize a square matrix.

## Searchlight: the n_centers² problem and how to avoid it

Searchlights have ~10⁵–10⁶ centers. The full center-by-center connectivity matrix is too large to materialize (a 10⁵-center matrix is 80 GB at double precision). Two complementary memory wins are built in.

### 1. Don't store RDMs; store fingerprints

`return_fingerprint = TRUE` reduces each center's neural pair vector — which can be hundreds of thousands of pair cells for many trials — to a length-K fingerprint where K is the rank of the model-RDM subspace (typically 2–10). Memory cost: O(n_centers × K). This is what makes the rest of the searchlight pipeline tractable.

### 2. k-means anchors: a smart-seed summary of `F · Fᵀ`

Instead of materializing the full ROI×ROI matrix, we cluster the per-center fingerprints with k-means and use the searchlight closest to each centroid as an *anchor*. Connectivity becomes an `n_centers × k` similarity matrix plus one brain map per anchor — a "this voxel's representational geometry resembles anchor j" map.

```{r run-searchlight, eval = FALSE}
res_sl <- run_searchlight(m, radius = 6, method = "standard")

conn_sl <- model_space_connectivity(
  res_sl,
  k = 20,             # number of anchors
  scale = "norm",     # cosine similarity
  random_seed = 1L,   # reproducible k-means starts
  build_maps = FALSE  # set TRUE when you want anchor similarity maps
)

names(conn_sl)
```

The returned `model_space_anchor_connectivity` object exposes:

| Field | What it is |
|:------|:-----------|
| `anchors` | Center IDs of the chosen anchor searchlights |
| `cluster_id` | Cluster assignment for every searchlight |
| `centroids` | The k-means centroids in fingerprint space |
| `similarity` | `n_centers × k` matrix of similarities to each anchor |
| `vol_results` | Named list of `NeuroVol` (or `NeuroSurface`) when `build_maps = TRUE` — one map per anchor |

Memory is O(n_centers × k) regardless of the model-RDM dimension or the number of trials.
Use `build_maps = FALSE` while choosing `k` or inspecting clusters; turn maps on
only once you know which anchor summary you want to write or visualize.

If you already have anchor centers in mind — for example, peak coordinates from a prior analysis or specific parcels — pass them with `seeds = c(<center IDs>)` and the function skips k-means entirely:

```{r run-searchlight-explicit, eval = FALSE}
conn_sl <- model_space_connectivity(res_sl, seeds = c(101L, 422L, 1037L))
```

### Choosing `k`

A practical default is `k = 20`. Reasonable lower bound: the rank of your model-RDM subspace. Reasonable upper bound: the number of distinct fingerprint-space "modes" you can interpret. The function clamps `k` automatically when there are fewer distinct fingerprints than requested anchors.

## Object zoo

| Class | Constructor | What it holds |
|:------|:------------|:--------------|
| `pair_rsa_design` | `pair_rsa_design()` | A pair table, model RDM vectors, optional nuisance vectors. Inherits from `rsa_design`. |
| `rsa_model` | `rsa_model(..., return_fingerprint = TRUE)` | A fittable RSA spec that emits fingerprints alongside the standard scores. |
| `regional_mvpa_result` | `run_regional()` | Per-ROI results. Carries fingerprints in `attr(., "fingerprints")` when `return_fingerprint = TRUE`. |
| `searchlight_result` | `run_searchlight()` | Per-center results. Same fingerprint attribute. |
| `rdm_model_space_connectivity` | `model_space_connectivity()` (regional) | ROI×ROI similarity matrix and decompositions. |
| `model_space_anchor_connectivity` | `model_space_connectivity()` (searchlight) | Anchor maps and `n_centers × k` similarity matrix. |

## Where to go next

- For decorrelating correlated model RDMs *before* you run any of the above, see `vignette("RSA")` and `?rdm_decorrelate`.
- For the *learned-feature-space* connectivity path — predicting one ROI's RDM from another via a feature-RSA model — see `vignette("Feature_RSA_Connectivity")`. The two paths are complementary: `model_space_connectivity()` works in your declared model-RDM subspace; `feature_rsa_connectivity()` works in a fitted feature space.
- For variance partitioning by signed contrasts, see `vignette("Contrast_RSA")`.
- For temporal / nuisance pair predictors, see `vignette("Temporal_Confounds_in_RSA")` and `?temporal_rdm`.

## Reproducibility note

All synthetic data in this vignette is generated with explicit seeds. Real datasets will give cleaner-looking similarity matrices because true representational structure dominates noise; the toy figures above are intentionally small so the document builds quickly.
