---
title: Naive Cross-Decoding
author: rMVPA authors
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Naive Cross-Decoding} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-teal');});</script>

---

```{r setup, include=FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(rMVPA)
  library(neuroim2)
})
```

## ReNA / REMAP terminology

This section of the documentation covers four related models for cross-domain representational analysis. They share a vocabulary that's easier to keep straight if defined once:

- **Naive cross-decoding** *(this vignette)* — the no-learning baseline: classify target patterns against source prototypes via correlation. Lives in `naive_xdec_model()`.
- **ERA-RSA** *(`vignette("ERA_RSA_Cross_Decoding")`)* — encoding-retrieval analysis combining first-order item match and second-order RDM similarity. Lives in `era_rsa_model()` and `era_partition_model()`.
- **REMAP-RRR** *(`vignette("REMAP_RRR")`)* — domain-adaptive cross-decoding via reduced-rank regression. Lives in `remap_rrr_model()`.
- **ReNA-Map** *(`vignette("repmap_model")`, `repmap_model()`)* — predicts an ROI's pattern from a low-rank seed feature space.
- **ReNA-RM** *(`vignette("repmed_model")`, `repmed_model()`)* — tests whether one ROI's geometry mediates another's relationship to a seed.
- **ReNA-RC** *(`vignette("repnet_model")`, `repnet_model()`)* — representational connectivity: correlates an ROI's RDM with a seed RDM, with optional confound RDMs.

Reach for naive cross-decoding *first*: it's the cheapest baseline. The other four answer different questions when naive transfer is insufficient.

## Overview

You use cross-decoding when the data come from two related domains, such as
perception and memory, and you want to know whether a pattern learned in the
source domain transfers to the target domain. Naive cross-decoding is the
direct-transfer baseline: it classifies target-domain patterns by correlating
them with source-domain prototypes, with no learned correction between the two
domains.

This makes the analysis deliberately conservative. If it works, the
representations are already aligned well enough to transfer directly. If it
does not, the result is still useful because it tells you how much work a
domain-adaptive model such as `remap_rrr_model()` or an encoding-retrieval
partition analysis has to explain.

**Algorithm:**

1. Compute a prototype (mean pattern) for each category in the training domain.
2. Correlate every test pattern with each prototype.
3. Assign the category whose prototype has the highest correlation.

Correlations are passed through softmax to produce pseudo-probabilities.

## Quick example

```{r basic_example}
set.seed(42)
data_info <- gen_sample_dataset(
  D = c(10, 10, 10), nobs = 80, nlevels = 4,
  blocks = 4, external_test = TRUE
)

# Three ROIs
roi_mask <- NeuroVol(
  sample(1:3, length(data_info$dataset$mask), replace = TRUE),
  space(data_info$dataset$mask)
)

model <- naive_xdec_model(
  dataset = data_info$dataset,
  design  = data_info$design,
  return_predictions = TRUE
)

print(model)
```

```{r run_regional}
results <- run_regional(model, roi_mask)
metric_cols <- intersect(
  c("roinum", "Accuracy", "AUC"),
  names(results$performance_table)
)
results$performance_table[, metric_cols, drop = FALSE]
```

```{r chance}
chance <- 1 / nlevels(data_info$design$y_test)
cat(sprintf("Chance level: %.0f%%\n", chance * 100))
```

```{r check-regional-results, include=FALSE}
stopifnot(
  is.data.frame(results$performance_table),
  nrow(results$performance_table) == 3,
  all(is.finite(results$performance_table$Accuracy)),
  all(results$performance_table$Accuracy >= 0),
  all(results$performance_table$Accuracy <= 1)
)
```

With random data, accuracy hovers around chance. With real fMRI data,
above-chance accuracy indicates that category representations transfer
across domains.

## The `link_by` parameter

By default (`link_by = NULL`), prototypes are keyed by class label:
all training trials for category A are averaged into one prototype, and
each test trial is classified against these category prototypes.

When the same individual items appear in both domains (e.g., the same
images during perception and recall), you can form **item-level
prototypes** instead:

```{r link_by, eval=FALSE}
model <- naive_xdec_model(
  dataset = dataset,
  design  = design,
  link_by = "item_id"
)
```

This creates one prototype per item rather than per category, giving
finer-grained matching when item identity is meaningful.

## How does this relate to ERA partitioning?

Naive cross-decoding answers a first-order question: does each target pattern
look most like the correct source prototype? That is often the first result to
check because it maps directly onto classification accuracy.

The `era_partition_model()` analysis uses the same direct-transfer idea as one
component of a broader encoding-retrieval analysis. It builds matched source
and target prototypes for the item key, then reports:

- direct transfer metrics such as `naive_top1_acc` and `xdec_Accuracy`,
- first-order variance partitioning via `first_order_delta_r2`,
- second-order geometry preservation via `second_order_delta_r2` and
  `geom_cor`, and
- optional leakage-free Procrustes alignment metrics.

Use `naive_xdec_model()` when you need a clean baseline map or ROI table for
direct transfer. Use `era_partition_model()` when your scientific question is
whether encoding-retrieval transfer reflects item-specific similarity,
preserved representational geometry, or nuisance structure such as block,
category, or temporal lag.

## Searchlight analysis

Naive cross-decoding plugs directly into `run_searchlight()`:

```{r searchlight, eval=FALSE}
model <- naive_xdec_model(dataset = dataset, design = design)
sl_results <- run_searchlight(model, radius = 3)
```

The resulting accuracy map shows where category structure transfers
across domains, voxel by voxel.

## Troubleshooting

**"requires external test set"** — Naive cross-decoding needs separate
train and test data. Provide `external_test = TRUE` in
`gen_sample_dataset()`, or construct your `mvpa_dataset()` with both
`train_data` and `test_data`.

**All predictions are the same class** — Prototypes may be too similar.
Check inter-prototype correlations; if all are above ~0.9, the training
domain does not carry enough discriminative information for the method to
work.

**Chance-level performance everywhere** — This is the expected outcome
when there is a large systematic domain shift. Consider domain-adaptive
models such as `remap_rrr_model()` (see `vignette("REMAP_RRR")`).

## Next steps

- `vignette("REMAP_RRR")` — domain-adaptive cross-decoding with low-rank
  corrections
- `vignette("ERA_RSA_Cross_Decoding")` — encoding-retrieval RSA and
  ERA partitioning
- `?naive_xdec_model` — full parameter documentation
