---
title: 'Choose a Safe Parallel Runtime for rMVPA'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: teal
  preset: study
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts
vignette: |
  %\VignetteIndexEntry{Choose a Safe Parallel Runtime for rMVPA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)

library(rMVPA)
library(neuroim2)
set.seed(20260828)
```

```{r albers-classes, echo = FALSE, results = 'asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

A large regional or searchlight analysis can be parallel in three different
ways at once. Future decides **which R processes run tasks**. rMVPA decides
**how image data reach those tasks**. A fitted model or one of its dependencies
may also create **native OpenMP or BLAS threads inside each R process**.

These controls solve different problems. `future::plan(multisession)` does not
activate shard; `backend = "shard"` does not select a Future plan; and
`OMP_NUM_THREADS` is not an rMVPA worker count. Most surprising behavior comes
from changing two or three layers without accounting for their product.

For a conservative local or cluster baseline, use separate R sessions, make
the worker count respect the current allocation, and pin native thread pools
to one before starting R:

```{bash safe-shell, eval = FALSE}
OMP_NUM_THREADS=1 \
OPENBLAS_NUM_THREADS=1 \
MKL_NUM_THREADS=1 \
VECLIB_MAXIMUM_THREADS=1 \
R
```

```{r safe-recipe, eval = FALSE}
run_safely <- function(model_spec, radius = 3) {
  old_plan <- future::plan()
  on.exit(future::plan(old_plan), add = TRUE)

  workers <- max(1L, parallelly::availableCores(omit = 1L))
  future::plan(future::multisession, workers = workers)

  run_searchlight(
    model_spec,
    radius = radius,
    backend = "default"
  )
}
```

That is the portable starting point, not a claim that it is fastest for every
analysis. Measure the real model, region sizes, and batch structure on the
machine where the analysis will run.

## Keep the three layers separate

```{r layer-table, echo = FALSE}
layers <- data.frame(
  layer = c("Process topology", "Data transport", "Native threads"),
  controlled_by = c(
    "future::plan()",
    "backend in run_regional()/run_searchlight()",
    "OMP/BLAS runtime settings"
  ),
  examples = c(
    "sequential, multisession, mirai, multicore",
    "default, shard, auto",
    "OMP_NUM_THREADS, BLAS-specific limits"
  ),
  main_cost = c(
    "process startup, scheduling, R-session memory",
    "ROI extraction, serialization, shared-memory setup",
    "oversubscription and fork safety"
  )
)
knitr::kable(layers, caption = "Three independent runtime layers in rMVPA.")
```

furrr is the mapping layer used by rMVPA's general iterator. It follows the
active Future plan; it is not another process backend. Batch size matters
because a batch is prepared by the controller and then mapped through furrr.
If an analysis creates only one batch, requesting eight workers does not make
eight batches appear.

## Which Future plan should you choose?

```{r plan-table, echo = FALSE}
plans <- data.frame(
  plan = c(
    "sequential", "multisession", "mirai_multisession", "multicore", "callr"
  ),
  execution = c(
    "current R process",
    "persistent independent R sessions",
    "independent mirai daemon processes",
    "forked copies of the current process",
    "fresh callr process per future"
  ),
  use_when = c(
    "debugging, a numerical reference, or one allocated core",
    "portable local/HPC parallelism and the safest default",
    "mirai scheduling is already deployed and measured",
    "Unix-only stack is known to be fork-safe",
    "strong task isolation or prompt worker memory reclamation matters"
  ),
  principal_cost = c(
    "no process parallelism",
    "startup, serialization, and per-session state",
    "optional dependency, startup, transport, and daemon lifecycle",
    "unsafe with some GUIs and multithreaded native runtimes",
    "high startup cost for short futures"
  )
)
knitr::kable(plans, caption = "Future process topologies relevant to rMVPA.")
```

`multisession` is the default recommendation because its workers are separate
R sessions and it is available across platforms. Those sessions load their own
R runtime, packages, model state, and task data, so their memory is not free.
Worker startup can dominate small jobs; reuse a plan for a sequence of analyses
when that is scientifically and operationally appropriate.

`future.mirai::mirai_multisession` also uses separate R processes. It changes
the scheduler and communication mechanism; it is not shared memory. It can be
combined with `backend = "shard"` because the two settings operate at different
layers:

```{r mirai-recipe, eval = FALSE}
run_with_mirai_and_shard <- function(model_spec, region_mask) {
  old_plan <- future::plan()
  on.exit(future::plan(old_plan), add = TRUE)

  future::plan(future.mirai::mirai_multisession, workers = 4)
  run_regional(
    model_spec,
    region_mask,
    backend = "shard"
  )
}
```

`multicore` can start quickly and initially benefits from copy-on-write pages,
but it is not available on Windows and is disabled in some front ends. More
importantly, R Core strongly discourages forked processing when the parent has
used multithreaded libraries: locks and runtime state can be inherited in an
inconsistent form, leading to hangs or crashes. Copy-on-write is a memory
mechanism, not a thread-safety guarantee. Use multicore only as an explicit,
Unix-only experiment after auditing the entire native stack.

## What do default and shard actually move?

The default backend does **not** export the complete rMVPA dataset with every
future. Before furrr sees the model specification, rMVPA removes its dataset.
The controller extracts each requested ROI, however, and the resulting ROI
matrices live in the task frame sent to workers. Large or overlapping ROIs can
therefore create substantial serialization and transient task memory even
though the original dataset was stripped.

Shard changes that transport path. rMVPA builds shared train (and, when
present, test) matrices using POSIX shared memory. Task frames carry voxel or
region indices; each worker maps the shared matrices and extracts the ROI on
demand. Model fitting still allocates ordinary private working memory, and the
controller still owns its original rMVPA objects. Shard therefore reduces data
transport; it does not make the complete analysis zero-copy or eliminate all
per-worker memory.

```{r data-table, echo = FALSE}
data_paths <- data.frame(
  backend = c("default", "shard", "auto"),
  task_payload = c(
    "controller-extracted ROI matrices",
    "ROI indices plus shared-memory handles",
    "shard when setup succeeds; otherwise default"
  ),
  advantages = c(
    "no optional dependency; broad dataset support",
    "small task frames; worker-side extraction; shared source matrix",
    "convenient opportunistic acceleration"
  ),
  tradeoffs = c(
    "serialization and transient ROI payloads",
    "experimental; setup/cleanup; POSIX platform and supported datasets",
    "fallback can hide which path actually ran"
  )
)
knitr::kable(data_paths, caption = "rMVPA data-transport backends.")
```

Use an explicit backend in a production receipt. `auto` is convenient for an
interactive searchlight, but it may fall back to `default` when shard is absent
or the dataset is unsupported. At present shard supports volumetric, surface,
and clustered datasets, but not multibasis image datasets. High-level runners
clean shared segments on normal exit.

```{r shard-recipe, eval = FALSE}
run_with_shared_data <- function(model_spec, radius = 3) {
  if (!requireNamespace("shard", quietly = TRUE)) {
    stop("Install shard before requesting the shared-memory backend.")
  }

  old_plan <- future::plan()
  on.exit(future::plan(old_plan), add = TRUE)
  future::plan(future::multisession, workers = 4)

  run_searchlight(
    model_spec,
    radius = radius,
    backend = "shard"
  )
}
```

Shard is most promising when the source matrices are large, ROIs overlap, or
the same data would otherwise be serialized repeatedly. Its setup cost can
lose on a tiny or single-task analysis. Benchmark both paths with the same
seed, regions, model, workers, and batch size.

## Why can Feature RSA still be slow?

Parallel transport is only one part of `feature_rsa_model()` runtime. For PLS
and PCR, the component selector determines how many numerical fits occur
inside every outer cross-validation fold:

```{r feature-rsa-selection-cost, echo = FALSE}
selection_cost <- data.frame(
  selection = c("blocked", "loo", "pve", "max"),
  approximate_inner_work = c(
    "one fit per training block",
    "one fit per training observation",
    "no validation refits",
    "no validation refits"
  ),
  interpretation = c(
    "block-level held-out error",
    "observation-level held-out error",
    "in-sample explained-variance heuristic",
    "fixed component count"
  )
)
knitr::kable(
  selection_cost,
  caption = "Component-selection work inside each Feature RSA outer fold."
)
```

`"loo"` remains the compatibility default. When rows within runs or sessions
are dependent, `ncomp_selection = "blocked"` is usually both cheaper and a
better match to the independence structure. It requires an observation-aligned
`block_var` in `feature_rsa_design()` and at least two training blocks in every
outer fold. Inner centering and scaling are learned without the held-out block.
`"pve"` and `"max"` are cheaper still, but they change the tuning estimand
rather than accelerating the same validation scheme. See
`vignette("Feature_RSA")` for the selection contract.

rMVPA now streams held-out segment errors, uses compact matrix-level PLS/PCR
fits, and computes the final out-of-fold geometry once. A worker also drops the
original quadratic similarity matrix `S` after its derived feature targets have
been cached, and transports one target matrix plus fold indices instead of
overlapping target slices for every fold. These changes reduce fitting and
serialization overhead for both the default and shard backends. Shard still
matters for the neural source matrix; it does not eliminate private
coefficient, prediction, or validation work inside each worker.

The source benchmark at `inst/benchmarks/feature_rsa_hotpaths.R` validates
prediction and component-selection parity before reporting repeated absolute
times, compact-fit sizes, and worker-payload sizes. Treat its timings as a
machine-specific characterization, not a universal threshold.

## How do OpenMP and BLAS interact?

The current rMVPA C source contains no OpenMP parallel regions. That does not
make every rMVPA analysis single-threaded: an optional classifier, numerical
library, BLAS, or another dependency may use native threads. The active native
code path, rather than a package's mere link to a runtime, determines whether
`OMP_NUM_THREADS` matters.

Set thread limits before R starts. The OpenMP specification defines
`OMP_NUM_THREADS` as an initial control for OpenMP parallel regions; BLAS
implementations have separate controls. Changing a variable after a runtime
has initialized may be too late or runtime-specific.

A useful capacity warning is

\[
  \text{potential runnable threads}
  \approx \text{Future workers} \times
  \text{maximum native threads per worker}.
\]

This is deliberately conservative. OpenMP and BLAS pools need not be nested or
active at the same instant, while some models add another independent pool.
The point is to budget the whole process tree, not to multiply every setting
blindly and call the product a measurement.

Start with one native thread per worker. Increase a native pool only after a
profile shows time in that library, and reduce the Future worker count so the
combined job stays within its CPU and memory allocation. Do not use a few
successful multicore runs with `OMP_NUM_THREADS = 1` as proof of fork safety:
the relevant native runtime may have initialized earlier, or the test workload
may never have entered a parallel region.

## An executable parity check

The following small fixture uses sequential execution as its numerical
reference. It also checks shard when the optional package is installed. This
is a correctness example, not a timing benchmark.

```{r parity-fixture}
example <- gen_sample_dataset(
  D = c(4, 4, 4), nobs = 24, nlevels = 2, blocks = 3
)
crossval <- blocked_cross_validation(example$design$block_var)
model_spec <- mvpa_model(
  model = load_model("sda_notune"),
  dataset = example$dataset,
  design = example$design,
  model_type = "classification",
  crossval = crossval,
  return_predictions = FALSE
)

labels <- integer(length(example$dataset$mask))
inside <- which(as.vector(example$dataset$mask) > 0)
labels[inside] <- rep_len(1:4, length(inside))
region_mask <- NeuroVol(labels, space(example$dataset$mask))

reference <- local({
  old_plan <- future::plan()
  on.exit(future::plan(old_plan), add = TRUE)
  future::plan(future::sequential)
  run_regional(
    model_spec, region_mask,
    backend = "default", preflight = "off"
  )
})

reference$performance_table
```

```{r parity-assertions, include = FALSE}
stopifnot(
  nrow(reference$performance_table) == 4L,
  all(vapply(reference$performance_table, function(x) {
    !is.numeric(x) || all(is.finite(x))
  }, logical(1)))
)

if (requireNamespace("shard", quietly = TRUE)) {
  shared <- local({
    old_plan <- future::plan()
    on.exit(future::plan(old_plan), add = TRUE)
    future::plan(future::sequential)
    run_regional(
      model_spec, region_mask,
      backend = "shard", preflight = "off"
    )
  })
  stopifnot(
    identical(names(shared$performance_table),
              names(reference$performance_table)),
    isTRUE(all.equal(shared$performance_table,
                     reference$performance_table,
                     tolerance = 1e-12,
                     check.attributes = FALSE))
  )
}
```

For parallel validation, keep the fixture fixed and compare the complete
canonical result vector, not just its length, class, or average accuracy. Run
each thread configuration in a fresh process with a timeout so a low-level
hang becomes an observed `timeout` rather than a stalled notebook.

## What did the benchmark show?

The checked-in benchmark used a 72-observation, `28 x 28 x 28` regional
classification fixture with 18 regions, five batches of at most four regions,
and `sda_notune`. Non-sequential plans configured two workers. OpenMP and BLAS
were fixed at one thread. Each scheduler/data-backend pairing ran in three
fresh R processes from a clean installed rMVPA 0.1.3 tarball.

```{r benchmark-receipt, echo = FALSE}
receipt_path <- system.file(
  "extdata", "parallel_runtime_benchmark_results.csv",
  package = "rMVPA"
)
if (!nzchar(receipt_path)) {
  receipt_path <- file.path(
    "..", "inst", "extdata", "parallel_runtime_benchmark_results.csv"
  )
}
parallel_receipt <- utils::read.csv(receipt_path, stringsAsFactors = FALSE)

stopifnot(
  nrow(parallel_receipt) == 8L,
  all(parallel_receipt$repetitions == 3L),
  all(parallel_receipt$n_batches == 5L),
  all(parallel_receipt$omp_threads == 1L),
  all(parallel_receipt$blas_threads == 1L),
  max(parallel_receipt$max_abs_result_error) <= 1e-12,
  all(parallel_receipt$frame_bytes_sum_median[
    parallel_receipt$data_backend == "shard"
  ] < parallel_receipt$frame_bytes_sum_median[
    parallel_receipt$data_backend == "default"
  ])
)

# Every claim the prose makes about this table is derived from it here, so a
# regenerated receipt cannot leave the surrounding sentences asserting the
# ordering of a previous run.
oxford <- function(x) {
  if (!length(x)) return("no plan")
  if (length(x) == 1L) return(x)
  if (length(x) == 2L) return(paste(x, collapse = " and "))
  paste0(paste(x[-length(x)], collapse = ", "), ", and ", x[[length(x)]])
}

shard_summary <- local({
  defaults <- parallel_receipt[parallel_receipt$data_backend == "default", ]
  shards <- parallel_receipt[parallel_receipt$data_backend == "shard", ]
  defaults <- defaults[order(defaults$future_backend), ]
  shards <- shards[order(shards$future_backend), ]
  plan <- sub("_multisession$", "", defaults$future_backend)
  faster <- shards$analysis_median_seconds < defaults$analysis_median_seconds
  overlap <- pmin(defaults$analysis_max_seconds, shards$analysis_max_seconds) >=
    pmax(defaults$analysis_min_seconds, shards$analysis_min_seconds)
  list(
    faster_clause = if (all(faster)) {
      "every plan measured"
    } else oxford(plan[faster]),
    slower_clause = if (any(!faster)) {
      paste0(", and raised it for ", oxford(plan[!faster]))
    } else "",
    overlap_clause = if (all(overlap)) {
      "overlap for every pairing"
    } else if (!any(overlap)) {
      "are separated for every pairing"
    } else {
      sprintf("overlap for %d of the %d pairings", sum(overlap), length(overlap))
    },
    default_bytes = format(defaults$frame_bytes_sum_median[[1L]],
                           big.mark = ",", scientific = FALSE),
    shard_bytes = format(shards$frame_bytes_sum_median[[1L]],
                         big.mark = ",", scientific = FALSE),
    fold = round(median(shards$default_to_shard_frame_sum_ratio))
  )
})

display_receipt <- transform(
  parallel_receipt,
  plan = sub("_multisession$", "", future_backend),
  backend = data_backend,
  median_seconds = round(analysis_median_seconds, 3),
  range_seconds = sprintf("%.3f--%.3f", analysis_min_seconds,
                          analysis_max_seconds),
  task_frame_MiB = round(frame_bytes_sum_median / 1024^2, 3)
)[, c("plan", "backend", "median_seconds", "range_seconds",
      "task_frame_MiB")]

knitr::kable(
  display_receipt,
  col.names = c(
    "Future plan", "Data backend", "Median (s)", "Range (s)",
    "Task frames (MiB)"
  ),
  caption = paste(
    "Local fresh-process characterization on an Apple M3 Max (36 GB).",
    "Timings are descriptive, not thresholds."
  )
)
```

All eight paths produced identical canonical result vectors in every fixture
repetition (`max_abs_result_error = 0`). The default frames totaled
`r shard_summary$default_bytes` bytes per run; shard frames totaled
`r shard_summary$shard_bytes` bytes, a `r shard_summary$fold`-fold reduction
for this design. This is direct evidence about the transport structure measured
by `object.size()`, not a claim that physical RAM fell by exactly the same
factor.

Shard reduced the median time for `r shard_summary$faster_clause` in this short
local run`r shard_summary$slower_clause`, and the three-repetition timing ranges
`r shard_summary$overlap_clause`. Startup dominates the separate-session plans.
Read that ordering as one machine on one afternoon rather than a property of
the backends: multicore's local timing does not override its safety
constraints, and mirai's fresh-process overhead here does not predict a
long-lived mirai deployment. Re-run the driver on the target system before
choosing on speed.

The receipt intentionally omits process-tree RSS as a publication claim.
Summed RSS can double-count shared and copy-on-write pages, while socket and
mirai workers may be detached from the controller's descendant tree. Frame
payload size, exact output parity, and separately reported timing phases were
the stable measurements.

The reproducibility driver is installed at
`inst/benchmarks/parallel_runtime_benchmark.R`; the more general grid and its
field definitions are in `scripts/sweep_parallel_runtime_grid.R` and
`scripts/README_parallel_runtime_grid.md` in the source repository.

## Practical decisions

Use this order of operations:

1. Run the complete analysis sequentially on a small fixture and save its
   canonical result.
2. Determine the CPU and memory allocation with
   `parallelly::availableCores()` and the scheduler, not the host's advertised
   core count.
3. Select `multisession`, start with native thread pools at one, and verify
   that the intended number of batches and workers actually ran.
4. Compare `default` and `shard` explicitly. Keep shard only if its smaller
   task frames or elapsed time matter for the real workload.
5. For PLS/PCR Feature RSA, choose the component-selection estimand before
   tuning workers. Prefer `"blocked"` when the block is the independence unit;
   exact `"loo"` requires one inner fit per training observation.
6. Consider mirai for its scheduler and lifecycle properties, not as a synonym
   for shared memory.
7. Increase one axis at a time. Recheck complete numerical results after every
   change.
8. Use multicore only when the full native stack is known to be fork-safe; do
   not combine it casually with OpenMP or another multithreaded runtime.

The underlying runtime references are the
[Future plan documentation](https://future.futureverse.org/reference/plan.html),
[Future multicore warning](https://future.futureverse.org/reference/multicore.html),
[future.mirai documentation](https://future.mirai.futureverse.org/),
[furrr gotchas](https://furrr.futureverse.org/articles/gotchas.html),
[parallelly core detection](https://parallelly.futureverse.org/reference/availableCores.html),
[R Core fork guidance](https://stat.ethz.ch/R-manual/R-devel/library/parallel/help/mcfork.html),
and the [OpenMP environment-variable specification](https://www.openmp.org/spec-html/5.0/openmpch6.html).
