---
title: "Confirming a frozen pattern model"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: |
  %\VignetteIndexEntry{Confirming a frozen pattern model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 6, fig.height = 3.5)
library(rMVPA)
set.seed(505)
```

This is the second pattern-model article. The first,
`vignette("Pattern_Model")`, chooses rank, penalties, and a residual
covariance on discovery data. Confirmation asks a narrower question: do
those **already chosen directions** still predict brain measurements in
independent rows?

It estimates fresh, unpenalized loadings while freezing the target coding,
whitening, and component directions learned during discovery. That
separation permits t and omnibus tests without treating a selected,
regularized discovery loading as an ordinary regression coefficient. It
does not turn cross-validated rank selection into a test of population
rank. Pooling confirmed subject loadings is `vignette("Pattern_Group")`.

## Freeze a discovery fit

The example has two continuous targets and 32 features. The first half of the
features responds to the first target; the second half responds to the second.
Shared run noise makes rows within a run dependent.

```{r data}
p <- 32
A <- cbind(rep(c(1, 0), each = 16), rep(c(0, 1), each = 16))
make_sample <- function(n = 120, runs = 12) {
  Y <- cbind(speed = rnorm(n), difficulty = rnorm(n))
  run <- rep(seq_len(runs), length.out = n)
  X <- Y %*% t(A) + matrix(rnorm(n * p), n, p) + rnorm(runs)[run]
  list(X = X, Y = Y, run = run)
}
discovery <- make_sample(runs = 4)
space <- neuroim2::NeuroSpace(c(4, 4, 2), c(1, 1, 1))
mask <- neuroim2::NeuroVol(array(1, c(4, 4, 2)), space)
images <- neuroim2::NeuroVec(array(t(discovery$X), c(4, 4, 2, 120)),
                            neuroim2::NeuroSpace(c(4, 4, 2, 120), c(1, 1, 1)))
dataset <- mvpa_dataset(images, mask = mask)
design <- mvpa_design(data.frame(y = discovery$Y[, 1], run = discovery$run),
                      y_train = ~ y, block_var = ~ run, targets = discovery$Y)
fit <- run_global(pattern_model(dataset, design, rank = 2), refit = TRUE)$refit
basis <- pattern_basis(fit)
round(basis$matrix, 3)
```

The rows of `basis$matrix` are named raw targets and its columns are frozen
scores. This matrix includes the training scaling and whitening; storing `C`
alone would lose that information. Independent display rotations from
`rotate_patterns()` do not redefine this inference basis.

## Estimate and test loadings on new rows

Use globally meaningful observation IDs, including the participant, session,
and observation identity in a real study. Declare **every** discovery row used
for preprocessing, tuning, selection, or fitting. Prefixing the same rows with
new names would defeat the check; IDs cannot establish independence by themselves.

```{r confirm}
heldout <- make_sample()
discovery_ids <- paste0("participant01:discovery:", seq_len(120))
confirmation_ids <- paste0("participant01:confirmation:", seq_len(120))
confirmation <- pattern_confirm(
  fit, heldout$X, heldout$Y,
  block_var = heldout$run,
  inference = confirmation_plan("sign_flip", n_resamples = 999, seed = 505),
  observation_ids = confirmation_ids,
  discovery_ids = discovery_ids,
  feature_ids = paste0("voxel", seq_len(p)),
  preprocessing_id = "simulated-BOLD-original-units-v1",
  subject_id = "participant01"
)
head(data.frame(feature = confirmation$feature_ids,
                F = confirmation$omnibus$F,
                p_max = confirmation$bootstrap$p_omnibus_max))
```

Each row tests whether **all** frozen score coefficients at that feature are
zero, conditional on any nuisance columns. Loadings and their standard errors
are in original measurement units. All input features enter the hypothesis
family, including ones screened or penalized to zero during discovery.

`confirmation$estimate`, `$se`, `$t`, and `$p` contain the component results.
`$p_holm` adjusts over every feature-component pair. The omnibus table has its
own Holm correction over features. Neither family correction simultaneously
covers every alternative analysis a researcher might try.

```{r confirmed-map, fig.alt="Omnibus confirmation statistics across 32 simulated signal features."}
plot(confirmation$omnibus$F, type = "h", xlab = "Input feature",
     ylab = "Omnibus Wald F")
```

```{r checks, include=FALSE}
stopifnot(all(is.finite(confirmation$estimate)),
          all(confirmation$omnibus$df2 == 11),
          mean(confirmation$bootstrap$p_omnibus_max < 0.05) > 0.75)
```

### Choose the error model before inspecting results

- `confirmation_plan("independent")` uses ordinary Gaussian t/F tests with
  residual degrees of freedom. It assumes independent, homoskedastic rows.
- `confirmation_plan("block_robust")` uses a CR1 sandwich and approximate
  t/Wald F inference with the number of independent blocks minus one as the
  denominator degrees of freedom. Few blocks can give unreliable inference. In the bundled 30-block null
  simulation, this approximation rejected 8.4% of nominal 5% omnibus tests;
  it should not be treated as exactly calibrated.
- `confirmation_plan("sign_flip", n_resamples = 999, seed = 1)` refits
  restricted residuals using one Rademacher sign per independent block. This
  is an **approximate restricted wild bootstrap**, not an exact permutation
  test with estimated nuisance effects. `$bootstrap$p_max` maximizes across
  features within each component and Bonferroni-adjusts across components;
  `$bootstrap$p_omnibus_max` maximizes omnibus statistics across features.
  The same signs are used across features, preserving their joint dependence.
  The bundled null simulation rejected 4.1% of nominal 5% omnibus tests
  with 199 draws; this simulation is evidence for that design, not a guarantee
  for every block structure. Use more draws for final analysis.

The [FSL randomise documentation](https://fsl.fmrib.ox.ac.uk/fsl/docs/statistics/randomise.html)
also distinguishes exact exchangeability from approximate residual-based
resampling with nuisance effects. No residual-resampling option repairs an
incorrect independence or exchangeability assumption. A Monte Carlo p value
cannot be smaller than `1 / (n_resamples + 1)`. For large feature families,
Holm correction can therefore be unable to reject anything at the chosen
resolution. Use the stated maximum-statistic family, or budget enough draws
for the intended marginal-p correction; inspect Monte Carlo uncertainty near
a decision threshold.

`block_var` identifies independent units; it does **not** add run intercepts.
For that estimand, supply, for example,
`nuisance = model.matrix(~ factor(heldout$run))[, -1, drop = FALSE]`.
An intercept is always included. Redundant nuisance columns or target scores
that cannot be distinguished from nuisance effects cause an error. Nonuniform
row weights are unsupported. No p value is returned for a singular sampling
distribution.

## Separate association from incremental prediction

A decoded component can correlate with its target without improving prediction
once other components are available. The two questions need different tests.

```{r components}
components <- pattern_component_tests(
  fit, confirmation, heldout$X, heldout$Y,
  calibration = list(dataset = discovery$X, design = discovery$Y,
                     observation_ids = discovery_ids)
)
components$association
components$incremental
```

Association is the partial correlation between each frozen decoded score and
its corresponding target score, adjusting for the confirmation nuisance
columns. It is marginal with respect to the other components.

For incremental value, full and reduced **linear decoding heads** are fitted
on calibration rows only. Removing a component refits all remaining head
coefficients. Both heads predict the untouched confirmation rows. Positive
`gain` means that removing the component increases squared error. Loss is
averaged over raw target columns, then within each block, then equally across
blocks; choose commensurate target units before discovery. A paired block-mean
t test provides approximate inference. For a sign-flip plan it is replaced
by an approximate centered, studentized block wild bootstrap.

These comparisons evaluate the separately calibrated linear heads, not the
original Gaussian posterior decoder. Categorical targets use one-hot squared
loss, and linear head outputs are not probabilities. The confirmation object
also includes descriptive held-out performance of the original decoder in
`$prediction`.

Data and fit hashes prevent accidentally reusing a confirmation record with a
different fit or different rows. Calibration IDs must belong to the declared
discovery set and cannot overlap confirmation IDs. A completely untouched
calibration split may be included in that declared set before confirmation.

## Preserve the inferential record

```{r save, eval=FALSE}
saveRDS(list(confirmation = confirmation, components = components),
        "participant01-confirmation.rds")
```

The confirmation object retains the frozen raw target basis, coefficient
covariance, preprocessing identifier and hash, observation IDs, block labels,
and nuisance specification. Independent-error covariance is factorized into a
small design covariance and per-feature residual variances. Block methods
retain a small component-by-component covariance at each feature. Neither
allocates a feature-by-feature covariance matrix.

Individual component effects depend on the frozen coordinates. Omnibus tests
are invariant to nonsingular changes of score coordinates. Confirmation of
several components is not a sequential rank test: `rank_supported` remains a
separate research extension with a separately validated null.

To pool these subject-level confirmations in a shared target basis, continue
with `vignette("Pattern_Group")`.
