---
title: "Fit a whole-brain pattern model"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: |
  %\VignetteIndexEntry{Fit a whole-brain pattern model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 6, fig.height = 3.5)
library(rMVPA)
set.seed(904)
```

A pattern model answers a joint question: **can the whole feature domain
decode the task, and where is that information expressed?** It fits one
low-rank forward model

\[
x = A C^{\top} y + \varepsilon, \qquad \varepsilon \sim (0, \Psi)
\]

and then reuses that fit for classification or multivariate decoding,
encoding, forward patterns, and restricted regional prediction. You do not
fit one model to predict labels and another to draw a map.

This is the first of three pattern-model articles. Confirmation of a frozen
fit is `vignette("Pattern_Confirmation")`; pooling confirmed subject loadings
is `vignette("Pattern_Group")`.

## When to use it

Use `pattern_model()` when you want **one covariance-aware fit** over a
whole domain (or a large ROI) and several derived answers from it:

- held-out classification or multi-response decoding
- model-implied forward patterns, not only decoding weights
- how much of the whole-domain prediction remains accessible inside a region
- later confirmation or group analysis of those patterns

Use a different family when the question is narrower:

| Question | Model family |
|---|---|
| Decode feature vectors inside independently fitted regions | `feature_rsa_model` (`vignette("Feature_RSA")`) |
| Voxelwise encoding from grouped predictors | `banded_ridge_model` (`vignette("Banded_Ridge_Encoding")`) |
| Map between perception and retrieval domains | `remap_rrr_model` (`vignette("REMAP_RRR")`) |
| Find locally predictive neighborhoods with separate local fits | searchlight analysis (`vignette("Searchlight_Analysis")`) |

The current residual covariance is diagonal plus a few shared noise
components. It can miss richer *local* noise correlations. Spatial
penalties act on the forward patterns \(A\), not on decoding weights.

## What you pass in and get back

| Object | Role |
|---|---|
| `mvpa_dataset` | Brain measurements. Rows are observations; columns are features. |
| design with `model_targets()` | Categorical labels (`mvpa_design`) or a numeric target matrix (`targets`, `feature_sets_design`, or `feature_rsa_design`). |
| `pattern_model()` | The specification: rank, noise, optional spatial penalty, optional observation weights. |
| `run_global()` | One domain-wide evaluation. Same specification also works with `run_regional()` and `run_searchlight()`. |
| `pattern_global_result` | Held-out `performance_table`, prediction ledger, optional fold fits, optional full-data `refit`. |

The usual next calls on that result are `performance()`,
`model_importance()` / `model_patterns()` on `$refit`,
`local_performance()`, and `predict()` on a retained `pattern_fit`.

## Plant four territories

The example has 120 observations in three runs and 48 voxels. Class labels
cycle through `a`, `b`, and `c`. Four contiguous territories are planted:

- **signal 1** (voxels 1–12): an A-versus-B code, plus a shared noise
  process
- **signal 2** (voxels 13–24): a B-versus-C code
- **shared noise** (voxels 25–36): the same noise process, no task loading
- **background** (voxels 37–48): independent noise only

The shared-noise territory is a *noise canceller*. Observing it can improve
whole-domain decoding even when its forward loading is near zero. That is
why the guide keeps forward patterns and decoding weights as separate
outputs.

```{r data}
n <- 120
p <- 48
class <- factor(rep(c("a", "b", "c"), length.out = n))
run <- rep(1:3, each = 40)
code <- model.matrix(~ class - 1)
X <- matrix(rnorm(n * p), n, p)
shared_noise <- rnorm(n, sd = 2)
X[, 1:12]  <- X[, 1:12]  + 1.2 * (code[, 1] - code[, 2]) + shared_noise
X[, 13:24] <- X[, 13:24] + 1.2 * (code[, 2] - code[, 3])
X[, 25:36] <- X[, 25:36] + shared_noise

space <- neuroim2::NeuroSpace(c(4, 4, 3), c(1, 1, 1))
mask <- neuroim2::NeuroVol(array(1, c(4, 4, 3)), space)
images <- neuroim2::NeuroVec(
  array(t(X), c(4, 4, 3, n)),
  neuroim2::NeuroSpace(c(4, 4, 3, n), c(1, 1, 1))
)
dataset <- mvpa_dataset(images, mask = mask)
design <- mvpa_design(data.frame(class, run), y_train = ~ class,
                      block_var = ~ run)
```

## Fit the domain and score held-out runs

```{r fit}
spec <- pattern_model(dataset, design, rank = 2,
                      noise = list(type = "diag_lowrank", rank = 1))
result <- run_global(spec, return_fits = TRUE, refit = TRUE)
performance(result)
```

`performance()` is leave-one-run-out accuracy. Chance is one third.
`rank_mean` is the mean rank selected across outer folds; here rank is
fixed at 2, so the column is exactly 2.

Two fits are retained for different jobs:

- **fold fits** (`return_fits = TRUE`) are the cross-validated estimators.
  `local_performance()` and the Haufe diagnostic need them.
- **`$refit`** uses every training row. It is the descriptive map and the
  object you would confirm later. It is **not** the fit that produced the
  accuracy table.

`return_fits = TRUE` costs memory, especially with many resamples.

```{r predict}
X_all <- get_feature_matrix(dataset)
head(predict(result$refit, X_all, type = "prob"))
```

`predict()` on a `pattern_fit` returns class probabilities, class labels,
decoded continuous targets, calibrated scores, or encoded brain
measurements. The probabilities above come from the descriptive refit, so
they are not the held-out scores in `performance()`.

## Read where the signal lives

```{r maps, fig.alt="Model-implied signal standard deviation and Gaussian conditional information across 48 input voxels, with planted territory boundaries."}
signal_sd <- model_importance(result$refit, type = "signal_sd")
information <- model_importance(result$refit, type = "conditional_info")
oldpar <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (panel in list(
  list(y = signal_sd, ylab = "Signal SD"),
  list(y = information, ylab = "Conditional information (nats)")
)) {
  plot(panel$y, type = "h", xlab = "Input voxel", ylab = panel$ylab)
  abline(v = c(12.5, 24.5, 36.5), lty = 2, col = "grey70")
  mtext(c("S1", "S2", "noise", "bg"), side = 3, line = 0,
        at = c(6, 18, 30, 42), cex = 0.8)
}
par(oldpar)
```

**Signal SD** is \(\sqrt{\mathrm{diag}(A\Phi A')}\) in the original
measurement units. It says where fitted task variance is expressed. In this
realization the B-versus-C territory (S2) is strongest, the A-versus-B
territory (S1) is next, and background is near zero. The shared-noise
territory is not exactly zero: finite-sample loadings rarely vanish without
a sparsity penalty.

**Conditional information** asks what a voxel adds once the other voxels
have been observed. It can be high in a noise canceller whose signal SD is
modest. It is a working Gaussian score-model quantity, in nats — not
empirical mutual information about class labels, and not a significance
map.

`model_importance(result)` builds a spatial image through the dataset.
`model_importance(result$refit)` returns a vector aligned to input columns;
screened columns are `NA`. Component-wise forward patterns and calibrated
decoder weights come from `model_patterns(fit, "forward")` and
`model_patterns(fit, "weights")`. Those columns depend on the chosen
component coordinates. Scalar maps do not.

## Ask what each region can decode

`local_performance()` keeps the whole-domain task representation and
restricts only the residual covariance to the region. It then predicts the
**same held-out observations** used for the whole-domain score. Cropping
whole-brain decoding weights would keep noise-cancellation terms that
require unobserved voxels; that is a different, generally incorrect,
estimand.

```{r local}
regions <- list(
  signal_1     = 1:12,
  signal_2     = 13:24,
  shared_noise = 25:36,
  background   = 37:48,
  whole_domain = 1:48
)
local <- local_performance(result, regions)
local[local$metric == "Accuracy", ]
```

Each name in `regions` is a set of **input-matrix column positions**, not
voxel IDs.

In this realization the table matches the planting:

- **signal 2** decodes well on its own.
- **signal 1** is near chance by itself. It carries both the A-versus-B
  code and the shared noise, and without the canceller it cannot separate
  them.
- **shared noise** is also near chance alone, as it should be: it has no
  task loading.
- **whole domain** recovers the high held-out accuracy. The canceller is
  useful only when the signal territories are observed with it.

The regional score is access under the shared whole-brain representation,
not the best accuracy a separately trained local model could reach. A
finite-sample region can beat the whole-domain number. For a true
independent-ROI comparison, pass fold-resolved ledgers through
`independent_roi`; those ledgers must match rows, folds, truth, and
baselines.

```{r locality-check, include=FALSE}
acc <- local[local$metric == "Accuracy", ]
row_of <- function(name) acc[acc$region == name, ]
stopifnot(
  isTRUE(all.equal(row_of("whole_domain")$whole_brain,
                   row_of("whole_domain")$local_restricted)),
  row_of("signal_2")$local_restricted > 0.8,
  row_of("signal_1")$local_restricted < 0.5,
  row_of("shared_noise")$local_restricted < 0.5,
  row_of("background")$local_restricted < 0.5,
  mean(information[25:36]) > mean(information[37:48]),
  mean(signal_sd[13:24]) > mean(signal_sd[37:48])
)
```

## Decode continuous feature targets

The same constructor accepts a numeric target matrix. The example below
encodes two continuous features into the first two territories.

```{r continuous}
features <- matrix(rnorm(n * 2), n, 2,
                   dimnames = list(NULL, c("feature_1", "feature_2")))
brain <- matrix(rnorm(n * p), n, p)
brain[, 1:12]  <- brain[, 1:12]  + features[, 1]
brain[, 13:24] <- brain[, 13:24] + features[, 2]
continuous_data <- mvpa_dataset(
  neuroim2::NeuroVec(
    array(t(brain), c(4, 4, 3, n)),
    neuroim2::NeuroSpace(c(4, 4, 3, n), c(1, 1, 1))
  ),
  mask = mask
)
continuous_design <- mvpa_design(
  data.frame(id = 1:n, run),
  cv_labels = 1:n, targets = features, block_var = ~ run
)
continuous <- run_global(
  pattern_model(continuous_data, continuous_design, rank = 2),
  return_fits = TRUE
)
performance(continuous)
local_performance(continuous, list(first = 1:12, second = 13:24))
```

`R2` is predictive R-squared against each fold's **training-target mean**,
not squared correlation, and it can be negative. The reported value
averages per-response R-squared. Targets are centred and whitened on
training rows only; `predict(..., type = "decode")` returns original
target units. Here each territory reconstructs its own feature only
partly; the whole domain does much better because both features are
observed together.

`feature_sets_design` is the other common continuous-target path. Its
`row_weights` are used by the estimator (see below).

```{r continuous-check, include=FALSE}
clocal <- local_performance(continuous, list(first = 1:12, second = 13:24))
r2 <- clocal[clocal$metric == "R2", ]
stopifnot(
  r2$whole_brain[1] > 0.7,
  all(r2$local_restricted < r2$whole_brain)
)
```

## Control rank, penalties, and weights

**Rank.** `"auto"` (the default) selects rank inside each outer training
split by held-out decoding loss, then reports `rank_mean`. A fixed integer
is used as given and still capped at the eligible rank (number of classes
minus one for categorical targets).

**Spatial penalties** act on \(A\). They are optional and they are
hypotheses about anatomy, not defaults.

```{r penalty, eval=FALSE}
# Localized support: a feature is in the patterns or out of them.
pattern_model(dataset, design, rank = 2, penalty = list(sparse = 0.1))

# Neighbouring voxels should carry similar signed loadings.
# A spatial_graph() is built from the dataset when one is not supplied.
pattern_model(dataset, design, rank = 2, penalty = list(signed_smooth = 1))
```

`sparse` is a fraction of the penalty that empties the model, so `0.1`
means the same thing across folds and domains. `signed_smooth` is the
weight of a graph-Laplacian term relative to the data-fit curvature.
Either entry may be `"auto"`, which cross-validates a short path jointly
with rank. Do **not** default to `sparse = "auto"` on a dense or weak
whole-brain code: inner folds can pick a penalty that hurts held-out
accuracy. Use sparsity when you believe support is localized.
`support_smooth` is reserved for a later envelope penalty and is rejected
rather than quietly remapped.

**Observation weights** enter the fit itself: centring, target whitening,
the residual covariance, the objective, and the inner tuning loss. They
do not reweight the reported metrics — every tested observation still
counts once.

```{r weights, eval=FALSE}
# Explicit weights override design-carried row_weights.
# A zero weight drops that row from training, exactly.
pattern_model(dataset, design, rank = 2, weights = w)
```

If the design is a `feature_sets_design`, its `row_weights` are used
automatically. Integer weights are equivalent to replicating rows.
Confirmation (`vignette("Pattern_Confirmation")`) does not accept
nonuniform weights.

## Rotate a display and compare folds

```{r rotation}
view <- rotate_patterns(result$refit)
view$reconstruction_error
result$component_stability[, c("fold1", "fold2", "rank1", "rank2",
                               "n_common", "overlap")]
```

`rotate_patterns()` changes display coordinates while preserving
\(L_b H L_t' = A C'\). The underlying fit is unchanged, so predictions and
scalar maps stay bit-identical. Only orthogonal rotations are supported.
Compare patterns across folds or subjects with the column space, the
support, or a coordinate-invariant map — never raw entries of \(A\).

`component_stability` compares spatial column spaces with principal angles
on features retained in both folds. An overlap of one means the two
nonempty subspaces coincide. Unequal ranks reduce overlap even if the
smaller space sits inside the larger. This is a subspace diagnostic, not a
test that a particular rank is supported.

```{r haufe}
vapply(result$haufe_diagnostics, function(x) x$relative_discrepancy,
       numeric(1))
```

The Haufe diagnostic compares empirical held-out covariance patterns with
the model-implied patterns. Large discrepancies can flag sampling noise or
misspecification; they are not *p*-values. Automatic results use
`train:<row ID>` for training and CV rows and `test:<row ID>` for an
external test partition. Independence of an external partition is a caller
obligation. For a separately held-out matrix, call
`pattern_haufe(fit, X_holdout, observation_ids)`.

## Save and continue

```{r save, eval=FALSE}
save_results(result, "pattern-output")
```

The RDS stores the result, including retained fits and ledgers. Spatial
refit maps are also written as images for supported datasets. Multibasis
image aggregation is refused because averaging channels would change the
estimand; those results save as RDS.

To test the frozen discovery directions on new rows, continue with
`vignette("Pattern_Confirmation")`. To pool confirmed subject loadings in
a shared target basis, use `vignette("Pattern_Group")`.
