---
title: "Group patterns in shared target coordinates"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: |
  %\VignetteIndexEntry{Group patterns in shared target coordinates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 6, fig.height = 3.5)
library(rMVPA)
set.seed(606)
```

This is the third pattern-model article. Fit a discovery model with
`vignette("Pattern_Model")`, confirm each subject on independent rows with
`vignette("Pattern_Confirmation")`, then pool those confirmed loadings here.

Group analysis starts from independently estimated subject loadings and
their uncertainty. Both axes of every loading matrix must already mean the
same thing: features correspond spatially, and target coordinates represent
the same task quantities. Similar-looking component columns are not enough.

`pattern_group()` transports each subject's estimates **and full coefficient
covariance** into a reference target basis, then estimates the group effect.
An orthogonal rotation of the reference changes component columns but preserves
omnibus tests, effect norms, and descriptive subject expression.

## Create independent subject confirmations

This runnable example uses one independently fitted discovery basis for all
subjects. A simulation then generates 12 subjects with Gaussian variation
around the discovery loading matrix. Conditioning on that fixed discovery fit
makes the population loading matrix known for this example.

```{r discovery}
ds <- gen_sample_dataset(c(4, 4, 3), 72, nlevels = 3, blocks = 3)
fit <- run_global(pattern_model(ds$dataset, ds$design, rank = 2),
                  refit = TRUE)$refit
reference <- pattern_basis(fit)
population <- model_patterns(fit, type = "forward")
p <- nrow(population)
feature_ids <- paste0("shared-voxel", seq_len(p))
discovery_ids <- paste0("external-discovery:", seq_len(72))
```

The same discovery cohort may supply a basis to multiple subjects. None of its
observations may appear in any subject's confirmation data. In a real study,
subject-specific discovery fits are also allowed if their raw target matrices
span exactly the reference subspace. Full-rank two-component models for the
same three class contrasts, for example, can differ in whitening and rotation
while representing the same target subspace.

```{r subjects}
labels <- factor(rep(reference$target_ids, each = 30),
                 levels = reference$target_ids)
code <- model.matrix(~ labels - 1)
colnames(code) <- reference$target_ids
scores <- sweep(code, 2, reference$center, "-") %*% reference$matrix
subjects <- lapply(seq_len(12), function(s) {
  subject_pattern <- population + matrix(rnorm(p * 2, sd = 0.2), p, 2)
  X <- scores %*% t(subject_pattern) + matrix(rnorm(90 * p), 90, p)
  pattern_confirm(
    fit, X, labels,
    inference = confirmation_plan("independent"),
    observation_ids = paste0("subject", s, ":confirmation:", seq_len(90)),
    discovery_ids = discovery_ids,
    feature_ids = feature_ids,
    preprocessing_id = "simulated-BOLD-original-units-v1",
    subject_id = paste0("subject", s)
  )
})
```

For real run-dependent observations, choose a suitable block error model in
each subject's confirmation; see `vignette("Pattern_Confirmation")`. A group
model does not repair biased subject estimates or discovery/confirmation
leakage.

## Estimate the population mean

```{r group}
group <- pattern_group(subjects, reference_basis = reference, effects = "random")
group
head(group$omnibus)
```

The random-effects estimand is the **equally weighted subject population mean**.
Its covariance is the sample covariance of subject coefficient vectors divided
by the number of subjects. This includes both between-subject variation and
subject estimation error. Adding the within-subject variance again would count
sampling error twice.

The omnibus test is Hotelling's T-squared expressed as an F statistic. With
12 subjects and two target dimensions it uses F(2, 10), regardless of how many
rows each subject contributed. It is exact for iid Gaussian subject estimates
with common total covariance, and approximate with heterogeneous precision or
nonnormal effects. More than rank plus one subjects are required; singular
covariance produces unavailable omnibus inference, not an invented zero effect.

Component t tests in `$p` have 11 degrees of freedom here. `$p_holm` adjusts all
feature-component tests together. `$omnibus$p_holm` separately adjusts the
omnibus family across features. These are distinct hypothesis families.

```{r group-map, fig.alt="Estimated group loading norms and known simulated population norms across features."}
plot(group$effect_norm, type = "l", xlab = "Shared feature",
     ylab = "Loading-vector norm", ylim = range(c(group$effect_norm,
                                               sqrt(rowSums(population^2)))))
points(sqrt(rowSums(population^2)), pch = 1, col = "steelblue")
legend("topright", c("Estimated group", "Simulation population"),
       lty = c(1, NA), pch = c(NA, 1), col = c("black", "steelblue"), bty = "n")
```

```{r checks, include=FALSE}
stopifnot(all(is.finite(group$estimate)), all(group$omnibus$df2 == 10),
          max(abs(group$estimate - population)) < 0.35)
Q <- matrix(c(cos(0.4), sin(0.4), -sin(0.4), cos(0.4)), 2)
rotated <- reference
rotated$matrix <- reference$matrix %*% Q
rotated_group <- pattern_group(subjects, rotated)
stopifnot(isTRUE(all.equal(group$effect_norm, rotated_group$effect_norm)),
          isTRUE(all.equal(group$omnibus, rotated_group$omnibus)))
```

Norms are descriptive and nonnegative; sampling noise biases small norms
upward. They are not thresholded significance maps. Their magnitude also depends
on the reference score scale. Only orthogonal changes preserve Euclidean norms.

## Inspect heterogeneity and subject expression

```{r heterogeneity}
head(group$heterogeneity)
group$prediction$summary
```

`$between_covariance` is a moment estimate: sample coefficient covariance minus
average within-subject covariance, projected onto positive-semidefinite
matrices. `$heterogeneity$trace` summarizes its magnitude without choosing
component axes. Truncation at zero introduces boundary bias; this is not REML.
The Q test concerns a common-effect null, treats within-subject covariance as
known, and is approximate when those covariances are estimated. Its p values
are unadjusted descriptive diagnostics, separate from the confirmation families.

`$subject_expression` projects each subject onto the normalized mean of the
**other** subjects at each feature. It describes agreement and sign relative to
that shared pattern. It is neither an out-of-sample group decoder nor an
independent validation of the group map. `$prediction` separately retains the
original frozen decoder's held-out metrics for each subject and their
unweighted means. An unavailable subject metric makes the group mean of that
metric unavailable; missing subjects are not silently dropped.

`effects = "fixed"` instead uses full inverse-covariance GLS weighting. Its
normal and chi-squared tests treat estimated subject covariance as known and
are asymptotic. This estimates a common effect among these subjects and does
not support population generalization in the presence of heterogeneity.

## Make spatial correspondence explicit

By default all subjects must have the same feature-ID set; their column order
may differ. If identifiers differ after an independently specified spatial
correspondence, supply a mapping for each subject. Each mapping names the shared
feature and gives that subject's source feature ID.

```{r mapping, eval=FALSE}
mapping <- list(
  subject1 = c(left_motor = "native_roi_17", right_motor = "native_roi_22"),
  subject2 = c(left_motor = "native_roi_08", right_motor = "native_roi_31")
  # ...one entry for every subject, with the same shared feature names
)
group <- pattern_group(subjects, reference, spatial_mapping = mapping)
```

Mappings are one-to-one correspondences, possibly selecting a common subset.
They do not resample images or average voxels. Interpolation or parcel averaging
requires cross-feature covariance to propagate uncertainty, which confirmation
objects do not store. Perform that spatial transformation before confirmation,
then fit and estimate the transformed measurements in their stated units.

Matching preprocessing identifiers and nuisance column names are checked, but
the caller must verify that preprocessing recipes, measurement units, target
units, and nuisance meanings agree. Target sampling and omitted target
effects must also permit the same conditional estimand across subjects.
Unequal target subspaces are rejected:
rotating them to look similar would silently change the group estimand.

```{r save, eval=FALSE}
saveRDS(group, "group-pattern-confirmation.rds")
```

The saved result includes aligned subject coefficients, full mean and
between-subject covariance, basis transformations, mappings, and provenance.
It transports subject covariance one feature at a time, avoiding a second
copy of every subject covariance array. The retained aligned estimates
require features \(\times\) rank \(\times\) subjects storage; mean and
heterogeneity covariance require features \(\times\) rank\(^2\) storage. No
feature-by-feature covariance is formed. The reference basis ID is
recomputed from the actual raw target coordinates; `source_basis_id` keeps
the discovery basis identifier.

Joint hierarchical fitting of discovery patterns, and sequential
supported-rank inference, remain separate extensions. They are not required
for the workflow in this article.
