---
title: "Pattern models: whole-brain fit, regional access"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: |
  %\VignetteIndexEntry{Pattern models: whole-brain fit, regional access}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                     fig.width = 6, fig.height = 3.5)
library(rMVPA)
set.seed(904)
```

A pattern model learns a low-rank relationship between task targets and brain
measurements across the whole feature domain. From the same fit you can predict
held-out targets, inspect where model-implied signal varies, and ask how much
of the prediction remains accessible within a region.

The distinction matters when a region helps cancel correlated noise. Its
forward loading can be zero even though observing it improves prediction.
Forward patterns and decoding weights describe different aspects of the fit.

## Fit and evaluate a classification model

This small example creates two signal territories and a third territory that
shares noise with the first. Rows represent observations; columns represent
voxels. Class labels are balanced within each of three runs.

```{r data}
n <- 120
p <- 48
class <- factor(rep(c("a", "b", "c"), length.out = n))
run <- rep(1:3, each = 40)
code <- model.matrix(~ class - 1)
X <- matrix(rnorm(n * p), n, p)
shared_noise <- rnorm(n, sd = 2)
X[, 1:12] <- X[, 1:12] + 1.2 * (code[, 1] - code[, 2]) + shared_noise
X[, 13:24] <- X[, 13:24] + 1.2 * (code[, 2] - code[, 3])
X[, 25:36] <- X[, 25:36] + shared_noise
space <- neuroim2::NeuroSpace(c(4, 4, 3), c(1, 1, 1))
mask <- neuroim2::NeuroVol(array(1, c(4, 4, 3)), space)
images <- neuroim2::NeuroVec(array(t(X), c(4, 4, 3, n)),
                           neuroim2::NeuroSpace(c(4, 4, 3, n), c(1, 1, 1)))
dataset <- mvpa_dataset(images, mask = mask)
design <- mvpa_design(data.frame(class, run), y_train = ~ class,
                      block_var = ~ run)
spec <- pattern_model(dataset, design, rank = 2,
                      noise = list(type = "diag_lowrank", rank = 1))
result <- run_global(spec, return_fits = TRUE, refit = TRUE)
performance(result)
```

The performance table scores held-out runs. The refit uses every training row
and supplies descriptive maps; it is not the fit used to report validation
accuracy. Here the rank is fixed at two. With `rank = "auto"`, rank is selected
inside each outer training split. A spatial penalty can be added with
`penalty = list(sparse = "auto", signed_smooth = 1)`; it acts on forward
patterns, and its strength is selected using held-out decoding loss.

`return_fits = TRUE` retains the fold fits needed for locality and diagnostics.
Retaining these objects costs memory, especially with many resamples.

## Read invariant maps

```{r maps, fig.alt="Model-implied signal standard deviation and conditional information across input voxels."}
signal_sd <- model_importance(result$refit, type = "signal_sd")
information <- model_importance(result$refit, type = "conditional_info")
oldpar <- par(mfrow = c(1, 2), mar = c(4, 4, 1, 1))
plot(signal_sd, type = "h", xlab = "Input voxel", ylab = "Signal SD")
plot(information, type = "h", xlab = "Input voxel", ylab = "Conditional information (nats)")
par(oldpar)
```

Signal SD is the standard deviation implied by the fitted signal covariance,
`sqrt(diag(A Phi A'))`, expressed in the original measurement units. It includes
all fitted task directions. Conditional information asks what a voxel adds
once the other voxels have been observed. It can be positive in a noise
canceller with zero signal SD. Finite-sample estimated loadings are rarely
exactly zero without sparsity, so the simulated nuisance territory need not
have an exactly zero estimated signal map.

Conditional information is a **working Gaussian score-model** quantity, in
nats. For classification it is not empirical mutual information about discrete
class labels. Both maps are descriptive, not significance maps.

Using `model_importance(result)` returns a spatial image through the dataset's
map builder. Using `result$refit` returns a vector aligned to input columns;
screened columns are `NA`. Forward component patterns and calibrated decoder
weights are available from `model_patterns(fit, "forward")` and
`model_patterns(fit, "weights")`. Their individual columns depend on the chosen
component coordinates. Multibasis datasets require separate channel maps;
automatic channel aggregation would change the quantity being shown.

## Ask what each region can decode

```{r local}
regions <- list(signal_1 = 1:12, signal_2 = 13:24,
                shared_noise = 25:36, background = 37:48,
                whole_domain = 1:48)
local <- local_performance(result, regions)
local[local$metric == "Accuracy", ]
```

In this realization, the first signal territory is near chance on its own
despite nonzero loadings. The whole-domain prediction benefits from observing
features beyond that territory.

Each entry in `regions` specifies **input matrix column positions**, not voxel
IDs. For every fold, the query keeps the whole-brain forward pattern within
the region and restricts the residual covariance to `Psi_RR`. It then
recomputes the decoder and predicts the same held-out observations used in the
whole-brain ledger. Cropping whole-brain decoding weights would retain noise
cancellation terms that require unobserved features and gives a different,
generally incorrect answer.

The regional score measures access under the shared whole-brain representation.
It does not estimate the best accuracy any separately trained local model
could reach. A finite-sample region can outperform the whole-brain prediction.
A noise-only territory may contribute conditionally while decoding poorly
alone. For independent ROI comparisons, `independent_roi` accepts a named list
of fold-resolved ledgers with matching rows, folds, truth, and baselines.
Repeated assessments are pooled to one prediction per observation before any
of these methods is scored.

```{r locality-check, include=FALSE}
all_region <- local[local$region == "whole_domain", ]
stopifnot(isTRUE(all.equal(all_region$whole_brain, all_region$local_restricted)))
```

## Rotate a display and compare folds

```{r rotation}
view <- rotate_patterns(result$refit)
view$reconstruction_error
result$component_stability[, c("fold1", "fold2", "rank1", "rank2", "n_common", "overlap")]
```

Rotation changes display coordinates, preserving `L_b H L_t' = A C'`.
The view retains the original fit, so predictions and invariant maps remain
exactly unchanged. Its matrices use fitted feature units and whitened target
coordinates; the view also stores their back-transforms. Only orthogonal
rotations are supported.

Stability compares spatial column spaces in original feature units using
principal angles on features retained in both folds. An overlap of one means
the two nonempty subspaces coincide. Different ranks reduce overlap even if
the smaller space lies within the larger. This is a subspace diagnostic, not
a test that individual components or a particular rank are supported.

```{r haufe}
vapply(result$haufe_diagnostics, function(x) x$relative_discrepancy, numeric(1))
```

The Haufe diagnostic compares empirical held-out covariance patterns with the
model-implied patterns using calibrated scores. Large discrepancies can flag
sampling noise or model misspecification. They are not p-values. Fold-level
row identities are checked against training IDs. For an additional independent
sample, use `pattern_haufe(fit, X_holdout, observation_ids)` and supply identifiers
in the fit's training-ID namespace. Automatic results use `train:<row ID>` for
training/CV rows and `test:<row ID>` for an external test partition; the user
must ensure an external partition contains genuinely independent observations.

## Decode continuous feature targets

The same model accepts multiple continuous targets. The example below uses
two continuous measurements generated with a low-rank encoding relationship.

```{r continuous}
features <- matrix(rnorm(n * 2), n, 2,
                   dimnames = list(NULL, c("feature_1", "feature_2")))
brain <- matrix(rnorm(n * p), n, p)
brain[, 1:12] <- brain[, 1:12] + features[, 1]
brain[, 13:24] <- brain[, 13:24] + features[, 2]
continuous_data <- mvpa_dataset(
  neuroim2::NeuroVec(array(t(brain), c(4, 4, 3, n)),
                    neuroim2::NeuroSpace(c(4, 4, 3, n), c(1, 1, 1))), mask = mask)
continuous_design <- mvpa_design(data.frame(id = 1:n, run),
                                 cv_labels = 1:n, targets = features,
                                 block_var = ~ run)
continuous <- run_global(pattern_model(continuous_data, continuous_design, rank = 2),
                         return_fits = TRUE)
performance(continuous)
local_performance(continuous, list(first = 1:12, second = 13:24))
```

Predictive R-squared uses each fold's training-target mean as its baseline;
it is not squared correlation and can be negative. The reported aggregate
averages per-response R-squared. Targets are centred and whitened using
training rows only. Predictions are returned in the original target units.

## Choose the model for the question

| Question | Model family |
|---|---|
| Whole-domain forward patterns and regional access under one covariance-aware fit | `pattern_model` |
| Decode feature vectors within independently fitted regions | `feature_rsa_model` |
| Voxelwise encoding from grouped predictors | `banded_ridge_model` |
| Map between perception and retrieval domains | `remap_rrr_model` |
| Find locally predictive neighborhoods using separate local fits | Searchlight analysis |

The current noise model combines diagonal variance and a few shared residual
components. It may miss richer local correlations. Signed Laplacian smoothing
encourages neighboring loadings to agree; it does not provide a coherent
support envelope for alternating-sign fine-scale codes. Confirmation inference,
rank tests, group estimates, and that support-envelope extension are separate
work.

To save retained fits, ledgers, and refit maps, call
`save_results(result, "pattern-output")`. The RDS preserves the result object;
spatial refit maps are also written as images for supported datasets.
