---
title: Advanced Regional MVPA Analysis in rMVPA
author: Bradley Buchsbaum
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: |
  %\VignetteIndexEntry{Advanced Regional MVPA Analysis in rMVPA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-teal');});</script>

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(albersdown::theme_albers(params$family))
}
library(neuroim2)
library(rMVPA)
```

# Introduction

Use regional MVPA when you already have an ROI mask and want one prediction
summary per region. The input is a 4D neuroimaging dataset plus a design table;
the output is a `regional_mvpa_result` containing cross-validated performance,
trial-level predictions, and map-ready regional summaries.

This vignette follows one compact workflow: make a small dataset, define three
ROIs, fit a blocked cross-validated classifier, and inspect the region-level
performance.

# Data Generation and Preparation

We begin by generating a synthetic volumetric dataset using the `gen_sample_dataset()` function. This function creates a 4D array (with spatial dimensions and multiple observations), along with a binary mask and an associated design for cross-validation.

```{r generate_data}
data_out <- rMVPA::gen_sample_dataset(D = c(6,6,6), nobs = 80, blocks = 4, nlevels = 2)

data.frame(
  observations = length(data_out$design$block_var),
  active_voxels = sum(data_out$dataset$mask),
  blocks = length(unique(data_out$design$block_var)),
  classes = length(levels(data_out$design$targets))
)
```

The returned list contains two objects:

- **dataset**: an MVPA dataset object with training data and a binary mask.
- **design**: an MVPA design object specifying the response variable and block structure.

## Creating a Region Mask

For regional analysis, we need to define ROIs. Here, we create a region mask by randomly assigning each active voxel in the binary mask a region label (from 1 to 3). This simulates a scenario where the brain is partitioned into three regions of interest.

```{r create_region_mask}
mask <- data_out$dataset$mask
nvox <- sum(mask)

set.seed(123)
region_mask <- neuroim2::NeuroVol(
  sample(1:3, size = nvox, replace = TRUE),
  neuroim2::space(mask),
  indices = which(mask > 0)
)
table(region_mask)
```

# Setting Up the MVPA Model

Next, we create an MVPA model to evaluate a classification task. In brief, we construct an `mvpa_dataset`, specify the design (including the block variable and response), and define the model with `mvpa_model()` using a chosen classifier and cross‑validation strategy.

```{r setup_mvpa_model}
dset <- mvpa_dataset(data_out$dataset$train_data, mask = data_out$dataset$mask)
cval <- blocked_cross_validation(data_out$design$block_var)

mod <- load_model("sda")
tune_grid <- data.frame(lambda = 0.01, diagonal = FALSE)

mvpa_mod <- mvpa_model(
  mod,
  dataset = dset,
  design = data_out$design,
  crossval = cval,
  tune_grid = tune_grid
)
mvpa_mod
```

The model object packages the dataset, design, classifier, tuning grid, and
cross-validation plan.

# Running the Regional Analysis

The regional analysis is executed with `run_regional()`. Internally it prepares ROI indices from the region mask, applies the MVPA model to each ROI, and then compiles performance metrics and prediction tables.

```{r run_regional_analysis, results='hide'}
regional_results <- run_regional(mvpa_mod, region_mask)
```

The output is a `regional_mvpa_result` object with a `performance_table` (cross‑validated metrics per region), a `prediction_table` (trial‑level predictions), and `vol_results` (volumetric maps of performance across the brain).

# Examining the Results

We can inspect the performance table to evaluate model accuracy in each region.

```{r performance_results}
regional_results$performance_table
```

```{r regional-checks, include = FALSE}
stopifnot(nrow(regional_results$performance_table) == length(unique(region_mask[region_mask > 0])))
metric_names <- intersect(c("Accuracy", "AUC"), names(regional_results$performance_table))
stopifnot(length(metric_names) > 0)
stopifnot(all(is.finite(as.matrix(regional_results$performance_table[metric_names]))))
```

```{r regional-perf-bars, echo = FALSE, fig.cap = "Cross-validated accuracy per region. The dashed line marks chance for a 2-class problem.", out.width = "70%", fig.width = 5.6, fig.height = 3.4}
perf <- regional_results$performance_table
metric <- intersect(c("Accuracy", "AUC"), names(perf))[1]
if (!is.na(metric) && nrow(perf) > 0) {
  vals <- as.numeric(perf[[metric]])
  ids  <- as.character(perf$roinum)
  op <- par(mar = c(3.5, 4, 1.5, 0.5))
  bp <- barplot(vals, names.arg = ids, ylim = c(0, 1),
                col = "#3a5b97", border = NA,
                xlab = "ROI", ylab = metric, main = "")
  abline(h = 0.5, lty = 2, col = "grey50")
  par(op)
}
```

The bar plot makes between-region differences visible at a glance and shows where each region sits relative to chance. For a compact trial-level check, show the first few predictions:

```{r prediction_results}
utils::head(regional_results$prediction_table)
```

Volumetric results (`vol_results`) can be further visualized with neuroimaging tools to determine spatial patterns of performance.

# Summary 

This vignette showed you how to generate synthetic neuroimaging data, define ROIs with region masks, set up MVPA models with cross-validation, run analyses across ROIs, and interpret the performance metrics. You now have the tools to conduct regional MVPA analyses on your own neuroimaging data.

For further details, see `vignette("Searchlight_Analysis")` for the searchlight counterpart and `vignette("CrossValidation")` for cross-validation strategies.
