---
title: "Advanced Regional MVPA Analysis in rMVPA"
author: "Bradley Buchsbaum"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Advanced Regional MVPA Analysis in rMVPA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

Regional multivariate pattern analysis (MVPA) allows you to evaluate prediction performance within predefined brain regions. In this vignette, we demonstrate a complete workflow for performing regional MVPA analysis using the `rMVPA` package. We will:

- Generate a synthetic neuroimaging dataset.
- Define regions of interest (ROIs) using a region mask.
- Construct appropriate MVPA dataset and design objects.
- Build an MVPA model with cross-validation.
- Run the regional analysis over the ROIs.
- Examine performance metrics and prediction outputs.

This approach, detailed in the source files (`regional.R`, `mvpa_model.R`, and `dataset.R`), ensures robust implementation of regional analysis within rMVPA.

# Data Generation and Preparation

We begin by generating a synthetic volumetric dataset using the `gen_sample_dataset()` function. This function creates a 4D array (with spatial dimensions and multiple observations), along with a binary mask and an associated design for cross-validation.

```{r, message=FALSE, warning=FALSE}
library(rMVPA)
library(neuroim2)
# Generate a synthetic dataset with dimensions 6x6x6, 80 observations, divided into 4 blocks
data_out <- rMVPA::gen_sample_dataset(D = c(6,6,6), nobs = 80, blocks = 4, nlevels = 2)
print(data_out)
```

The returned list contains:

- **dataset**: an MVPA dataset object with training data and a binary mask.
- **design**: an MVPA design object specifying the response variable and block structure.

## Creating a Region Mask

For regional analysis, we need to define ROIs. Here, we create a region mask by randomly assigning each active voxel in the binary mask a region label (from 1 to 3). This simulates a scenario where the brain is partitioned into three regions of interest.

```{r, message=FALSE, warning=FALSE}
# Extract the binary mask from the dataset
mask <- data_out$dataset$mask
nvox <- sum(mask)

# Create a regional mask: assign each voxel a random region number (1 to 3)
set.seed(123)  # for reproducibility
region_mask <- neuroim2::NeuroVol(sample(1:3, size = nvox, replace = TRUE), neuroim2::space(mask), indices = which(mask > 0))
table(region_mask)
```

# Setting Up the MVPA Model

Next, we create an MVPA model to evaluate a classification task. This involves:

1. Constructing an MVPA dataset using `mvpa_dataset()`.
2. Specifying the design (including the block variable and response) via `mvpa_design()`.
3. Defining the model with `mvpa_model()` using a chosen classifier and cross-validation strategy.

```{r, message=FALSE, warning=FALSE}
# Create MVPA dataset object from the generated training data and mask
 dset <- mvpa_dataset(data_out$dataset$train_data, mask = data_out$dataset$mask)

# Build cross-validation structure using block information from the design
cval <- blocked_cross_validation(data_out$design$block_var)

# Load a classification model; here we use "sda" (Shrinkage Discriminant Analysis)
mod <- load_model("sda")
tune_grid <- data.frame(lambda = 0.01, diagonal = FALSE)

# Create the MVPA model object
mvpa_mod <- mvpa_model(mod, dataset = dset, design = data_out$design, crossval = cval, tune_grid = tune_grid)
print(mvpa_mod)
```

The `mvpa_model()` function, as defined in `mvpa_model.R`, packages all necessary parameters including cross-validation and performance computation.

# Running the Regional Analysis

The regional analysis is executed with the `run_regional()` function, which:

- Prepares the ROIs by extracting voxel indices from the region mask.
- Iterates over the regions, applying the MVPA model to each ROI.
- Compiles performance metrics and prediction tables.

```{r, message=FALSE, warning=FALSE, results='hide'}
# Run the regional analysis on the defined region mask
regional_results <- run_regional(mvpa_mod, region_mask)
```

The output is a `regional_mvpa_result` object containing:

- **performance_table**: Cross-validated performance metrics per region.
- **prediction_table**: Detailed predictions for each ROI and trial.
- **vol_results**: Volumetric maps representing performance distributed across the brain.

# Examining the Results

We can inspect the performance table to evaluate model accuracy in each region.

```{r}
# Display performance metrics for each region
print(regional_results$performance_table)
```

For a more detailed view, the prediction table shows trial-by-trial predictions:

```{r}
# Display the prediction table
print(regional_results$prediction_table)
```

Volumetric results (`vol_results`) can be further visualized with neuroimaging tools to determine spatial patterns of performance.

# Under the Hood: How It Works

The `run_regional()` function internally calls `prep_regional()` (from `regional.R`) to process the region mask, and then uses `mvpa_iterate()` to apply the MVPA model across each ROI. Functions such as `combine_regional_results()` and `combine_prediction_tables()` merge the individual regional outputs into a comprehensive result.

This modular design, laid out in `mvpa_model.R` and `dataset.R`, ensures that:

- Data integrity is maintained.
- Cross-validation is properly applied.
- Results are aggregated for clear interpretation at the regional level.

# Summary 

In this vignette, we have demonstrated:

- **Data Generation**: Creating a synthetic neuroimaging dataset using `gen_sample_dataset()`.
- **ROI Definition**: Building a region mask for defining brain areas.
- **Model Setup**: Constructing an MVPA model with proper design and cross-validation.
- **Regional Analysis**: Running the model across each ROI and aggregating results.
- **Results Interpretation**: Examining performance and prediction metrics.

For further details, please refer to the source files:

- `regional.R` for regional analysis methods.
- `mvpa_model.R` for MVPA model creation and result formatting.
- `dataset.R` for dataset generation routines.

Happy analyzing!
