---
title: "Advanced Regional MVPA Analysis in rMVPA"
author: "Bradley Buchsbaum"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Advanced Regional MVPA Analysis in rMVPA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

Regional MVPA evaluates prediction performance within predefined brain regions. This vignette walks through the complete workflow using rMVPA:

We'll generate synthetic data, define ROIs with a region mask, build the MVPA model with cross-validation, run the analysis, and examine the results. The implementation follows the approach in `regional.R`, `mvpa_model.R`, and `dataset.R`.

# Data Generation and Preparation

We begin by generating a synthetic volumetric dataset using the `gen_sample_dataset()` function. This function creates a 4D array (with spatial dimensions and multiple observations), along with a binary mask and an associated design for cross-validation.

```{r, message=FALSE, warning=FALSE}
library(rMVPA)
library(neuroim2)
# Generate a synthetic dataset with dimensions 6x6x6, 80 observations, divided into 4 blocks
data_out <- rMVPA::gen_sample_dataset(D = c(6,6,6), nobs = 80, blocks = 4, nlevels = 2)
print(data_out)
```

The returned list contains:

- **dataset**: an MVPA dataset object with training data and a binary mask.
- **design**: an MVPA design object specifying the response variable and block structure.

## Creating a Region Mask

For regional analysis, we need to define ROIs. Here, we create a region mask by randomly assigning each active voxel in the binary mask a region label (from 1 to 3). This simulates a scenario where the brain is partitioned into three regions of interest.

```{r, message=FALSE, warning=FALSE}
# Extract the binary mask from the dataset
mask <- data_out$dataset$mask
nvox <- sum(mask)

# Create a regional mask: assign each voxel a random region number (1 to 3)
set.seed(123)  # for reproducibility
region_mask <- neuroim2::NeuroVol(sample(1:3, size = nvox, replace = TRUE), neuroim2::space(mask), indices = which(mask > 0))
table(region_mask)
```

# Setting Up the MVPA Model

Next, we create an MVPA model to evaluate a classification task. In brief, we construct an `mvpa_dataset`, specify the design (including the block variable and response), and define the model with `mvpa_model()` using a chosen classifier and cross‑validation strategy.

```{r, message=FALSE, warning=FALSE}
# Create MVPA dataset object from the generated training data and mask
 dset <- mvpa_dataset(data_out$dataset$train_data, mask = data_out$dataset$mask)

# Build cross-validation structure using block information from the design
cval <- blocked_cross_validation(data_out$design$block_var)

# Load a classification model; here we use "sda" (Shrinkage Discriminant Analysis)
mod <- load_model("sda")
tune_grid <- data.frame(lambda = 0.01, diagonal = FALSE)

# Create the MVPA model object
mvpa_mod <- mvpa_model(mod, dataset = dset, design = data_out$design, crossval = cval, tune_grid = tune_grid)
print(mvpa_mod)
```

The `mvpa_model()` function, as defined in `mvpa_model.R`, packages all necessary parameters including cross-validation and performance computation.

# Running the Regional Analysis

The regional analysis is executed with `run_regional()`. Internally it prepares ROI indices from the region mask, applies the MVPA model to each ROI, and then compiles performance metrics and prediction tables.

```{r, message=FALSE, warning=FALSE, results='hide'}
# Run the regional analysis on the defined region mask
regional_results <- run_regional(mvpa_mod, region_mask)
```

The output is a `regional_mvpa_result` object with a `performance_table` (cross‑validated metrics per region), a `prediction_table` (trial‑level predictions), and `vol_results` (volumetric maps of performance across the brain).

# Examining the Results

We can inspect the performance table to evaluate model accuracy in each region.

```{r}
# Display performance metrics for each region
print(regional_results$performance_table)
```

For a more detailed view, the prediction table shows trial-by-trial predictions:

```{r}
# Display the prediction table
print(regional_results$prediction_table)
```

Volumetric results (`vol_results`) can be further visualized with neuroimaging tools to determine spatial patterns of performance.

# Under the Hood: How It Works

The `run_regional()` function internally calls `prep_regional()` (from `regional.R`) to process the region mask, and then uses `mvpa_iterate()` to apply the MVPA model across each ROI. Functions such as `combine_regional_results()` and `combine_prediction_tables()` merge the individual regional outputs into a comprehensive result.

This modular design, laid out in `mvpa_model.R` and `dataset.R`, ensures that:

- Data integrity is maintained.
- Cross-validation is properly applied.
- Results are aggregated for clear interpretation at the regional level.

# Summary 

This vignette showed you how to generate synthetic neuroimaging data, define ROIs with region masks, set up MVPA models with cross-validation, run analyses across ROIs, and interpret the performance metrics. You now have the tools to conduct regional MVPA analyses on your own neuroimaging data.

For further details, please refer to the source files:

- `regional.R` for regional analysis methods.
- `mvpa_model.R` for MVPA model creation and result formatting.
- `dataset.R` for dataset generation routines.

Happy analyzing!
