---
title: "Regional Analysis"
author: "Bradley Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Searchlight Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## MVPA Analysis on Regions of Interest

```{r, echo=FALSE, message=FALSE}
library(neuroim)
library(devtools)
load_all()
```

If we have one or more pre-defined regions of interests (ROIs) we can use `rMVPA` to run a clasisifation/regression model in each region, yielding peformance measures for each region.

### Generate a volumetric dataset with 100 observations and two classes

To generate a dataset we use the `gen_sample_dataset` function. We are creating a 4-dimensional neuroimaging dataset, with 6-by-6-by-6 spatial dimensions and 80 observations in the 4th dimension. These 80 observations are divided into 4 blocks, each consisting of 20 trials. The generated `y` variable is a `factor` with 2 levels ('a' and 'b'). the `gen_sample_dataset` function creates a list with two elements: an `mvpa_dataset` object (`dataset`) and an `mvpa_design` object (`design`). The first contains information about the data itself and the second contains information about the experimental design.



```{r}

dset <- gen_sample_dataset(D=c(6,6,6), nobs = 80, blocks=4, nlevels=2)
print(dset)
```

```

Now we generate an arbitrary set of ROIs by assigning each voxel in the input mask a value ranging from 1 to 3. 

```{r}

mask <- dset$dataset$mask
nvox <- sum(mask)

region_mask <- neuroim::BrainVolume(sample(1:3,replace=TRUE,size=nvox), space(mask), indices=which(mask>0))
table(region_mask)

```

The `region_mask` volume now defines a set of three ROIs where each ROI is defined by an integer value. Next, we can run a classification/regression model within each ROI using the `run_regional` function:

```{r, message=FALSE, results='hide'}


mod <- load_model("sda")
tune_grid <- data.frame(lambda=.01, diagonal=FALSE)
cval <- blocked_cross_validation(dset$design$block_var)
mspec <- mvpa_model(mod, dataset=dset$dataset, design=dset$design, tune_grid=tune_grid,crossval=cval)
res <- run_regional(mspec,region_mask)

```

`run_regional` produces seeveral output values, including `performance_table`, which includes cross-validated performance metrics for each ROI:

```{r}
print(res$performance_table)

```

To get fine-grained information, the individual predictions for each ROI and trial are included in `prediction_table`:

```{r}
print(res$prediction_table)
```

