---
title: "Decode a held-out fMRI run with rMVPA"
author: Bradley Buchsbaum
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Decode a held-out fMRI run with rMVPA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-teal');});</script>
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 4.6,
  fig.align = "center"
)

library(rMVPA)
library(neuroim2)

if (!requireNamespace("sda", quietly = TRUE)) {
  stop("This vignette requires the suggested package 'sda'.", call. = FALSE)
}

set.seed(20260813)
set_log_level("WARN")

resolve_haxby_path <- function() {
  installed <- system.file(
    "extdata", "haxby2001_subj1", "patterns.rds",
    package = "rMVPA"
  )
  if (nzchar(installed) && file.exists(installed)) {
    return(installed)
  }

  candidates <- c(
    file.path("..", "inst", "extdata", "haxby2001_subj1", "patterns.rds"),
    file.path("inst", "extdata", "haxby2001_subj1", "patterns.rds")
  )
  found <- candidates[file.exists(candidates)]
  if (!length(found)) {
    stop("The bundled Haxby derivative could not be found.", call. = FALSE)
  }
  normalizePath(found[[1]])
}
```

Can a distributed fMRI pattern identify what a person is viewing **in a run
the model has never seen**? This guide answers that question with Subject 1
from Haxby et al. (2001): 96 ventral-temporal patterns, eight visual
categories, and 12 scanning runs. We will hold out one entire run at a time,
fit a classifier on the other 11, and collect 96 genuinely out-of-sample
predictions.

By the end, you will have a checked analysis specification, a cross-validated
accuracy estimate, and a confusion matrix that shows where the classifier
succeeds and fails. The example is small enough to run during package checks,
but the objects and analysis engine are the same ones used for atlas regions
and whole-brain searchlights.

> **Scientific scope.** This is a within-subject demonstration of
> generalization across runs. It is not evidence about new participants, and
> its accuracy is not a population-level inferential test.

## What data are we decoding?

The package ships a compact derivative of the public Haxby dataset. Each row
is the mean activation pattern for one category in one run; each column is a
voxel in the published ventral-temporal (VT) mask. The raw BOLD time series
have already been reduced to analysis-ready patterns, so no download is needed.
This is a reanalysis of that derivative, not a literal reproduction of Haxby
et al.'s original classifier and validation procedure.
The derivative was built from the
[PyMVPA Subject 1 tutorial archive](http://data.pymvpa.org/datasets/haxby2001/subj1-2010.01.14.tar.gz);
`data-raw/haxby2001_subj1.R` is the complete rebuild script, and
`inst/extdata/haxby2001_subj1/README.md` records the contents and
redistribution note. The SHA-256 digest of the analyzed `patterns.rds` file is
`0d2c8cd36ec955a31e201b10efeae6c4a478da6dcc1f5bf05af0c9ef22c21706`.

```{r load-data}
bundle <- readRDS(resolve_haxby_path())

data_overview <- data.frame(
  patterns = nrow(bundle$patterns),
  voxels = ncol(bundle$patterns),
  categories = length(unique(bundle$category)),
  runs = length(unique(bundle$run))
)
knitr::kable(data_overview, caption = "The bundled Haxby derivative.")
```

The design is balanced: every run contributes one pattern for each of the
eight categories. That matters because the nominal accuracy expected from
uniform guessing is `1 / 8 = 12.5%`.

```{r check-balance, include = FALSE}
category_by_run <- table(bundle$run, bundle$category)
stopifnot(
  identical(dim(bundle$patterns), c(96L, 577L)),
  identical(dim(category_by_run), c(12L, 8L)),
  all(category_by_run == 1L),
  all(is.finite(bundle$patterns)),
  any(apply(bundle$patterns, 2L, stats::sd) > 0)
)
```

In your own study, these rows would usually be trial-wise or condition-wise
beta estimates from a first-level model. See
`vignette("Constructing_Datasets")` for file-backed images, masks, and other
input layouts.

## What must remain independent?

Measurements from the same fMRI run share drift, motion, and temporal noise.
A random row-wise split can therefore put correlated observations on both
sides of the train/test boundary and exaggerate performance. Our estimand is
instead explicit:

> How accurately can a model trained on 11 runs classify the eight category
> patterns in the twelfth run?

We encode the category as the response, the run as the blocking variable, and
use leave-one-run-out cross-validation.

```{r design-and-cv}
design_table <- data.frame(
  category = factor(bundle$category),
  run = bundle$run
)

design <- mvpa_design(
  design_table,
  y_train = ~ category,
  block_var = ~ run
)
crossval <- blocked_cross_validation(bundle$run)
```

```{r inspect-folds, include = FALSE}
folds <- crossval_samples(
  crossval,
  data = data.frame(row = seq_len(nrow(design_table))),
  y = design_table$category
)

fold_integrity <- vapply(seq_len(nrow(folds)), function(i) {
  train_id <- folds$train[[i]]$idx
  test_id <- folds$test[[i]]$idx
  length(intersect(bundle$run[train_id], bundle$run[test_id])) == 0L &&
    length(test_id) == 8L &&
    length(unique(design_table$category[test_id])) == 8L
}, logical(1))

stopifnot(nrow(folds) == 12L, all(fold_integrity))
```

There are 12 folds. In each fold, 88 patterns train the model and the eight
patterns from one untouched run test it. Every category appears once in every
test fold.

## How does rMVPA represent the analysis?

An rMVPA workflow has four scientific objects. Keeping them separate makes the
analysis auditable: the imaging measurements live in a dataset, the response
and grouping variables live in a design, the estimator and resampling rule
live in a model specification, and an engine decides *where* to fit it.

| Question | Object in this analysis |
|:--|:--|
| What are the measurements? | An `mvpa_dataset` with 96 VT patterns |
| What is predicted, and what defines independence? | An `mvpa_design` with category and run |
| What is fitted and resampled? | An `mvpa_model` with blocked CV |
| Where is it fitted? | One labelled VT region passed to `run_regional()` |

The bundle stores a matrix plus spatial metadata. The next chunk back-projects
the voxel columns into a 4-D sparse neuroimaging object. With your own data,
`neuroim2::read_vec()` and `neuroim2::read_vol()` usually provide these
objects directly.

```{r reconstruct-images}
mask_array <- array(0L, bundle$mask_dim)
mask_array[bundle$mask_idx] <- 1L
vt_mask <- LogicalNeuroVol(mask_array, bundle$mask_space)

voxel_by_pattern <- t(bundle$patterns)
storage.mode(voxel_by_pattern) <- "double"
vector_space <- neuroim2::add_dim(bundle$mask_space, ncol(voxel_by_pattern))
bold_patterns <- SparseNeuroVec(
  voxel_by_pattern,
  space = vector_space,
  mask = as.logical(mask_array)
)
vt_region <- NeuroVol(mask_array, bundle$mask_space)
```

```{r check-images, include = FALSE}
stopifnot(
  ncol(voxel_by_pattern) == nrow(design_table),
  sum(vt_mask) == ncol(bundle$patterns),
  all(mask_array %in% c(0L, 1L))
)
```

```{r build-model}
dataset <- mvpa_dataset(bold_patterns, mask = vt_mask)

model <- mvpa_model(
  model = load_model("sda_notune"),
  dataset = dataset,
  design = design,
  crossval = crossval,
  return_predictions = TRUE
)
```

`sda_notune` is shrinkage discriminant analysis with no hyperparameter search.
It is a strong default when the number of correlated voxels is large relative
to the number of observations. The estimator is refitted from scratch inside
each training fold; the held-out run is used only for prediction.
This model requires the suggested CRAN package `sda`; install that optional
dependency before running the vignette if it is not already available.

## Can we catch a bad design before fitting?

`validate_analysis()` checks the specification for common failures such as a
cross-validation rule that ignores run structure, missing classes in a fold,
or test folds that are too small.

```{r preflight}
preflight <- validate_analysis(model, verbose = FALSE)

preflight_summary <- data.frame(
  passed = preflight$n_pass,
  warnings = preflight$n_warn,
  failures = preflight$n_fail
)
knitr::kable(preflight_summary, caption = "Static checks before model fitting.")
```

```{r assert-preflight, include = FALSE}
stopifnot(
  inherits(preflight, "validation_result"),
  preflight$n_pass > 0L,
  preflight$n_warn == 0L,
  preflight$n_fail == 0L
)
```

A clean preflight is necessary, not magical. It can verify the design and fold
structure represented in these objects; it cannot prove that upstream
preprocessing was scientifically appropriate. Any data-driven scaling,
feature selection, or tuning must also be learned without looking at the test
observations. `vignette("FeatureSelection")` shows how to place feature
selection inside the resampling loop.

## What does the held-out analysis find?

The VT mask contains one non-zero region, so `run_regional()` fits the model
once per fold in that region and then pools the held-out predictions. Setting
`preflight = "error"` prevents execution if a methodological check fails.

```{r run-analysis}
result <- run_regional(
  model,
  region_mask = vt_region,
  preflight = "error",
  verbose = FALSE
)
```

`result$performance_table` contains region-level metrics;
`result$prediction_table` contains one row per held-out prediction. Accuracy is
the most transparent primary measure for this balanced eight-way problem.

```{r summarize-result}
predictions <- as.data.frame(result$prediction_table)
accuracy <- result$performance_table$Accuracy[[1]]
auc_centered <- result$performance_table$AUC[[1]]
chance <- 1 / length(unique(design_table$category))

result_summary <- data.frame(
  held_out_predictions = nrow(predictions),
  correct = sum(predictions$correct),
  accuracy = sprintf("%.1f%%", 100 * accuracy),
  chance_accuracy = sprintf("%.1f%%", 100 * chance),
  accuracy_above_chance = sprintf("%.1f points", 100 * (accuracy - chance)),
  chance_centered_AUC = sprintf("%.3f", auc_centered)
)
knitr::kable(result_summary, caption = "Performance pooled across 12 held-out runs.")
```

The model correctly classifies `r sum(predictions$correct)` of
`r nrow(predictions)` patterns: **`r sprintf("%.1f%%", 100 * accuracy)` accuracy**, compared
with `r sprintf("%.1f%%", 100 * chance)` under uniform guessing. The result
supports the narrow claim that category information in this subject's VT
patterns generalizes across runs.

rMVPA reports multiclass AUC as the mean one-versus-rest AUC transformed to
`2 * raw AUC - 1`. Its range is therefore -1 to 1, with 0 as the
chance reference and 1 as perfect ranking. It is not raw AUC and it is not
`AUC - 0.5`.

```{r prepare-diagnostics, include = FALSE}
stopifnot(
  nrow(predictions) == nrow(design_table),
  identical(sort(predictions$.rownum), seq_len(nrow(design_table))),
  all(predictions$correct %in% c(TRUE, FALSE)),
  is.finite(accuracy),
  accuracy >= 0 && accuracy <= 1,
  accuracy >= 0.75,
  is.finite(auc_centered),
  auc_centered >= -1 && auc_centered <= 1,
  auc_centered >= 0.8,
  abs(accuracy - mean(predictions$correct)) < 1e-12,
  accuracy > chance
)

predictions$run <- bundle$run[predictions$.rownum]
run_accuracy <- stats::aggregate(correct ~ run, predictions, mean)

class_levels <- sort(unique(as.character(predictions$observed)))
confusion_counts <- table(
  observed = factor(predictions$observed, levels = class_levels),
  predicted = factor(predictions$predicted, levels = class_levels)
)
confusion_rate <- prop.table(confusion_counts, margin = 1L)
class_accuracy <- diag(confusion_rate)

stopifnot(
  identical(dim(confusion_counts), c(8L, 8L)),
  sum(confusion_counts) == nrow(predictions),
  all(abs(rowSums(confusion_rate) - 1) < 1e-12),
  diff(range(run_accuracy$correct)) > 0,
  diff(range(confusion_rate)) > 0,
  min(run_accuracy$correct) >= 0.5,
  min(class_accuracy) >= 0.5,
  all(is.finite(confusion_rate))
)
```

```{r diagnostics, echo = FALSE, fig.cap = "Held-out performance by test run (left) and the row-normalized confusion matrix (right). The dashed line is eight-way chance accuracy; the solid line is pooled accuracy. Every plotted value comes from a prediction made while that observation's run was held out.", fig.alt = "Two-panel diagnostic figure. On the left, accuracy in each held-out run ranges from 0.75 to 1.0, well above the dashed chance line at 0.125, with pooled accuracy at 0.917. On the right, a row-normalized eight-category confusion matrix has a strong diagonal; bottle has the lowest category accuracy at 0.75.", fig.width = 9, fig.height = 4.5, out.width = "100%"}
old_par <- par(no.readonly = TRUE)
layout(matrix(c(1, 2), nrow = 1), widths = c(0.85, 1.15))

par(mar = c(4.2, 4.2, 2.2, 0.8))
plot(
  run_accuracy$run,
  run_accuracy$correct,
  type = "b",
  pch = 19,
  lwd = 2,
  col = "#15616D",
  ylim = c(0, 1.04),
  xlab = "Held-out run",
  ylab = "Accuracy",
  main = "Generalization across runs",
  xaxt = "n"
)
axis(1, at = run_accuracy$run)
abline(h = chance, lty = 2, col = "#777777")
abline(h = accuracy, lty = 1, col = "#C44536")
legend(
  "bottomleft",
  legend = c("Per-run", "Pooled", "Chance"),
  col = c("#15616D", "#C44536", "#777777"),
  lty = c(1, 1, 2),
  pch = c(19, NA, NA),
  bty = "o",
  bg = "white",
  box.col = "white",
  cex = 0.8
)

par(mar = c(7.3, 8.3, 2.2, 0.8))
n_class <- nrow(confusion_rate)
image(
  x = seq_len(n_class),
  y = seq_len(n_class),
  z = t(confusion_rate[n_class:1L, , drop = FALSE]),
  col = hcl.colors(64, "YlGnBu", rev = TRUE),
  zlim = c(0, 1),
  axes = FALSE,
  xlab = "Predicted category",
  ylab = "",
  main = "Which categories are confused?"
)
axis(1, at = seq_len(n_class), labels = colnames(confusion_rate), las = 2, cex.axis = 0.78)
axis(2, at = seq_len(n_class), labels = rev(rownames(confusion_rate)), las = 2, cex.axis = 0.78)
mtext("Observed category", side = 2, line = 6.7)
box()

for (i in seq_len(n_class)) {
  for (j in seq_len(n_class)) {
    value <- confusion_rate[i, j]
    text(
      x = j,
      y = n_class - i + 1L,
      labels = sprintf("%.2f", value),
      cex = 0.63,
      col = if (value >= 0.6) "white" else "#222222"
    )
  }
}

par(old_par)
```

All 12 folds exceed chance, so the pooled result is not carried by a single
run. The confusion matrix also prevents a high average from hiding a failed
category: accuracy ranges from
`r sprintf("%.1f%%", 100 * min(class_accuracy))` for
`r names(which.min(class_accuracy))` to
`r sprintf("%.1f%%", 100 * max(class_accuracy))`. Face and house patterns are
classified at `r sprintf("%.1f%%", 100 * class_accuracy[["face"]])` and
`r sprintf("%.1f%%", 100 * class_accuracy[["house"]])`, respectively.

These diagnostics describe prediction, not mechanism. A classifier can use a
distributed signal without identifying a unique set of causally responsible
voxels, and a high score in one participant does not establish a group effect.

## How do you adapt this workflow?

The scientific contract stays fixed while the spatial question changes:

1. Build an `mvpa_dataset` whose observations align exactly with the design.
2. Put the response and the independence unit in an `mvpa_design`.
3. Choose cross-validation that respects that unit.
4. Validate the specification before fitting.
5. Inspect fold-level and class-level behavior, not only a pooled score.

| This example uses | Replace it with |
|:--|:--|
| Block-mean VT patterns | Trial-wise or condition-wise patterns from your first-level model |
| `category` | The outcome your model should predict |
| `run` | The acquisition or grouping unit that must not cross a fold boundary |
| The published VT mask | A region defined independently of the held-out outcomes |
| `sda_notune` | A classifier chosen before final evaluation, with any tuning nested inside CV |

For several atlas regions, supply an integer-labelled region mask to
`run_regional()`. For a local decoding map, the same model specification can
be sent to the searchlight engine:

```{r searchlight-next, eval = FALSE}
searchlight_result <- run_searchlight(
  model,
  radius = 4,
  preflight = "error"
)
```

A searchlight changes the spatial estimand from “does this predefined region
carry information?” to “where do local neighbourhoods carry information?” It
also introduces many spatial comparisons and substantially more computation;
`vignette("Searchlight_Analysis")` treats those choices directly.

To save the regional tables and a reproducibility manifest:

```{r save-next, eval = FALSE}
save_results(result, dir = "haxby-vt-results")
```

## Where should you go next?

- **Bring your own images and masks:** `vignette("Constructing_Datasets")`.
- **Design leakage-resistant resampling:** `vignette("CrossValidation")`.
- **Select voxels inside the training folds:** `vignette("FeatureSelection")`.
- **Move from one region to a decoding map:** `vignette("Searchlight_Analysis")`.
- **Study representational geometry instead of category prediction:**
  `vignette("RSA")`, followed by `vignette("Kriegeskorte_92_Images")` for a
  real-data example.

The package's core public workflow is now in view:

```text
mvpa_dataset + mvpa_design + cross-validation
                    |
                    v
                mvpa_model
                    |
          +---------+----------+
          |                    |
  run_regional()       run_searchlight()
          |                    |
          +---------+----------+
                    v
       checked predictions, metrics, and maps
```

## References

Haxby JV, Gobbini MI, Furey ML, Ishai A, Schouten JL, Pietrini P (2001).
Distributed and overlapping representations of faces and objects in ventral
temporal cortex. *Science*, **293**, 2425–2430.
[doi:10.1126/science.1063736](https://doi.org/10.1126/science.1063736).

Varoquaux G, Raamana PR, Engemann DA, Hoyos-Idrobo A, Schwartz Y, Thirion B
(2017). Assessing and tuning brain decoders: cross-validation, caveats, and
guidelines. *NeuroImage*, **145**, 166–179.
[doi:10.1016/j.neuroimage.2016.10.038](https://doi.org/10.1016/j.neuroimage.2016.10.038).
