---
title: "Represenational Similarity Analysis"
author: "Bradley Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Represenational Similarity Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Representational Similarity Analysis
```{r, echo=FALSE, message=FALSE}
suppressPackageStartupMessages(library(neuroim2))
suppressPackageStartupMessages(library(devtools))
library(rMVPA)
```

### Generate a volumetric dataset with 100 observations and an external "model" RSA dissimilarity matrix

To generate a dataset we use the `gen_sample_dataset` function. We are creating a 4-dimensional neuroimaging dataset, with 6-by-6-by-6 spatial dimensions and 80 observations in the 4th dimension. These 80 observations are divided into 4 blocks, each consisting of 20 trials.

To run RSA with an external model matrix, we need to create and `rsa_design` object. First we generate a sample dataset with 80 trials as follows.


```{r}

dataset <- gen_sample_dataset(D=c(6,6,6), nobs = 80, blocks=4)

```

Now we create a external dissimilarity matrix with the same number of trials as our fMRI dataset. Then we create an `rsa_design` object and an `rsa_model` object. Finally, we run a searchlight analysis with a 4mm radius using a pearson correlation as a metric to compare the similarity between the external model matrix (Dmat) and 

```{r}
Dmat <- dist(matrix(rnorm(80*100), 80, 100))
rsades <- rsa_design(~ Dmat, list(Dmat=Dmat))

dset <- mvpa_dataset(dataset$dataset$train_data, mask=dataset$dataset$mask)

### if we had a file called "betas.nii.gz" and associated mask, "mask.nii.gz" we would do as follows:
### traindat <- neuroim2::read_vector("betas.nii.gz")
### mask <- neuroim2::read_vector("mask.nii.gz")   
## dset <- mvpa_dataset(traindat, mask=mask)


rsamod <- rsa_model(dset, rsades)
result <- run_searchlight(rsamod, radius=4, method="standard",regtype="pearson")

```

The return object `result` contains a list of `NeuroVol` objects, one for each model matrix. Here, since there is only one external model matrix, `Dmat` there is only one element in the returned list.

```{r}

result

### we can save the result to disk, as follows:

## write_vol(result$Dmat, "RSA_Dmat.nii.gz") 
```



