---
title: Representational Similarity Analysis (RSA) in rMVPA
author: Bradley Buchsbaum
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Representational Similarity Analysis (RSA) in rMVPA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-teal');});</script>

---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
suppressPackageStartupMessages({
  library(neuroim2)
  library(rMVPA)
})
```

## Introduction to Representational Similarity Analysis

Representational Similarity Analysis (RSA) compares neural activity patterns with computational models by measuring pattern similarities, matching them against model predictions, and quantifying how well models explain the neural data. The `rMVPA` package implements this technique for neuroimaging analysis.

## Basic concepts

### Dissimilarity matrices
A dissimilarity matrix represents pairwise differences between conditions or stimuli. Each cell (i, j) quantifies how different two conditions are, and the matrix can be derived from either neural data or theoretical models. Common measures include correlation distance (1 − correlation) and Euclidean distance.

### RSA workflow in rMVPA
In practice you (i) create an MVPA dataset, (ii) define one or more model dissimilarity matrices, (iii) build an RSA design, and (iv) fit and evaluate the RSA model.


## Step-by-Step Example

### 1. Creating Sample Data

Let's create a simple example dataset with known structure:

```{r create_sample_data, message=FALSE, warning=FALSE}
# Generate a sample dataset (20x20x8 volume, 80 observations, 4 blocks)
dataset <- rMVPA::gen_sample_dataset(D=c(20,20,8), nobs = 80, blocks=4)
```

### 2. Creating Model Dissimilarity Matrices

You can use different types of dissimilarity matrices:

```{r create_model_rdms, message=FALSE, warning=FALSE}
# Method 1: Using dist() on feature vectors
model_features <- matrix(rnorm(80*10), 80, 10)  # 80 trials, 10 features
model_rdm <- dist(model_features)  # Default is Euclidean distance

# Method 2: Direct correlation distance matrix
model_matrix <- 1 - cor(t(model_features))  # Correlation distance
```

### 3. Creating an RSA Design

The RSA design specifies how to compare neural and model dissimilarity patterns:

```{r rsa_design_creation, message=FALSE, warning=FALSE}
# Basic design with one model RDM
basic_design <- rsa_design(
  formula = ~ model_rdm,
  data = list(model_rdm = model_rdm),
  block_var = factor(dataset$design$block_var)
)

# Design with multiple model RDMs
model_rdm2 <- dist(matrix(rnorm(80*10), 80, 10))
complex_design <- rsa_design(
  formula = ~ model_rdm + model_rdm2,
  data = list(
    model_rdm = model_rdm,
    model_rdm2 = model_rdm2
  ),
  block_var = factor(dataset$design$block_var),
  keep_intra_run = FALSE  # Exclude within-run comparisons
)
```

### Decorrelating correlated model RDMs

Sometimes your model RDMs are not independent hypotheses. CNN layer RDMs are a
common example: layers such as VGG-16 layer 4, 7, 11, and 16 are ordered and
often share substantial representational structure. If you enter these RDMs
directly into a multiple-RDM RSA model, shared geometry can make the individual
layer effects hard to interpret.

`rdm_decorrelate()` is a preprocessing helper for this situation. It works on
the lower-triangular RDM vectors, estimates layer-specific innovation RDMs, and
returns adjusted RDMs that can be passed directly to `rsa_design()`.

```{r decorrelate_cnn_rdms, message=FALSE, warning=FALSE}
set.seed(12)

shared_visual <- matrix(rnorm(80 * 4), 80, 4)
layer_noise <- lapply(c(0.9, 0.7, 0.5, 0.35), function(sd) {
  matrix(rnorm(80 * 4, sd = sd), 80, 4)
})

cnn_features <- Map(function(noise, weight) {
  base::scale(weight * shared_visual + noise)
}, layer_noise, c(0.45, 0.60, 0.75, 0.90))
names(cnn_features) <- c("vgg4", "vgg7", "vgg11", "vgg16")

cnn_rdms <- lapply(cnn_features, dist)

raw_rdm_vectors <- vapply(cnn_rdms, as.vector, numeric(length(cnn_rdms[[1]])))
round(cor(raw_rdm_vectors, method = "spearman"), 2)
```

The raw RDM correlations show how much representational geometry is shared
across the model layers. The ordered innovation method treats the supplied
order as meaningful: later-layer RDMs are decomposed into the part predicted by
earlier layer innovations and the residual layer-specific component.

```{r rdm_decorrelate_example, message=FALSE, warning=FALSE}
decorrelated <- rdm_decorrelate(
  cnn_rdms,
  method = "ordered_innovation",
  similarity = "spearman",
  epsilon = 0.05,
  gamma_grid = seq(0, 1, by = 0.05)
)

decorrelated$mean_abs_cor_before
decorrelated$mean_abs_cor_after
decorrelated$preservation
```

The result records both the decorrelation achieved and how much each adjusted
RDM still resembles the original. The adjusted RDMs are ordinary `dist`
objects, so they can be used in a standard RSA design.

```{r decorrelated_rsa_design, message=FALSE, warning=FALSE}
cnn_design <- rsa_design(
  formula = ~ vgg4 + vgg7 + vgg11 + vgg16,
  data = decorrelated$rdms,
  block_var = factor(dataset$design$block_var),
  keep_intra_run = FALSE
)
```

```{r check-rdm-decorrelation, include=FALSE}
stopifnot(
  inherits(decorrelated, "rdm_decorrelation_result"),
  decorrelated$mean_abs_cor_after <= decorrelated$mean_abs_cor_before,
  all(is.finite(decorrelated$preservation)),
  all(vapply(decorrelated$rdms, inherits, logical(1), what = "dist")),
  inherits(cnn_design, "rsa_design")
)
```

Use this when you want to test ordered, related model geometries while reducing
shared RSA structure among them. The adjusted RDMs should be interpreted as
layer-specific representational innovations, not as ordinary covariance
shrinkage estimates.

### 4. Creating and Running an RSA Model

The `rsa_model()` function supports different methods for computing neural dissimilarities and analyzing relationships:

```{r run_rsa_searchlight, message=FALSE, warning=FALSE}
# Create MVPA dataset
dset <- mvpa_dataset(dataset$dataset$train_data, mask=dataset$dataset$mask)

# Create RSA model with different options
rsa_spearman <- rsa_model(
  dataset = dset,
  design = basic_design,
  distmethod = "spearman",  # Method for computing neural dissimilarities
  regtype = "spearman"      # Method for comparing neural and model RDMs
)

# Run searchlight analysis
results <- run_searchlight(
  rsa_spearman,
  radius = 4,
  method = "standard"
)
```

## Advanced Features

### Multiple Comparison Methods

`rMVPA` supports several methods for comparing neural and model RDMs:

```{r comparison_methods, message=FALSE, warning=FALSE}
# Pearson correlation
rsa_pearson <- rsa_model(dset, basic_design, 
                        distmethod = "pearson", 
                        regtype = "pearson")

# Linear regression
rsa_lm <- rsa_model(dset, basic_design, 
                    distmethod = "spearman", 
                    regtype = "lm")

# Rank-based regression
rsa_rfit <- rsa_model(dset, basic_design, 
                      distmethod = "spearman", 
                      regtype = "rfit")
```

### Handling Run Structure

RSA can account for the run/block structure of fMRI data. A critical consideration in fMRI analysis is whether to include comparisons between patterns from the same run.

#### Understanding keep_intra_run

The `keep_intra_run = FALSE` parameter tells RSA to exclude comparisons between patterns within the same run/block. This is important because:

1. **Temporal Autocorrelation**: BOLD responses within the same run are temporally autocorrelated
2. **Scanner Drift**: Within-run patterns may share scanner drift effects
3. **Physiological Noise**: Within-run patterns may share structured noise from breathing, heart rate, etc.

Here's a visualization of what `keep_intra_run = FALSE` does:

```{r run_structure_example, message=FALSE, warning=FALSE}
# Create a small example with 2 runs, 4 trials each
mini_data <- matrix(1:8, ncol=1)  # Trial numbers 1-8
run_labels <- c(1,1,1,1, 2,2,2,2)  # Two runs with 4 trials each

# Create distance matrix
d <- dist(mini_data)
d_mat <- as.matrix(d)

# Show which comparisons are kept (TRUE) or excluded (FALSE)
comparison_matrix <- outer(run_labels, run_labels, "!=")
# Only show lower triangle to match distance matrix structure
comparison_matrix[upper.tri(comparison_matrix)] <- NA

# Display the matrices
cat("Trial numbers:\n")
print(matrix(1:8, nrow=8, ncol=8)[lower.tri(matrix(1:8, 8, 8))])
cat("\nRun comparisons (TRUE = across-run, FALSE = within-run):\n")
print(comparison_matrix[lower.tri(comparison_matrix)])
```

When we create an RSA design with `keep_intra_run = FALSE`:

```{r blocked_rsa_design, message=FALSE, warning=FALSE}
# Create design excluding within-run comparisons
blocked_design <- rsa_design(
  formula = ~ model_rdm,
  data = list(model_rdm = model_rdm),
  block_var = factor(dataset$design$block_var),
  keep_intra_run = FALSE  # Exclude within-run comparisons
)
```

This creates an RSA design that includes only between‑run comparisons, excludes within‑run pairs, and focuses the analysis on more reliable across‑run similarities.

#### When to use keep_intra_run = FALSE

Set `keep_intra_run = FALSE` when your design includes multiple runs and you want to control for temporal autocorrelation and run‑specific noise; this is the conservative choice for most confirmatory analyses. Keeping within‑run comparisons (`keep_intra_run = TRUE`, the default) can be reasonable in short‑run designs, when sample size is limited, or for exploratory work where maximizing comparisons is more important than strict control of temporal structure.

### Visualizing Results

You can examine and visualize the RSA results:

```{r visualize_rsa_results, message=FALSE, warning=FALSE}
# Extract the searchlight map
rsa_map <- results$results$model_rdm

# Compute range of correlation values
rsa_values <- neuroim2::values(rsa_map)
range(rsa_values, na.rm = TRUE)

# Basic summary of the searchlight result
print(results)

# Save results (commented out)
# neuroim2::write_vol(rsa_map, "RSA_results.nii.gz")
```

## Summary

The rMVPA package provides a comprehensive RSA implementation with flexible model specification, multiple dissimilarity computation methods, and support for complex experimental designs with run/block structures. It integrates seamlessly with searchlight analysis and offers various statistical approaches including correlation, regression, and rank-based methods.

When using RSA in rMVPA, carefully consider your experimental design when setting block variables and intra-run parameters, choose distance methods that match your theoretical framework, and select statistical approaches appropriate for your analysis goals.

## From within-ROI scores to ROI-to-ROI connectivity

Once you have RSA scores per ROI, a natural follow-up is "where else does this geometry live?" `rsa_model()` accepts `return_fingerprint = TRUE`, which stores a small per-ROI projection of the neural pair vector onto your model RDM subspace. `model_space_connectivity()` then turns those fingerprints into ROI-by-ROI representational connectivity in one call. The same flag also feeds k-means anchor maps for searchlight runs without materialising an `n_centers × n_centers` matrix. See `vignette("Model_Space_Connectivity")` for the full workflow, including cross-domain pair designs (`pair_rsa_design(..., pairs = "between")`).

## Further reading

- `vignette("Model_Space_Connectivity")` -- model-space fingerprints, ROI-to-ROI connectivity, pair_rsa_design, and searchlight anchor maps
- `vignette("Advanced_RSA")` -- Feature-Based and Vector-Based RSA methods
- `vignette("Contrast_RSA")` -- MS-ReVE: contrast-based decomposition of representational geometry
- `vignette("Temporal_Confounds_in_RSA")` -- controlling for temporal proximity confounds

## References

- Kriegeskorte et al. (2008). Representational similarity analysis - connecting the branches of systems neuroscience. Front Syst Neurosci.
- Nili et al. (2014). A toolbox for representational similarity analysis. PLoS Comput Biol.
