---
title: "Vector-Based RSA"
author: Bradley Buchsbaum
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 3
vignette: |
  %\VignetteIndexEntry{Vector-Based RSA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-teal');});</script>
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(albersdown::theme_albers(params$family))
}
library(rMVPA)
```

## What problem does Vector-Based RSA solve?

Standard RSA correlates the lower triangle of a neural RDM with the lower triangle of a model RDM and gives you **one number per ROI**. Vector-Based RSA does something subtly different: for **each trial `i`** it correlates that trial's *row* of the neural distance matrix with the same row of a reference distance matrix — over the trials in *other* blocks only — and returns one similarity score per trial. The mean across trials becomes the ROI's `rsa_score`.

This per-trial framing buys you three things:

1. **Built-in across-block masking.** Within-run pairs (which suffer from temporal autocorrelation, scanner drift, breathing, etc.) are excluded automatically: for trial `i`, only columns where `block_var != block_var[i]` are used.
2. **Per-observation scores.** With `return_predictions = TRUE` you get the full vector of trial-level RSA values, useful for trial-level regressions, item-level analyses, or single-trial QC.
3. **A clean permutation null.** Set `nperm > 0` and a row-permutation null is computed *inside the ROI*, returning `p_rsa_score` and `z_rsa_score` alongside the mean.

## When to use it

Use Vector-Based RSA when:

- You have **one reference RDM** (from features, behaviour, a model, or a prior dataset) and want a single, interpretable RSA score per ROI/searchlight centre.
- Your design has **runs/blocks** and you want within-block comparisons excluded — this is the default, not an opt-in.
- You want **per-trial RSA scores** so you can correlate them with behaviour or a single-trial covariate.
- You want a **permutation-based p-value and z-score** without writing the permutation loop yourself.

If you instead have a feature matrix and want regression-style metrics (pattern correlation, R²) under cross-validation, see `vignette("Feature_RSA")`. If you want to fit multiple model RDMs as competing predictors, see `vignette("RSA")`.

## Inputs and outputs

| Component | What it is |
|---|---|
| `D` | Reference distance matrix, **with `rownames(D)` set**. Rows/columns are the *unique* labels (one row per condition or stimulus). Need not be the same length as the number of trials. |
| `labels` | Length `n_trials` vector. Each entry must appear in `rownames(D)`. The design expands `D` to an `n_trials × n_trials` matrix (`Dexpanded`) by row matching. |
| `block_var` | Length `n_trials` vector of block IDs (e.g. run number). Trials sharing a block are excluded from each other's correlation neighbourhood. |
| `mvpa_dataset` | Neural data: `n_trials × n_voxels` per ROI. |
| `distfun` | A distance object from `create_dist()` (e.g. `cordist()`, `eucdist()`) used on the *neural* data. |
| `rsa_simfun` | `"pearson"` (default) or `"spearman"` — the correlation between the neural and reference distance vectors. |
| Output | Per-ROI `rsa_score` (mean across trials). With `nperm > 0`: also `p_rsa_score`, `z_rsa_score`. With `return_predictions = TRUE`: a `prediction_table` containing the per-trial scores. |

## Minimal example

```{r minimal-data}
set.seed(2026)

# Synthetic neural data: 50 trials in 5 blocks, small mask.
sim <- gen_sample_dataset(D = c(6, 6, 6), nobs = 50, blocks = 5, nlevels = 2)

# A reference distance matrix derived from a feature space.
n_trials   <- 50
n_features <- 5
F          <- matrix(rnorm(n_trials * n_features), n_trials, n_features)
labels     <- paste0("trial_", seq_len(n_trials))
rownames(F) <- labels

D <- as.matrix(dist(F))
rownames(D) <- colnames(D) <- labels
```

Build the design and the model:

```{r minimal-design}
design <- vector_rsa_design(
  D         = D,
  labels    = labels,
  block_var = sim$design$block_var
)

mod <- vector_rsa_model(
  dataset    = sim$dataset,
  design     = design,
  distfun    = cordist(),  # neural distance: 1 - Pearson correlation
  rsa_simfun = "pearson"
)
```

Run regionally over a few ROIs:

```{r minimal-run}
mask_vol <- sim$dataset$mask
nvox     <- sum(mask_vol)
region_mask <- neuroim2::NeuroVol(
  sample(1:3, size = nvox, replace = TRUE),
  neuroim2::space(mask_vol),
  indices = which(mask_vol > 0)
)

res <- run_regional(mod, region_mask)
res$performance_table
```

Each row is one ROI; `rsa_score` is the average across-block second-order correlation.

## What each row actually computes

For trial `i`:

```
neural_dist_row_i  =  pairwise_dist(distfun, X)[i, valid]
reference_row_i    =  Dexpanded[i, valid]
score_i            =  cor(neural_dist_row_i, reference_row_i, method = rsa_simfun)
```

where `valid = which(block_var != block_var[i])`. The ROI's `rsa_score` is `mean(score_i, na.rm = TRUE)`. If a trial has no valid out-of-block partners, its score is `NA` and it contributes nothing to the mean.

## Per-trial scores

If you want the full vector of scores back from `run_regional` or `run_searchlight`, opt in:

```{r return-predictions}
mod_pred <- vector_rsa_model(
  dataset            = sim$dataset,
  design             = design,
  distfun            = cordist(),
  rsa_simfun         = "pearson",
  return_predictions = TRUE
)

res_pred <- run_regional(mod_pred, region_mask)
head(res_pred$prediction_table)
```

The `prediction_table` is one row per trial-per-ROI, with the trial-level `rsa_score`. This lets you correlate RSA strength against trial-level behavioural covariates without re-running the analysis.

## Permutation testing

Setting `nperm > 0` permutes the rows of the ROI's data matrix and recomputes the mean second-order similarity under the null. The output adds `p_rsa_score` and `z_rsa_score`:

```{r perm}
mod_perm <- vector_rsa_model(
  dataset    = sim$dataset,
  design     = design,
  distfun    = cordist(),
  rsa_simfun = "pearson",
  nperm      = 50            # use a larger value (e.g. 1000+) in real analyses
)

res_perm <- run_regional(mod_perm, region_mask)
res_perm$performance_table
```

Permutation is done *within each ROI* — it shuffles which neural row goes with which reference row, holding the block structure of the reference fixed. With `save_distributions = TRUE` the full null distributions are also returned.

## How it differs from standard RSA

| | Standard RSA (`rsa_model`) | Vector-Based RSA (`vector_rsa_model`) |
|---|---|---|
| Granularity | One scalar per ROI from the **upper-triangle vector** of pair dissimilarities | One scalar **per trial**, averaged across trials per ROI |
| Block handling | Optional, via `keep_intra_run = FALSE` | **Built-in** — within-block columns excluded for every trial |
| Multiple model RDMs | Supported as a regression formula | Single reference RDM at a time |
| Output | Coefficients or correlations per model RDM | `rsa_score` + optional `p_rsa_score`, `z_rsa_score`, per-trial `prediction_table` |
| Best for | Comparing competing geometric hypotheses | Trial-level RSA, single-RDM tests with strong block control |

## What's next

- `vignette("Feature_RSA")` — when you have a feature matrix rather than a single reference RDM.
- `vignette("RSA")` — the standard whole-RDM workflow with multiple model RDMs and `keep_intra_run`.
- `?vector_rsa_design`, `?vector_rsa_model`, `?second_order_similarity` — argument reference and the underlying scoring function.
