---
title: Building Plugin Analyses for rMVPA
author: Bradley Buchsbaum
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 3
vignette: |
  %\VignetteIndexEntry{Building Plugin Analyses for rMVPA} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-lapis');});</script>

---

```{r setup, include=FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r libraries}
library(rMVPA)
library(neuroim2)
future::plan(future::sequential)
```

# Why plugin analyses?

`rMVPA` already provides execution harnesses for searchlight and regional analyses (ROI extraction, iteration, parallel orchestration, and spatial map assembly). Plugin analyses let you add new method families in a separate package while reusing that harness.

The plugin contract is small:

1. Constructor: returns a model specification via `create_model_spec()`.
2. `output_schema.<your_class>()`: declares metric names and layout.
3. `fit_roi.<your_class>()`: computes one ROI result.
4. Optional `strip_dataset.<your_class>()`: trims large fields for workers.

# Classifier Registry vs Analysis Plugins

These are different extension paths:

1. Classifier registry (`register_mvpa_model()`): add a new estimator backend (fit/predict/prob) to use inside the standard `mvpa_model` analysis.
2. Analysis plugin contract (`create_model_spec()` + `fit_roi.<class>`): add a new per-ROI analysis type with custom logic.

Use the classifier registry when your workflow is still standard classification/regression and only the estimator changes. Use analysis plugins when the ROI computation itself is novel.

# Step 1: Define a constructor

```{r constructor}
toy_plugin_model <- function(dataset, design, center_stat = c("mean", "median")) {
  center_stat <- match.arg(center_stat)
  create_model_spec(
    name = "toy_plugin_model",
    dataset = dataset,
    design = design,
    center_stat = center_stat,
    compute_performance = TRUE,
    return_predictions = FALSE
  )
}
```

# Step 2: Declare the metric schema

```{r schema}
output_schema.toy_plugin_model <- function(model) {
  list(
    score = "scalar",
    spread = "scalar"
  )
}
```

# Step 3: Implement `fit_roi`

```{r fit-roi}
fit_roi.toy_plugin_model <- function(model, roi_data, context, ...) {
  x <- roi_data$train_data
  score <- if (identical(model$center_stat, "median")) stats::median(x) else mean(x)
  spread <- stats::sd(as.numeric(x))

  roi_result(
    metrics = c(score = score, spread = spread),
    indices = roi_data$indices,
    id = context$id
  )
}
```

For classification/regression-style plugins, you can delegate fold execution to `cv_evaluate_roi()` instead of reimplementing CV loops in each plugin:

```{r fit-roi-cv-wrapper, eval=FALSE}
fit_roi.my_classif_plugin <- function(model, roi_data, context, ...) {
  cv_evaluate_roi(
    model_spec = model,
    roi_data = roi_data,
    context = context,
    mode = "auto"
  )
}
```

# Step 4: Register methods in-session

For a package, these methods are discovered automatically when your package is loaded. In an interactive session, register them explicitly:

```{r register}
registerS3method(
  "fit_roi",
  "toy_plugin_model",
  fit_roi.toy_plugin_model,
  envir = asNamespace("rMVPA")
)
registerS3method(
  "output_schema",
  "toy_plugin_model",
  output_schema.toy_plugin_model,
  envir = asNamespace("rMVPA")
)
```

# Fast plugin validation without full pipelines

Use `mock_roi_data()`, `mock_context()`, and `validate_plugin_model()` to verify your contract before expensive runs.

```{r validate-plugin}
ds <- gen_sample_dataset(c(4, 4, 4), nobs = 20, nlevels = 2, blocks = 2)
mspec <- toy_plugin_model(ds$dataset, ds$design)

plugin_check <- validate_plugin_model(
  mspec,
  roi_data = mock_roi_data(n_train = 20, n_features = 8, n_test = 0, seed = 1),
  context = mock_context(design = ds$design, id = 101L)
)

plugin_check
```

```{r validate-model-spec}
validate_model_spec(mspec, require_schema = TRUE, dry_run = TRUE)
```

# Run on searchlight and regional harnesses

```{r run-harness}
sl_res <- run_searchlight(mspec, radius = 2, method = "standard")
names(sl_res$results)

region_mask <- NeuroVol(
  sample(1:3, size = length(ds$dataset$mask), replace = TRUE),
  space(ds$dataset$mask)
)
reg_res <- run_regional(mspec, region_mask)
names(reg_res$vol_results)
```

# Optional: worker-safe object trimming

If your model spec stores heavy transient objects, define a class method that strips them before distributed execution.

```{r strip-dataset, eval=FALSE}
strip_dataset.toy_plugin_model <- function(obj, ...) {
  obj$dataset <- NULL
  obj
}
```

# Plugin package checklist

1. Export your constructor.
2. Provide `fit_roi.<class>` and `output_schema.<class>`.
3. Keep metric names stable; schema and runtime metrics must match exactly.
4. Add unit tests around `validate_plugin_model()`.
5. Add one lightweight integration test for `run_searchlight()` and/or `run_regional()`.
